### 2018-05-26 DenotSem/document/root.tex
initialization

\documentclass[11pt,a4paper]{article}
\usepackage{isabelle,isabellesym}

% further packages required for unusual symbols (see also
% isabellesym.sty), use only when needed

%\usepackage{amssymb}
  %for \<leadsto>, \<box>, \<diamond>, \<sqsupset>, \<mho>, \<Join>,
  %\<lhd>, \<lesssim>, \<greatersim>, \<lessapprox>, \<greaterapprox>,
  %\<triangleq>, \<yen>, \<lozenge>

%\usepackage{eurosym}
  %for \<euro>

%\usepackage[only,bigsqcap]{stmaryrd}
  %for \<Sqinter>

%\usepackage{eufrak}
  %for \<AA> ... \<ZZ>, \<aa> ... \<zz> (also included in amssymb)

%\usepackage{textcomp}
  %for \<onequarter>, \<onehalf>, \<threequarters>, \<degree>, \<cent>,
  %\<currency>

% this should be the last package used
\usepackage{pdfsetup}

% urls in roman style, theory text in math-similar italics
\urlstyle{rm}
\isabellestyle{it}

% for uniform font size
%\renewcommand{\isastyle}{\isastyleminor}

\begin{document}

\title{DenotSem}
\author{By jsiek}
\maketitle

\tableofcontents

% sane default for proof documents
\parindent 0pt\parskip 0.5ex

% generated text of all theories
\input{session}

% optional bibliography
%\bibliographystyle{abbrv}
%\bibliography{root}

\end{document}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: t
%%% End:

### 2018-07-21 DenotSem/document/root.tex
abstract

@@
 
-%\usepackage[only,bigsqcap]{stmaryrd}
+\usepackage[only,bigsqcap]{stmaryrd}
   %for \<Sqinter>
@@
 
-\title{DenotSem}
-\author{By jsiek}
+\title{Modeling the Functional Character \\ of the Lambda Calculus}
+\author{Jeremy G. Siek}
 \maketitle
+
+\begin{abstract}
+  Denotational semantics for the lambda calculus that are given by graph models
+  or filter models (i.e. intersection type systems) use finite relations
+  to model functions. In this document I use a notion of consistency 
+  to restrict these relations to be functions (or finite approximations of them) 
+  and I prove that the model is deterministic in the sense that any two outputs
+  from the same program are equal, and furthermore, that the result of merging
+  the two outputs is also an output of the program. Phrasing this in terms
+  of an intersection type system, I show that the subsumption rule and the
+  intersection introduction rule are both admissible.
+\end{abstract}
 
@@
 \parindent 0pt\parskip 0.5ex
+
+\pagebreak
 

### 2018-08-07 DenotSem/determ.tex
start of blog post

\documentclass{article}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{hyperref}

\newcommand{\lam}[1]{\lambda #1.\,}
\newcommand{\app}[0]{\;}
\newcommand{\IF}[0]{\textbf{if}\;}
\newcommand{\THEN}[0]{\;\textbf{then}\;}
\newcommand{\ELSE}[0]{\;\textbf{else}\;}
\newcommand{\ESEM}[1]{\mathcal{E}[\![ #1 ]\!]}
\newcommand{\FSEM}[1]{\mathcal{F}[\![ #1 ]\!]}
\newcommand{\SET}[1]{\mathcal{P}(#1)}

\title{Intersection Types, Cut Elimination, and the Functional
  Character of the Lambda Calculus}

\author{Jeremy G. Siek}

\begin{document}
\maketitle

Last December I proved that my graph model of the lambda calculus,
once suitable restricted, is deterministic. That is, I defined a
notion of \emph{consistency} between values, written $v_1 \sim v_2$,
and showed that any two outputs of the same program are consistent. \\

\noindent \textbf{Theorem} (Determinism of $\mathcal{E}$)
  If $v \in \ESEM{e}\rho$, $v' \in \ESEM{e}\rho'$, and $\rho \sim
  \rho'$, then $v \sim v'$. \\

\noindent Recall that values are integers or finite relations;
consistency for integers is equality and consistency for relations
means mapping consistent inputs to consistent outputs.

Having proved the Determinism Theorem, I thought it would be
straightforward to prove the following related theorem about the join
of two values. \\

\noindent \textbf{Theorem} (Join in $\mathcal{E}$)
  If $v \in \ESEM{e}\rho$, $v' \in \ESEM{e}\rho'$, and $\rho \sim
  \rho'$, \\
  then $v \sqcup v' \in \ESEM{e}(\rho\sqcup\rho')$. \\

Recall that we have defined a partial order $\sqsubseteq$ on values,
and that, in most partial orders, there is a close connection between
notions of consistency and least upper bounds (joins). One typically
has that $v \sim v'$ iff $v \sqcup v'$ exists. So my thinking was that
it should be easy to adapt my proof of the determinism theorem to
prove this theorem about joins, and I set out hoping to finish in a
couple weeks. Hah! Here we are 8 months later and the proof is
complete, but my journey to find the proof was rather circuitous and
ended up depending on a recent result about intersection types and cut
elimination from Olivier Laurent. In this blog post I'll try to
recount the journey and describe the proof, hopefully remembering all
the challenges and motivations that led to the proof.

Many of the challenges revolved around the definitions of
$\sqsubseteq$, consistency, and $\sqcup$. Given that I already had
definitions for $\sqsubseteq$ and consistency, the obvious thing to
try was to define $\sqcup$ such that it would be the least upper bound
of $\sqsubseteq$. So I arrived at this partial function:
\begin{align*}
  n \sqcup n &= n \\
  f_1 \sqcup f_2 &= f_1 \cup f_2
\end{align*}
Now suppose we prove the Join Theorem by induction on $e$ and consider
the case for application: $e = (e_1 \app e_2)$.
We have
\begin{itemize}
\item $f \in \ESEM{e_1}\rho$, $v_2 \in \ESEM{e_2}\rho$,  
  $v_3 \mapsto v_4 \in f$, $v_3 \sqsubseteq v_2$, and $v \sqsubseteq v_4$
  for some $f, v_2, v_3, v_4$.
\item $f' \in \ESEM{e_2}\rho'$, $v'_2 \in \ESEM{e_2}\rho'$,   
  $v'_3 \mapsto v'_4 \in f$, $v'_3 \sqsubseteq v'_2$, and $v' \sqsubseteq v'_4$
  for some $f', v'_2, v'_3, v'_4$.
\end{itemize}
By the induction hypothesis we have $f \sqcup f' \in \ESEM{e_1}$
and $v_2 \sqcup v'_2 \in \ESEM{e_2}$.
We need to show that 
\[
   v''_3 \mapsto v''_4 \in f \sqcup f' 
   \qquad
   v''_3 \sqsubseteq v_2 \sqcup v'_2
   \qquad
   v \sqcup v' \sqsubseteq v''_4
\]
But we have a problem, given our definition of $\sqcup$ in terms of
set union, there won't necissarily be a single entry in $f \sqcup f'$
that combines the information from both $v_3 \mapsto v_4$ and $v'_3
\mapsto v'_4$. After all, $f \sqcup f'$ contains all the entries of
$f$ and all the entries of $f'$, but no entries that mix together the
information from entries in $f$ and $f'$.

At this point I started thinking that my definitions of $\sqsubseteq$,
consistency, and $\sqcup$ were too simple, and that I needed to
incorporate ideas from the literature on filter models and
intersection types. As I've written about previously, my graph model
corresponds to a particular intersection type system, and perhaps a
different intersection type system would do the job. Recall that the
correspondence goes as follows: values correspond to types,
$\sqsubseteq$ corresponds to subtyping $<:$ (in reverse), and $\sqcup$
corresponds to intersection $\sqcap$. The various intersection type
systems primarily differ in their definitions of subtyping.  Given the
above proof attempt, I figured that I would need usual
co/contra-variant rule for function types and also the following rule
for distributing intersections over function types.
\[
  (A\to B) \sqcap (A \to C) <: A \to (B \sqcap C)
\]
This distributivity rule enables the ``mixing'' of information from
two different entries.

So I defined types as follows:
\[
  A,B,C,D ::= n \mid A \to B \mid A \sqcap B
\]
and defined subtyping to follow that of the BCD intersection type
system.
\begin{gather*}
A <: A \qquad \frac{A <: B \quad B <: C}{A <: C} \\[2ex]
A \sqcap B <: A \qquad
A \sqcap B <: B \qquad
\frac{C <: A \quad C <: B}{C <: A \sqcap B} \\[2ex]
\frac{C <: A \quad B <: D}{A \to B <: C \to D}
\qquad
(A\to B) \sqcap (A \to C) <: A \to (B \sqcap C)
\end{gather*}
I then adapted the definition of consistency to work over types.
Because of the use of negation in this definition, it is easier to
define consistency as a recursive function in Isabelle instead of as
an inductively defined relation.
\begin{align*}
 n \sim n' &= (n = n') \\
 n \sim (C \to D) &= \mathit{false} \\
 n \sim (C \sqcap D) &= n \sim C \text{ and } n \sim D \\
 (A \to B) \sim n' &= \mathit{false} \\
 (A \to B) \sim (C \to D) &= 
    (A \sim C \text{ and } B \sim D) \text{ or } A \not\sim C \\
 (A \to B) \sim (C \sqcap D) &=
    (A \to B) \sim C \text{ and } (A \to B) \sim D \\
 (A \sqcap B) \sim n' &= A \sim n' \text{ and } B \sim n' \\
 (A \sqcap B) \sim (C \sqcap D) &= 
    A \sim C \text{ and } A \sim D \text{ and } 
    B \sim C \text{ and } B \sim D
\end{align*}

Turning back to the Join Theorem, let us restate it in terms of an
intersection type system. Instead of using the letter $\rho$ for
environments, we shall switch to $\Gamma$ because they now contain
types instead of values. \\

\noindent \textbf{Theorem} (Intersection Introduction is Admissible)
  If $\Gamma \vdash e : A$, $\Gamma' \vdash e : B$, and $\Gamma \sim
  \Gamma'$, 
  then $\Gamma\sqcap\Gamma' \vdash e : A \sqcap B$. \\

By restating the theorem in terms of intersection types, we have
essentially arrived at the rule for intersection introduction.  In
other words, if we can prove this theorem we will have shown that the
intersection introduction rule is admissible in our system.

While the change to intersection types enabled this top-level proof to
go through, I got stuck on one of the lemmas that it requires, which
is an adaptation of Proposition 3 of the prior blog post. \\

\noindent \textbf{Lemma} (Consistency and Subtyping)
\begin{enumerate}
\item  If $A \sim B$, $A <: C$, and $B <: D$,
  then $C \sim D$.
\item If $A \not\sim B$, $C <: A$, $D <: B$, then $C \not\sim D$.
\end{enumerate}
In particular, I got stuck in the cases where the subtyping $A <: C$
or $B <: D$ was derived using the transitivity subtyping rule.

\end{document}

%%  LocalWords:  denotationally ctx subexpression eq denot lrcl env

### 2018-08-08 DenotSem/determ.tex
check

@@
 
-\title{Intersection Types, Cut Elimination, and the Functional
+\title{Intersection Types, Sub-formula Property, and the Functional
   Character of the Lambda Calculus}
@@
 
-\noindent \textbf{Theorem} (Determinism of $\mathcal{E}$)
+\noindent \textbf{Theorem} (Determinism)\\
   If $v \in \ESEM{e}\rho$, $v' \in \ESEM{e}\rho'$, and $\rho \sim
@@
 consistency for integers is equality and consistency for relations
-means mapping consistent inputs to consistent outputs.
+means mapping consistent inputs to consistent outputs. I then
+restricted values to be well formed, meaning that they must be
+consistent with themselves (and similarly for their parts).
 
@@
 
-\noindent \textbf{Theorem} (Join in $\mathcal{E}$)
-  If $v \in \ESEM{e}\rho$, $v' \in \ESEM{e}\rho'$, and $\rho \sim
-  \rho'$, \\
+\noindent \textbf{Theorem} (Join)\\
+If $v \in \ESEM{e}\rho$, $v' \in \ESEM{e}\rho'$,
+$\rho$ is well formed, $\rho'$ is well formed,
+and $\rho \sim \rho'$, \\
 then $v \sqcup v' \in \ESEM{e}(\rho\sqcup\rho')$. \\
+
+I am particularly interested in this theorem because $\beta$-equality
+can be obtained as a corollary.
+\[
+\ESEM{(\lam{x}e)\app e'}\rho = \ESEM{[x{:=}e']e}\rho
+\]
+This would enable the modeling of the call-by-name $\lambda$-calculus
+and it would also enable the use of $\beta$-equality in a
+call-by-value setting when $e'$ is terminating (instead of restricting
+$e'$ to be a syntactic value).
 
@@
 has that $v \sim v'$ iff $v \sqcup v'$ exists. So my thinking was that
-it should be easy to adapt my proof of the determinism theorem to
-prove this theorem about joins, and I set out hoping to finish in a
-couple weeks. Hah! Here we are 8 months later and the proof is
-complete, but my journey to find the proof was rather circuitous and
-ended up depending on a recent result about intersection types and cut
-elimination from Olivier Laurent. In this blog post I'll try to
-recount the journey and describe the proof, hopefully remembering all
-the challenges and motivations that led to the proof.
+it should be easy to adapt my proof of the Determinism Theorem to
+prove the Join Theorem, and I set out hoping to finish in a couple
+weeks. Hah! Here we are 8 months later and the proof is complete; it
+was a long journey that ended up depending on a result that was
+published just this summer, concerning intersection types, the
+sub-formula property, and cut elimination by Olivier Laurent. In this
+blog post I'll try to recount the journey and describe the proof,
+hopefully remembering the challenges and motivations.
 
@@
 Now suppose we prove the Join Theorem by induction on $e$ and consider
-the case for application: $e = (e_1 \app e_2)$.
-We have
+the case for application: $e = (e_1 \app e_2)$. From $v \in \ESEM{e_1
+  \app e_2}$ and $v' \in \ESEM{e_1 \app e_2}$ we have
 \begin{itemize}
@@
 \]
-But we have a problem, given our definition of $\sqcup$ in terms of
-set union, there won't necissarily be a single entry in $f \sqcup f'$
-that combines the information from both $v_3 \mapsto v_4$ and $v'_3
-\mapsto v'_4$. After all, $f \sqcup f'$ contains all the entries of
-$f$ and all the entries of $f'$, but no entries that mix together the
-information from entries in $f$ and $f'$.
+But here we have a problem. Given our definition of $\sqcup$ in terms
+of set union, there won't necissarily be a single entry in $f \sqcup
+f'$ that combines the information from both $v_3 \mapsto v_4$ and
+$v'_3 \mapsto v'_4$. After all, $f \sqcup f'$ contains all the entries
+of $f$ and all the entries of $f'$, but the set union operation does
+not mix together information from entries in $f$ and $f'$ to form new
+entries.
 
@@
 \]
-and defined subtyping to follow that of the BCD intersection type
-system.
+and defined subtyping according to the BCD intersection type system.
 \begin{gather*}
@@
 I then adapted the definition of consistency to work over types.
-Because of the use of negation in this definition, it is easier to
-define consistency as a recursive function in Isabelle instead of as
-an inductively defined relation.
+(Because this definition uses negation, it is easier to define
+consistency as a recursive function in Isabelle instead of as an
+inductively defined relation.)
 \begin{align*}
@@
 
-Turning back to the Join Theorem, let us restate it in terms of an
+Turning back to the Join Theorem, I restated it in terms of the
 intersection type system. Instead of using the letter $\rho$ for
@@
 
-\noindent \textbf{Theorem} (Intersection Introduction is Admissible)
+\noindent \textbf{Theorem} (Intersection Introduction)\\
   If $\Gamma \vdash e : A$, $\Gamma' \vdash e : B$, and $\Gamma \sim
@@
 
-While the change to intersection types enabled this top-level proof to
-go through, I got stuck on one of the lemmas that it requires, which
-is an adaptation of Proposition 3 of the prior blog post. \\
+While the switch to intersection types and subtyping enabled this
+top-level proof to go through, I got stuck on one of the lemmas that
+it requires, which is an adaptation of Proposition 3 of the prior blog
+post. \\
 
@@
 In particular, I got stuck in the cases where the subtyping $A <: C$
-or $B <: D$ was derived using the transitivity subtyping rule.
+or $B <: D$ was derived using the transitivity rule.
+
+For a long time I've disliked definition of subtyping in which
+transitivity is given as a rule instead of proved as a theorem.  There
+are several reasons for this: a subtyping algorithm can't directly
+implement a transitivity rule (or any rule that is not syntax
+directed), reasoning by induction or cases (inversion) is more
+difficult, and it is redundant. Furthermore, the presense of the
+transitivity rule means that subtyping does not satisfy the subformula
+property. This term subformula property comes from logic, and means
+that a derivation (proof) of a formula only mentions propositions that
+are a part of the formulate to be proved. The transitivity rule breaks
+this property because the type $B$ comes out of nowhere, it is not
+part of $A$ or $C$, the types in the conclusion of the rule.
+
+So I removed the transitivity rule and tried to prove transitivity.
+For most type systems, proving the transitivity of subtyping is
+straightforward. But what I soon realized that the addition of the
+distributivity rule makes it significantly more difficult.  After
+trying and failing to prove transitivity for some time, I resorted to
+reading the literature. Unfortunately, it turns out that none of the
+published intersection type systems satisfied the subformula property
+and vast majority of them included the transitivity rule.  However,
+there was one paper that offered some hope.  In a 2012 article in
+Fundamenta Informaticae titled \emph{Intersection Types with Subtyping
+  by Means of Cut Elimination}, Olivier Laurent defined subtyping
+without transitivity and instead proved it, but his system still did
+not satisfy the sub-formula property because of an addition rule for
+function types. Nevertheless, Olivier indicated that he was interested
+in finding a version of the system that did, writing
+\begin{quote}
+``it would be much nicer and much more natural to go through a
+  sub-formula property''
+\end{quote}
+
+A lot of progress can happen in six years, so I sent an email to
+Olivier. He replied,
+\begin{quote}
+``Indeed! I now have two different sequent-calculus systems which are
+  equivalent to BCD subtyping and satisfy the sub-formula property.  I
+  am currently writting a paper on this but it is not ready yet.''
+\end{quote}
+and he attached the paper draft and the Coq mechanization. What great
+timing!  Furthermore, Olivier would be presenting the paper, titled
+\emph{Intersection Subtyping with Constructors}, at the Workshop on
+Intersection Types and Related System in Oxford on July 8, part of the
+Federated Logic Conference (FLOC). I was planning to attend FLOC
+anyways, for the DOMAINS workshop to celabrate Dana Scott's 85th
+birthday.
+
+Olivier's systems makes two important changes compared to prior work:
+he combines the distributivity rule and the usual arrow rule into a
+single elegant rule, and to enable this, he generalizes the form of
+subtyping from $A <: B$ to $A_1,\ldots,A_n \vdash B$, which should be
+interpreted as meaning $A_1 \sqcap \cdots \sqcap A_n <: B$.  Having a
+sequence of formulas (types) on the left is characteristic of proof
+systems in logic, including both natural deduction systems and
+sequence calculi. (Sequent calculi, in addition, typically have a
+sequence on the right that means the disjunction of the formulas.)
+Here is one of Olivier's systems, adapted to my setting, which I'll
+describe below. Let $\Gamma$ range over sequences of types.
+\begin{gather*}
+  \frac{\Gamma_1, \Gamma_2 \vdash A}
+       {\Gamma_1 , n, \Gamma_2 \vdash A} \qquad
+  \frac{\Gamma_1, \Gamma_2 \vdash A}
+       {\Gamma_1 , B \to C, \Gamma_2 \vdash A}
+   \\[2ex]
+   \frac{\Gamma \vdash A \quad \Gamma \vdash B}{\Gamma \vdash A \sqcap B}
+   \qquad
+   \frac{\Gamma_1,B,C,\Gamma_2 \vdash A}{\Gamma_1,B\sqcap C,\Gamma_2 \vdash A}
+   \\[2ex]
+   \frac{}{n \vdash n}
+   \qquad
+   \frac{A \vdash C_1, \ldots, C_n \quad
+         D_1, \ldots, D_n \vdash B}
+        {C_1\to D_1,\ldots, C_n\to D_n \vdash A \to B}
+\end{gather*}
+The first two rules are weakening rules for singleton integers and
+function types. There is no weakening rule for intersections.  The
+third and fourth rules are introduction and elimination rules for
+intersection. The fifth rule is reflexivity for integers, and the last
+is the combined rule for function types.
+
+The combined rule for function types says that the intersection of a
+sequence of function types $\bigsqcap_{i=1\ldots n} (C_i\to D_i)$ is a
+subtype of $A \to B$ if
+\[
+  A <: \bigsqcap_{i\in\{1\ldots n\}} C_i
+  \qquad \text{and}\qquad
+  \bigsqcap_{i\in\{1\ldots n\}} D_i <: B
+\]
+Interestingly, the inversion principle for this rule is the
+$\beta$-sound property described in Chapter 14 of \emph{Lambda
+  Calculus with Types} by Barendregt et al., and is the key to proving
+$\beta$-equality. In Olivier's system, $\beta$-soundness falls out
+immediately, instead of by a somewhat involved proof.
+
+The regular subtyping rule for function types is simply an instance of
+the combined rule in which the sequence on the left contains just one
+function type.
+
+The next step for me was to enter Olivier's definitions into Isabelle
+and prove transitivity via cut elimination. That is, I needed to prove
+the following generalized statement via a sequence of lemmas laid out
+by Olivier in his draft.\\
+
+\noindent \textbf{Theorem} (Cut Elimination)\\
+  If $\Gamma_2 \vdash B$ and $\Gamma_1,B,\Gamma_3 \vdash C$,
+  then $\Gamma_1,\Gamma_2,\Gamma_3 \vdash C$. \\
+
+The transitivity rule is the instance of cut elimination where
+$\Gamma_2 = A$ and both $\Gamma_1$ and $\Gamma_3$ are empty.
+
+Unfortunately, I couldn't resist making changes to Olivier's subtyping
+system as I entered it into Isabelle, which cost me considerable time.
+Some of Olivier's lemmas show that the collection of types on the
+left, that is, the $A's$ in $A_1,\ldots, A_n \vdash B$, behave like a
+set instead of a sequence. I figured that if the left-hand-side was
+represented as a set, then I would be able to bypass several lemmas
+and obtain a shorter proof. I got stuck in proving Lemma $\cap L_e$
+which states that $\Gamma_1,A\sqcap B,\Gamma_2 \vdash C$ implies
+$\Gamma_1,A, B,\Gamma_2 \vdash C$. Olivier's subtyping rules are
+carefully designed to minimize the amount of overlap between the
+rules, and switching to a set representation increases the amount of
+overlap, making the proof of this lemma more difficult (perhaps
+impossible?).
+
+So after struggling with the set representation for some time, I went
+back to sequences and was able to complete the proof of cut
+elimination, with a little help from Olivier at FLOC.  I proved the
+required lemmas in the following order. \\
+
+\noindent \textbf{Lemma} (Weakening)\\
+  If $\Gamma_1,\Gamma_2 \vdash A$,
+  then $\Gamma_1,B,\Gamma_2 \vdash A$. \\
+  (Proved by induction on $A$.) \\
+
+\noindent \textbf{Lemma} (Axiom)\\
+$A \vdash A$ \\
+  (Proved by induction on $A$.)\\
+
+\noindent \textbf{Lemma} (Permutation)\\
+  If $\Gamma_1 \vdash A$ and $\Gamma_2$ is a permutation of $\Gamma_1$,
+  then $\Gamma_2 \vdash A$. \\
+  (Proved by induction on the derivation of $\Gamma_1 \vdash A$,
+   using many lemmas about permutations.)\\
+
+\noindent \textbf{Lemma} ($\cap L_e$) \\
+If $\Gamma_1,A\sqcap B,\Gamma_2 \vdash C$, then
+$\Gamma_1,A, B,\Gamma_2 \vdash C$. \\
+(Proved by induction on the derivation of $\Gamma_1,A\sqcap B,\Gamma_2
+\vdash C$.)\\
+
+\noindent \textbf{Lemma} (Collapse Duplicates) \\ If
+$\Gamma_1,A,A,\Gamma_2 \vdash C$, then $\Gamma_1,A,\Gamma_2 \vdash
+C$. \\
+%
+(This is proved by well-founded induction on the lexicographical
+ordering of the pair $(n,k)$ where $n$ is the size of $A$ and $k$ is
+the depth of the derivation of $\Gamma_1,A,A,\Gamma_2 \vdash C$. Proof
+assistants such as Isabelle and Coq do not directly provide the depth
+of a derivation, but the depth can be manually encoded as an extra
+argument of the relation, as in $\Gamma_1,A,A,\Gamma_2 \vdash_k C$.) \\
+
+The Cut Elimination Theorem is then proved by well-founded induction
+on the triple $(n,k_1,k_2)$ where $n$ is the size of B, $k_1$ is the
+depth of the derivation of $\Gamma_2 \vdash B$, and $k_2$ is the depth
+of the derivation of $\Gamma_1,B,\Gamma_3 \vdash C$.
 

### 2018-08-08 DenotSem/determ.tex
check

@@
 \newcommand{\app}[0]{\;}
-\newcommand{\IF}[0]{\textbf{if}\;}
-\newcommand{\THEN}[0]{\;\textbf{then}\;}
-\newcommand{\ELSE}[0]{\;\textbf{else}\;}
+\newcommand{\IF}[0]{\mathrm{if}\;}
+\newcommand{\THEN}[0]{\;\mathrm{then}\;}
+\newcommand{\ELSE}[0]{\;\mathrm{else}\;}
 \newcommand{\ESEM}[1]{\mathcal{E}[\![ #1 ]\!]}
@@
 \newcommand{\SET}[1]{\mathcal{P}(#1)}
+\newcommand{\WF}[1]{\mathsf{wf}(#1)}
+\newcommand{\UP}[1]{\mathord{\uparrow} #1}
 
@@
 
+\section{Intersection Types to the Rescue}
+
 At this point I started thinking that my definitions of $\sqsubseteq$,
@@
 or $B <: D$ was derived using the transitivity rule.
+
+\section{Subtyping and the Sub-formula Property}
 
@@
 
+\section{The Home Stretch}
+
+Armed with a good definition of intersection types, I turned back to
+the proof of the Join Theorem, but first I needed to adjust my
+denotational semantics to use intersection types. For this we'll need
+the definition of well formed types that we alluded to earlier.
+
+\begin{gather*}
+  \frac{}{\WF{n}}
+  \qquad
+  \frac{\WF{A} \quad \WF{B}}{\WF{A \to B}}
+  \qquad
+  \frac{A \sim B \quad \WF{A} \quad \WF{B}}{\WF{A \sqcap B}}
+\end{gather*}
+
+We also introduce the following notation for all the well-formed types
+that are super-types of a given type.
+\[
+   \UP{A} \stackrel{def}{=} \{ B\mid A <: B \text{ and } \WF{B} \}
+\]
+
+We shall represent variables with de Bruijn indices, so an environment
+$\rho$ is a sequence of types.  The denotational semantics of the CBV
+lambda calculus is defined as follows.
+\begin{align*}
+  \ESEM{n}\rho &= \UP{n} \\
+  \ESEM{x}\rho &=  \IF x < |\rho| \THEN \UP{\rho_k} \ELSE \emptyset \\
+  \ESEM{\lam{x}e }\rho &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\rho) \} \\
+  \ESEM{e_1\app e_2}\rho &= \{ C \mid
+    A \in \ESEM{e_1}\rho,
+    B \in \ESEM{e_2}\rho,
+    A <: B \to C, \text{ and } \WF{C} \} \\
+  \ESEM{f(e_1,e_2)}\rho &=
+    \{ \}
+\end{align*}
+
 \end{document}

### 2018-08-08 DenotSem/determ.tex
progress

@@
 \newcommand{\ELSE}[0]{\;\mathrm{else}\;}
-\newcommand{\ESEM}[1]{\mathcal{E}[\![ #1 ]\!]}
-\newcommand{\FSEM}[1]{\mathcal{F}[\![ #1 ]\!]}
+\newcommand{\SEM}[1]{[\![ #1 ]\!]}
+\newcommand{\ESEM}[1]{\mathcal{E}\SEM{#1}}
+\newcommand{\FSEM}[1]{\mathcal{F}\SEM{#1}}
 \newcommand{\SET}[1]{\mathcal{P}(#1)}
@@
 \newcommand{\UP}[1]{\mathord{\uparrow} #1}
+\newcommand{\ATOMS}[1]{\mathit{atoms}(#1)}
 
@@
 
-\noindent \textbf{Lemma} (Collapse Duplicates) \\ If
-$\Gamma_1,A,A,\Gamma_2 \vdash C$, then $\Gamma_1,A,\Gamma_2 \vdash
+\noindent \textbf{Lemma} (Collapse Duplicates) \\
+%
+If $\Gamma_1,A,A,\Gamma_2 \vdash C$, then $\Gamma_1,A,\Gamma_2 \vdash
 C$. \\
@@
 
+Here are some examples and non-examples of well-formed types.
+\begin{gather*}
+  \WF{4} \qquad \WF{3 \sqcap 3} \qquad \neg \WF{3 \sqcap 4} \\
+  \WF{(0\to 1) \sqcap (2 \to 3)} \qquad \neg \WF{(0 \to 1) \sqcap (0 \to 2)}
+\end{gather*}
+It is sometimes helpful to think of well-formed types in terms of the
+equivalence classes determined by subtype equivalence:
+\[
+A \approx B \quad = \quad
+  A <: B \text{ and } B <: A
+\]
+For example, we have $3 \approx (3 \sqcap 3)$, so they are
+in the same equivalence class and $3$ would be the representative.
+
 We also introduce the following notation for all the well-formed types
@@
 \[
-   \UP{A} \stackrel{def}{=} \{ B\mid A <: B \text{ and } \WF{B} \}
+   \UP{A} \quad = \quad \{ B\mid A <: B \text{ and } \WF{B} \}
 \]
@@
 $\rho$ is a sequence of types.  The denotational semantics of the CBV
-lambda calculus is defined as follows.
+$\lambda$-calculus is defined as follows.
 \begin{align*}
@@
   \ESEM{x}\rho &=  \IF x < |\rho| \THEN \UP{\rho_k} \ELSE \emptyset \\
-  \ESEM{\lam{x}e }\rho &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\rho) \} \\
-  \ESEM{e_1\app e_2}\rho &= \{ C \mid
+  \ESEM{\lambda e }\rho &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\rho) \} \\
+  \ESEM{e_1\app e_2}\rho &= \left\{ C\, \middle| \exists A,B.\;
+      \begin{array}{l}
       A \in \ESEM{e_1}\rho,
-    B \in \ESEM{e_2}\rho,
-    A <: B \to C, \text{ and } \WF{C} \} \\
+      B \in \ESEM{e_2}\rho,\\
+      A <: B \to C, \text{ and } \WF{C}
+      \end{array}
+        \right\} \\
   \ESEM{f(e_1,e_2)}\rho &=
-    \{ \}
+      \left\{ C\, \middle| \begin{array}{l}
+       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\rho, B \in \ESEM{e_2}\rho,\\
+      \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
+      \end{array} \right\} \\
+  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\rho &=
+    \left\{ B\, \middle| \exists A.\;
+    \begin{array}{l}A \in \ESEM{e_1}\rho, A <: n,\\
+           n = 0 \Rightarrow B \in \ESEM{e_3}\rho,\\
+           n \neq 0 \Rightarrow B \in \ESEM{e_2}\rho
+    \end{array}
+    \right\}
+    \\[2ex]
+    \intertext{where}
+   \FSEM{n}e\rho &= \mathit{false} \\
+   \FSEM{A \sqcap B}e \rho &= \FSEM{A}e\rho \text{ and } \FSEM{B}e\rho\\
+   \FSEM{A \to B}e \rho &= B \in \ESEM{e} (A, \rho)
 \end{align*}
+
+\noindent \textbf{Proposition} (Consistency is Symmetric and Reflexive) \\
+%
+\begin{enumerate}
+\item If $A \sim B$, then $B \sim A$.
+\item If $\WF{A}$, then $A \sim A$.
+\end{enumerate}
+
+It will often be convenient to decompose a type into its set of
+\emph{atoms}, defined as follows.
+\begin{align*}
+  \ATOMS{n} &= \{ n \} \\
+  \ATOMS{A\to B} &= \{ A \to B \} \\
+  \ATOMS{A \sqcap B} &= \ATOMS{A} \cup \ATOMS{B}
+\end{align*}
+
+The consistency of two types is determined by the consistency of its
+atoms. \\
+
+\noindent \textbf{Lemma} (Atomic Consistency) 
+\begin{enumerate}
+\item If $A \sim B$, $C \in \ATOMS{A}$, and $D \in \ATOMS{B}$,
+  then $C \sim D$.
+\item If (for any $C \in \ATOMS{A}$ and $D \in \ATOMS{B}$, 
+  $C \sim D$), then $A \sim B$.
+\item If $A \not\sim B$, then $C \not\sim D$ for some
+  $C \in \ATOMS{A}$ and $D \in \ATOMS{B}$.
+\item If $C \not\sim D$, $C \in \ATOMS{A}$, and $D \in \ATOMS{B}$,
+  then $A \not\sim B$. 
+\end{enumerate}
+
+There are also several properties of subtyping and the atoms of a type.\\
+
+\noindent \textbf{Lemma} (Atomic Subtyping)
+\begin{enumerate}
+\item If $A <: B$ and $C \in \ATOMS{B}$, then $A <: C$.
+\item If $A <: n$, then $n \in \ATOMS{A}$.
+\item $n <: A$ if and only if $\ATOMS{A} \subseteq \{ n \}$.
+\item If $C <: A \to B$, then $D\to E \in \ATOMS{C}$ for some $D,E$.
+\item If $\Gamma \vdash A$ and every atom in $\Gamma$ is a function type,
+  then every atom of $A$ is a function type.
+\end{enumerate}
+
+And we have the following important inversion lemma for function
+types. We use the following abbreviations:
+\begin{align*}
+  \mathrm{dom}(\Gamma) &= \{ A \mid \exists B.\; A \to B \in \Gamma \}\\
+  \mathrm{cod}(\Gamma) &= \{ B \mid \exists A.\; A \to B \in \Gamma \}
+\end{align*}
+
+\noindent \textbf{Lemma} (Subtyping Inversion for Funcion Types)\\
+%
+If $C <: A \to B$, then there is a sequence of function types $\Gamma$
+such that
+\begin{enumerate}
+\item each element of $\Gamma$ is an atom of $C$,
+\item For every $D\to E \in \Gamma$, we have $A <: D$, and
+\item $\bigsqcap \mathrm{cod}(\Gamma) <: B$.
+\end{enumerate}
+Note that item 2 above implies that $A <: \bigsqcap \mathrm{dom}(\Gamma)$.
+
+With the above lemmas, we are ready to prove the Consistency and
+Subtyping Lemma that previously stumped me.\\
+
+\noindent \textbf{Lemma} (Consistency and Subtyping)
+\begin{enumerate}
+\item  If $A \sim B$, $A <: C$, and $B <: D$,
+  then $C \sim D$.
+\item If $A \not\sim B$, $C <: A$, $D <: B$, then $C \not\sim D$.
+\end{enumerate}
+
+(1) The proof is by strong induction on the sum of the depths of $A$,
+$B$, $C$, and $D$. To show that $C \sim D$ it suffices to show that
+all of their atoms are consistent. Suppose $C' \in \ATOMS{C}$
+and $D'\in\ATOMS{D}$. So we need to show that $C' \sim D'$.
+%Note that $A <: C'$ and $B <: D'$
+We proceed by cases on $C'$.
+\begin{itemize}
+\item Case $C'=n_1$:\\
+  We have $A <: C'$ and therefore $n_1 \in \ATOMS{A}$.
+  Then because $A \sim B$, we have $\ATOMS{B} \subseteq \{n_1\}$.
+  We have $B <: D'$, so we also have $\ATOMS{D} \subseteq \{n_1\}$.
+  Therefore $C' \sim D'$.
+
+\item Case $C'=C_1\to C_2$:\\
+  We have $A <: C_1 \to C_2$, so by inversion we have
+  some sequence of function types $\Gamma_1$ such that
+  every element of $\Gamma_1$ is an atom of $A$,
+  $C_1 <: \bigsqcap \mathrm{dom}(\Gamma_1)$,
+  and $\bigsqcap \mathrm{cod}(\Gamma_1) <: C_2$.
+
+  We also know that $D'$ is a function type, say $D'=D_1 \to D_2$.
+  (This is because we have $A <: C'$, so we know that $A_1\to A_2 \in
+  \ATOMS{A}$ for some $A_1,A_2$. Then because $A \sim B$, we know that
+  all the atoms in $B$ are function types.  Then because $B <: D$ and
+  $D' \in \ATOMS{D}$, we have that $D'$ is a function type.)
+  %
+  So by inversion on $B <: D_1 \to D_2$, we have
+  some sequence of function types $\Gamma_2$ such that
+  every element of $\Gamma_2$ is an atom of $B$,
+  $D_1 <: \bigsqcap \mathrm{dom}(\Gamma_2)$,
+  and $\bigsqcap \mathrm{cod}(\Gamma_2) <: D_2$.
+
+  It's the case that either $C_1 \sim D_1$ or $C_1 \not\sim D_1$.
+  \begin{itemize}
+  \item Sub-case $C_1 \sim D_1$.\\
+    It suffices to show that $C_2 \sim D_2$.
+    By the induction hypothesis, we have
+    $\bigsqcap \mathrm{dom}(\Gamma_1) \sim \bigsqcap \mathrm{dom}(\Gamma_2)$.
+
+    As an intermediate step, we shall prove that
+    $\bigsqcap \mathrm{cod}(\Gamma_1) \sim \bigsqcap \mathrm{cod}(\Gamma_2)$,
+    which we shall do by showing that all their atoms are consistent.
+    Suppose $A' \in \ATOMS{\bigsqcap \mathrm{cod}(\Gamma_1)}$
+    and $B' \in \ATOMS{\bigsqcap \mathrm{cod}(\Gamma_2)}$.
+    There is some $A_1\to A_2 \in \Gamma_1$ where $A' \in \ATOMS{A_2}$.
+    Similarly, there is $B_1 \to B_2 \in \Gamma_2$ where $B' \in \ATOMS{B_2}$.
+    Also, we have $A_1 \to A_2 \in \ATOMS{A}$ and $B_1 \to B_2 \in \ATOMS{B}$.
+    Then because $A \sim B$, we have $A_1 \to A_2 \sim B_1 \to B_2$.
+    Furthermore, we have $A_1 \sim B_1$ because
+    $\bigsqcap \mathrm{dom}(\Gamma_1) \sim \bigsqcap \mathrm{dom}(\Gamma_2)$,
+    so it must be the case that $A_2 \sim B_2$.
+    Then because $A' \in \ATOMS{A_2}$ and $B' \in \ATOMS{B_2}$, we
+    have $A' \sim B'$. Thus concludes this intermediate step.
+
+    By another use of the induction hypothesis, we have
+    $C_2 \sim D_2$, and this case is finished.
+
+  \item Sub-case $C_1 \not\sim D_1$.\\
+    Then we immediately have $C_1 \to C_2 \sim D_1 \to D_2$.
+  \end{itemize}
+
+\item Case $C'=C_1\sqcap C_2$:\\
+  We already know that $C'$ is an atom, so we have a contradiction
+  and this case is vacously true.
+\end{itemize}
+
+Now we can move on to lemmas about the denotational semantics that
+will be needed to prove the Join Theorem. \\
+
+\noindent \textbf{Lemma} (Weakening)
+\begin{enumerate}
+\item If $\FSEM{A}e \rho_1$, $\rho_1 <: \rho_2$ and
+  $(\forall B \rho_1 \rho_2,\; B \in \ESEM{e}\rho_1, \rho_2 <: \rho_1
+  \Rightarrow B \in \ESEM{e}\rho_2)$, then
+  $\FSEM{A}e \rho_2$.
+\item If $A \in \ESEM{e}\rho_1$ and $\rho_2 <: \rho_1$,
+  then $A \in \ESEM{e}\rho_2$.
+\end{enumerate}
+(Part 1 is proved by induction on $A$. Part 2 is proved by induction
+on $e$ and uses part 1.)
+
+\noindent \textbf{Lemma} (Consistent Application) \\
+%
+If $A_1 \sim A_2$, $B_1 \sim B_2$, $A_1 <: B_1 \to C_1$,
+$A_2 <: B_2 \to C_2$, and all these types are well formed,
+then $C_1 \sim C_2$.
 

### 2018-08-09 DenotSem/determ.tex
check
longer proof for var case

@@
 We shall represent variables with de Bruijn indices, so an environment
-$\rho$ is a sequence of types.  The denotational semantics of the CBV
+$\Gamma$ is a sequence of types.  The denotational semantics of the CBV
 $\lambda$-calculus is defined as follows.
 \begin{align*}
-  \ESEM{n}\rho &= \UP{n} \\
-  \ESEM{x}\rho &=  \IF x < |\rho| \THEN \UP{\rho_k} \ELSE \emptyset \\
-  \ESEM{\lambda e }\rho &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\rho) \} \\
-  \ESEM{e_1\app e_2}\rho &= \left\{ C\, \middle| \exists A,B.\;
+  \ESEM{n}\Gamma &= \UP{n} \\
+  \ESEM{x}\Gamma &=  \IF x < |\Gamma| \THEN \UP{\Gamma[k]} \ELSE \emptyset \\
+  \ESEM{\lambda e }\Gamma &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\Gamma) \} \\
+  \ESEM{e_1\app e_2}\Gamma &= \left\{ C\, \middle| \exists A,B.\;
       \begin{array}{l}
-      A \in \ESEM{e_1}\rho,
-      B \in \ESEM{e_2}\rho,\\
+      A \in \ESEM{e_1}\Gamma,
+      B \in \ESEM{e_2}\Gamma,\\
       A <: B \to C, \text{ and } \WF{C}
@@
         \right\} \\
-  \ESEM{f(e_1,e_2)}\rho &=
+  \ESEM{f(e_1,e_2)}\Gamma &=
       \left\{ C\, \middle| \begin{array}{l}
-       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\rho, B \in \ESEM{e_2}\rho,\\
+       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\Gamma, B \in \ESEM{e_2}\Gamma,\\
       \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
       \end{array} \right\} \\
-  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\rho &=
+  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\Gamma &=
     \left\{ B\, \middle| \exists A.\;
-    \begin{array}{l}A \in \ESEM{e_1}\rho, A <: n,\\
-           n = 0 \Rightarrow B \in \ESEM{e_3}\rho,\\
-           n \neq 0 \Rightarrow B \in \ESEM{e_2}\rho
+    \begin{array}{l}A \in \ESEM{e_1}\Gamma, A <: n,\\
+           n = 0 \Rightarrow B \in \ESEM{e_3}\Gamma,\\
+           n \neq 0 \Rightarrow B \in \ESEM{e_2}\Gamma
     \end{array}
@@
     \intertext{where}
-   \FSEM{n}e\rho &= \mathit{false} \\
-   \FSEM{A \sqcap B}e \rho &= \FSEM{A}e\rho \text{ and } \FSEM{B}e\rho\\
-   \FSEM{A \to B}e \rho &= B \in \ESEM{e} (A, \rho)
+   \FSEM{n}e\Gamma &= \mathit{false} \\
+   \FSEM{A \sqcap B}e \Gamma &= \FSEM{A}e\Gamma \text{ and } \FSEM{B}e\Gamma\\
+   \FSEM{A \to B}e \Gamma &= B \in \ESEM{e} (A, \Gamma)
 \end{align*}
@@
 
-Now we can move on to lemmas about the denotational semantics that
-will be needed to prove the Join Theorem. \\
+The next two lemmas follow from the Consistency and Subtyping Lemma
+and help prepare to prove the case for application in the Join
+Theorem.
+
+\noindent \textbf{Lemma} (Application Consistency) \\
+%
+If $A_1 \sim A_2$, $B_1 \sim B_2$, $A_1 <: B_1 \to C_1$,
+$A_2 <: B_2 \to C_2$, and all these types are well formed,
+then $C_1 \sim C_2$.\\
+(This lemma is proved directly, without induction.)\\
+
+\noindent \textbf{Lemma} (Application Intersection) \\
+%
+If $A_1 <: B_1 \to C_1$, $A_2 <: B_2 \to C_2$, $A_1 \sim A_2$,
+$B_1 \sim B_2$, and $C_1 \sim C_2$, then
+$(A_1\sqcap A_2) <: (B_1 \sqcap B_2) \to (C_1 \sqcap C_2)$.\\
+(This lemma is proved directly, without induction.) \\
+
+We move on to the Weakening Lemma, which will be needed to prove the
+Join Theorem. \\
 
@@
 \begin{enumerate}
-\item If $\FSEM{A}e \rho_1$, $\rho_1 <: \rho_2$ and
-  $(\forall B \rho_1 \rho_2,\; B \in \ESEM{e}\rho_1, \rho_2 <: \rho_1
-  \Rightarrow B \in \ESEM{e}\rho_2)$, then
-  $\FSEM{A}e \rho_2$.
-\item If $A \in \ESEM{e}\rho_1$ and $\rho_2 <: \rho_1$,
-  then $A \in \ESEM{e}\rho_2$.
+\item If $\FSEM{A}e \Gamma_1$, $\Gamma_1 <: \Gamma_2$ and
+  $(\forall B \Gamma_1 \Gamma_2,\; B \in \ESEM{e}\Gamma_1, \Gamma_2 <: \Gamma_1
+  \Rightarrow B \in \ESEM{e}\Gamma_2)$, then
+  $\FSEM{A}e \Gamma_2$.
+\item If $A \in \ESEM{e}\Gamma_1$ and $\Gamma_2 <: \Gamma_1$,
+  then $A \in \ESEM{e}\Gamma_2$.
 \end{enumerate}
 (Part 1 is proved by induction on $A$. Part 2 is proved by induction
-on $e$ and uses part 1.)
+on $e$ and uses part 1.) \\
 
-\noindent \textbf{Lemma} (Consistent Application) \\
+Now for the main event, the proof of the Join Theorem!
+Though we'll rebrand it as the Meet Theorem.\\
+
+\noindent \textbf{Theorem} (Meet)\\
+If $A_1 \in \ESEM{e}\Gamma_1$, $A_2 \in \ESEM{e}\Gamma_2$,
+both $\Gamma_1$ and $\Gamma_2$ are well formed,
+and $\Gamma_1 \sim \Gamma_2$, \\
+then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$. \\
 %
-If $A_1 \sim A_2$, $B_1 \sim B_2$, $A_1 <: B_1 \to C_1$,
-$A_2 <: B_2 \to C_2$, and all these types are well formed,
-then $C_1 \sim C_2$.
+\textbf{Proof}
+We proceed by induction on $e$.
+\begin{itemize}
+\item Case $e=x$:
+  We have $\Gamma_1[k] <: A_1$ and $\Gamma_2[k] <: A_2$.
+
+  nts. $\Gamma_1[k] \sqcap \Gamma_2[k] <: A_1 \sqcap A_2$
+\end{itemize}
 

### 2018-08-09 DenotSem/determ.tex
longer proof for nat case

@@
 
-\section{The Home Stretch}
-
-Armed with a good definition of intersection types, I turned back to
-the proof of the Join Theorem, but first I needed to adjust my
-denotational semantics to use intersection types. For this we'll need
-the definition of well formed types that we alluded to earlier.
-
-\begin{gather*}
-  \frac{}{\WF{n}}
-  \qquad
-  \frac{\WF{A} \quad \WF{B}}{\WF{A \to B}}
-  \qquad
-  \frac{A \sim B \quad \WF{A} \quad \WF{B}}{\WF{A \sqcap B}}
-\end{gather*}
-
-Here are some examples and non-examples of well-formed types.
-\begin{gather*}
-  \WF{4} \qquad \WF{3 \sqcap 3} \qquad \neg \WF{3 \sqcap 4} \\
-  \WF{(0\to 1) \sqcap (2 \to 3)} \qquad \neg \WF{(0 \to 1) \sqcap (0 \to 2)}
-\end{gather*}
-It is sometimes helpful to think of well-formed types in terms of the
-equivalence classes determined by subtype equivalence:
+We define subtyping as follows.
 \[
-A \approx B \quad = \quad
-  A <: B \text{ and } B <: A
-\]
-For example, we have $3 \approx (3 \sqcap 3)$, so they are
-in the same equivalence class and $3$ would be the representative.
-
-We also introduce the following notation for all the well-formed types
-that are super-types of a given type.
-\[
-   \UP{A} \quad = \quad \{ B\mid A <: B \text{ and } \WF{B} \}
+  A <: B \quad = \quad A \vdash B
 \]
 
-We shall represent variables with de Bruijn indices, so an environment
-$\Gamma$ is a sequence of types.  The denotational semantics of the CBV
-$\lambda$-calculus is defined as follows.
-\begin{align*}
-  \ESEM{n}\Gamma &= \UP{n} \\
-  \ESEM{x}\Gamma &=  \IF x < |\Gamma| \THEN \UP{\Gamma[k]} \ELSE \emptyset \\
-  \ESEM{\lambda e }\Gamma &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\Gamma) \} \\
-  \ESEM{e_1\app e_2}\Gamma &= \left\{ C\, \middle| \exists A,B.\;
-      \begin{array}{l}
-      A \in \ESEM{e_1}\Gamma,
-      B \in \ESEM{e_2}\Gamma,\\
-      A <: B \to C, \text{ and } \WF{C}
-      \end{array}
-        \right\} \\
-  \ESEM{f(e_1,e_2)}\Gamma &=
-      \left\{ C\, \middle| \begin{array}{l}
-       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\Gamma, B \in \ESEM{e_2}\Gamma,\\
-      \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
-      \end{array} \right\} \\
-  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\Gamma &=
-    \left\{ B\, \middle| \exists A.\;
-    \begin{array}{l}A \in \ESEM{e_1}\Gamma, A <: n,\\
-           n = 0 \Rightarrow B \in \ESEM{e_3}\Gamma,\\
-           n \neq 0 \Rightarrow B \in \ESEM{e_2}\Gamma
-    \end{array}
-    \right\}
-    \\[2ex]
-    \intertext{where}
-   \FSEM{n}e\Gamma &= \mathit{false} \\
-   \FSEM{A \sqcap B}e \Gamma &= \FSEM{A}e\Gamma \text{ and } \FSEM{B}e\Gamma\\
-   \FSEM{A \to B}e \Gamma &= B \in \ESEM{e} (A, \Gamma)
-\end{align*}
+The BCD subtyping rules and other derived rules follow from the above
+lemmas.\\
 
-\noindent \textbf{Proposition} (Consistency is Symmetric and Reflexive) \\
+\noindent \textbf{Proposition} (Properties of Subtyping) 
+\begin{enumerate}
+\item $A <: A$.
+\item If $A <: B$ and $B <: C$, then $A <: C$. 
+\item If $C <: A$ and $B <: D$, then $A \to B <: C \to D$.
+\item If $A_1 <: B$, then $A_1 \sqcap A_2 <: B$.
+\item If $A_2 <: B$, then $A_1 \sqcap A_2 <: B$.
+\item If $B <: A_1$ and $B <: A_2$, then $B <: A_1 \sqcap A_2$.
+\item If $A <: C$ and $B <: D$, then $A \sqcap B <: C \sqcap D$.
+\item $(A\to B) \sqcap (A \to C) <: A \to (B \sqcap C)$.
+\item $(A \to C) \sqcap (B \to D) <: (A\sqcap B) \to (C \sqcap D)$
+\end{enumerate}
+
+\section{Consistency and Subtyping, Resolved}
+
+Recall that my switch to intersection types was motivated by my
+failure to prove the Consistency and Subtyping Lemma.  We now return
+to the proof of that Lemma. We start with a handful of lemmas that are
+needed for that proof. \\
+
+\noindent \textbf{Lemma} (Consistency is Symmetric and Reflexive) 
 %
@@
 
-\noindent \textbf{Lemma} (Subtyping Inversion for Funcion Types)\\
+\noindent \textbf{Lemma} (Subtyping Inversion for Function Types)\\
 %
@@
 
-With the above lemmas, we are ready to prove the Consistency and
-Subtyping Lemma that previously stumped me.\\
-
 \noindent \textbf{Lemma} (Consistency and Subtyping)
@@
 (1) The proof is by strong induction on the sum of the depths of $A$,
-$B$, $C$, and $D$. To show that $C \sim D$ it suffices to show that
+$B$, $C$, and $D$. We define the depth of a type as follows.
+\begin{align*}
+  \mathit{depth}(n) &= 0 \\
+  \mathit{depth}(A \to B) &= 1 + \mathrm{max}(\mathit{depth}(A),\mathit{depth}(B)) \\
+  \mathit{depth}(A \sqcap B) &= \mathrm{max}(\mathit{depth}(A),\mathit{depth}(B)) 
+\end{align*}
+To show that $C \sim D$ it suffices to show that
 all of their atoms are consistent. Suppose $C' \in \ATOMS{C}$
@@
 and help prepare to prove the case for application in the Join
-Theorem.
+Theorem. \\
 
@@
 
-We move on to the Weakening Lemma, which will be needed to prove the
-Join Theorem. \\
+\section{Updating the Denotational Semantics}
+
+Armed with the Consistency and Subtyping Lemma, I turned back to the
+proof of the Join Theorem, but first I needed to update my
+denotational semantics to use intersection types instead of
+values. For this we'll need the definition of well formed types that
+we alluded to earlier.
+
+\begin{gather*}
+  \frac{}{\WF{n}}
+  \qquad
+  \frac{\WF{A} \quad \WF{B}}{\WF{A \to B}}
+  \qquad
+  \frac{A \sim B \quad \WF{A} \quad \WF{B}}{\WF{A \sqcap B}}
+\end{gather*}
+
+Here are some examples and non-examples of well-formed types.
+\begin{gather*}
+  \WF{4} \qquad \WF{3 \sqcap 3} \qquad \neg \WF{3 \sqcap 4} \\
+  \WF{(0\to 1) \sqcap (2 \to 3)} \qquad \neg \WF{(0 \to 1) \sqcap (0 \to 2)}
+\end{gather*}
+It is sometimes helpful to think of well-formed types in terms of the
+equivalence classes determined by subtype equivalence:
+\[
+A \approx B \quad = \quad
+  A <: B \text{ and } B <: A
+\]
+For example, we have $3 \approx (3 \sqcap 3)$, so they are
+in the same equivalence class and $3$ would be the representative.
+
+We also introduce the following notation for all the well-formed types
+that are super-types of a given type.
+\[
+   \UP{A} \quad = \quad \{ B\mid A <: B \text{ and } \WF{B} \}
+\]
+
+We shall represent variables with de Bruijn indices, so an environment
+$\Gamma$ is a sequence of types.  The denotational semantics of the CBV
+$\lambda$-calculus is defined as follows.
+\begin{align*}
+  \ESEM{n}\Gamma &= \UP{n} \\
+  \ESEM{x}\Gamma &=  \IF x < |\Gamma| \THEN \UP{\Gamma[k]} \ELSE \emptyset \\
+  \ESEM{\lambda e }\Gamma &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\Gamma) \} \\
+  \ESEM{e_1\app e_2}\Gamma &= \left\{ C\, \middle| \exists A,B.\;
+      \begin{array}{l}
+      A \in \ESEM{e_1}\Gamma,
+      B \in \ESEM{e_2}\Gamma,\\
+      A <: B \to C, \text{ and } \WF{C}
+      \end{array}
+        \right\} \\
+  \ESEM{f(e_1,e_2)}\Gamma &=
+      \left\{ C\, \middle| \begin{array}{l}
+       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\Gamma, B \in \ESEM{e_2}\Gamma,\\
+      \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
+      \end{array} \right\} \\
+  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\Gamma &=
+    \left\{ B\, \middle| \exists A.\;
+    \begin{array}{l}A \in \ESEM{e_1}\Gamma, A <: n,\\
+           n = 0 \Rightarrow B \in \ESEM{e_3}\Gamma,\\
+           n \neq 0 \Rightarrow B \in \ESEM{e_2}\Gamma
+    \end{array}
+    \right\}
+    \\[2ex]
+    \intertext{where}
+   \FSEM{n}e\Gamma &= \mathit{false} \\
+   \FSEM{A \sqcap B}e \Gamma &= \FSEM{A}e\Gamma \text{ and } \FSEM{B}e\Gamma\\
+   \FSEM{A \to B}e \Gamma &= B \in \ESEM{e} (A, \Gamma)
+\end{align*}
+
+It is easy to show that swapping in a ``super'' environment does not
+change the semantics.
 
@@
 
-Now for the main event, the proof of the Join Theorem!
-Though we'll rebrand it as the Meet Theorem.\\
+\section{The Home Stretch}
+
+Now for the main event, the proof of the Join Theorem!  Though we'll
+rebrand it as the Meet Theorem, given our switch from values to
+intersection types.\\
 
@@
 and $\Gamma_1 \sim \Gamma_2$, \\
-then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$. \\
+then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$. \\[1ex]
 %
@@
 \begin{itemize}
-\item Case $e=x$:
-  We have $\Gamma_1[k] <: A_1$ and $\Gamma_2[k] <: A_2$.
+\item Case $e=k$ ($k$ is a de Bruijn index for a variable):\\
+  We have $\Gamma_1[k] <: A_1$ and $\Gamma_2[k] <: A_2$,
+  so $\Gamma_1[k] \sqcap \Gamma_2[k] <: A_1 \sqcap A_2$.
+  Also, because $\Gamma_1 \sim \Gamma_2$ we have
+  $\Gamma_1[k] \sim \Gamma_2[k]$ and therefore
+  $A_1 \sim A_2$, by the Consistency and Subtyping Lemma.
+  So we have $\WF{A_1 \sqcap A_2}$ and this case is finished.
 
-  nts. $\Gamma_1[k] \sqcap \Gamma_2[k] <: A_1 \sqcap A_2$
+\item Case $e=n$: \\
+  We have $n <: A_1$ and $n <: A_2$, so $n <: A_1 \sqcap A_2$.
+  Also, we have $A_1 \sim A_2$ by the Consistency and Subtyping Lemma.
+  So we have $\WF{A_1 \sqcap A_2}$ and this case is finished.
+
+\item Case $e=\lambda e$: \\
+
 \end{itemize}

### 2018-08-09 DenotSem/determ.tex
check

@@
 
-I am particularly interested in this theorem because $\beta$-equality
-can be obtained as a corollary.
+\noindent I am particularly interested in this theorem because
+$\beta$-equality can be obtained as a corollary.
 \[
@@
 Turning back to the Join Theorem, I restated it in terms of the
-intersection type system. Instead of using the letter $\rho$ for
-environments, we shall switch to $\Gamma$ because they now contain
-types instead of values. \\
+intersection type system and rebranded it the Meet Theorem. Instead of
+using the letter $\rho$ for environments, we shall switch to $\Gamma$
+because they now contain types instead of values. \\
 
-\noindent \textbf{Theorem} (Intersection Introduction)\\
+\noindent \textbf{Theorem} (Meet)\\
   If $\Gamma \vdash e : A$, $\Gamma' \vdash e : B$, and $\Gamma \sim
@@
   \ESEM{x}\Gamma &=  \IF x < |\Gamma| \THEN \UP{\Gamma[k]} \ELSE \emptyset \\
-  \ESEM{\lambda e }\Gamma &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\Gamma) \} \\
+  \ESEM{\lambda e }\Gamma &= \{ A \mid \WF{A} \text{ and } \FSEM{A}e\Gamma \} \\
   \ESEM{e_1\app e_2}\Gamma &= \left\{ C\, \middle| \exists A,B.\;
@@
 
-Now for the main event, the proof of the Join Theorem!  Though we'll
-rebrand it as the Meet Theorem, given our switch from values to
-intersection types.\\
+Now for the main event, the proof of the Meet Theorem! \\
 
@@
 and $\Gamma_1 \sim \Gamma_2$, \\
-then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$. \\[1ex]
+then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$ and
+  $\WF{A_1 \sqcap A_2}$. \\[1ex]
 %
@@
 \item Case $e=\lambda e$: \\
+  We need to show that $\WF{A_1 \sqcap A_2}$ and
+  $\FSEM{A_1 \sqcap A_2}e(\Gamma_1\sqcap\Gamma_2)$.
+  For the later, it suffices to show that $A_1 \sim A_2$,
+  which we shall do by showing that their atoms are consistent.
+  Suppose $A'_1 \in \ATOMS{A_1}$ and $A'_2 \in \ATOMS{A_2}$.
+  Because $\FSEM{A_1}e\Gamma_1$ 
+  we have $A'_1 =A'_{11} \to A'_{12}$ and
+  $A'_{12} \in \ESEM{e}(A'_{11},\Gamma_1)$. Similarly,
+  from $\FSEM{A_2}e\Gamma_2$ we have
+  $A'_2 =A'_{21} \to A'_{22}$ and $A'_{22} \in \ESEM{e}(A'_{21},\Gamma_2)$.
+  We proceed by cases on whether $A'_{11} \sim A'_{21}$.
+  \begin{itemize}
+  \item Sub-case $A'_{11} \sim A'_{21}$:\\
+    By the induction hypothesis, we have
+    $\WF{A'_{12} \sqcap A'_{22}}$ from which we
+    have $A'_{12} \sim A'_{22}$ and therefore
+    $A'_{11}\to A'_{12} \sim A'_{21} \to A'_{22}$.
+
+    %% \[
+    %%   A'_{12} \sqcap A'_{22} \in
+    %%   \ESEM{e}(A'_{11}\sqcap A'_{21},\Gamma_1 \sqcap \Gamma_2)
+    %%   \quad\text{and} \quad
+    %%   \WF{A'_{12} \sqcap A'_{22}}
+    %% \]
+
+  \item Sub-case $A'_{11} \not\sim A'_{21}$:\\
+      It immediately follows that $A'_{11}\to A'_{12} \sim A'_{21} \to A'_{22}$.
+  \end{itemize}
+
+  It remains to show $\FSEM{A_1 \sqcap A_2}e(\Gamma_1\sqcap\Gamma_2)$.
+  This follows from two uses of the Weakening Lemma to obtain
+  $\FSEM{A_1}e(\Gamma_1\sqcap\Gamma_2)$ and
+  $\FSEM{A_2}e(\Gamma_1\sqcap\Gamma_2)$.
+
+\item Case $e = (e_1 \app e_2)$: \\
+  We have
+  \[
+  B_1 \in \ESEM{e_1}\Gamma_1 \quad
+  C_1 \in \ESEM{e_2}\Gamma_1 \quad
+  B_1 <: C_1 \to A_1 \quad
+  \WF{A_1}
+  \]
+  and
+  \[
+  B_2 \in \ESEM{e_1}\Gamma_2 \quad
+  C_2 \in \ESEM{e_2}\Gamma_2 \quad
+  B_2 <: C_2 \to A_2 \quad
+  \WF{A_2}
+  \]
+  By the induction hypothesis, we have
+  \[
+  B_1 \sqcap B_2 \in \ESEM{e_1}(\Gamma_1 \sqcap \Gamma_2)
+  \quad
+  \WF{B_1 \sqcap B_2}
+  \]
+  and
+  \[
+  C_1 \sqcap C_2 \in \ESEM{e_2}(\Gamma_1 \sqcap \Gamma_2)
+  \quad
+  \WF{C_1 \sqcap C_2}
+  \]
+  We obtain $A_1 \sim A_2$ by the Application Consistency Lemma,
+  and then by the Application Intersection Lemma we have
+  \[
+    B_1 \sqcap B_2 <: (C_1 \sqcap C_2) \to (A_1 \sqcap A_2)
+  \]
+  So we have $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1 \sqcap \Gamma_2)$.
+
+  Also, from $A_1 \sim A_2$, $\WF{A_1}$, and $\WF{A_2}$, we conclude
+  that $\WF{A_1 \sqcap A_2}$.
+\item Case $e= f(e_1,e_2)$: \\
+  (This case is not very interesting. See the Isabelle proof
+   for the details.)
+
+ \item Case $e= \IF e_1 \THEN e_2 \ELSE e_3$: \\
+  (This case is not very interesting. See the Isabelle proof
+   for the details.)
 
 \end{itemize}
+
+I thought that the following Subsumption Theorem would be needed to
+prove the Meet Theorem, but it turned out not to be necessary, which
+is especially nice because the proof of the Subsumption Theorem turned
+out to depend on the Meet Theorem!\\
+
+\noindent \textbf{Theorem} (Subsumption)\\
+If $A \in \ESEM{e}\Gamma$, $A <: B$, and both $B$ and $\Gamma$ are
+well-formed, then $B \in \ESEM{e}\Gamma$. \\
+
+\noindent The proof is by induction on $e$ and all but the case
+$e=\lambda e'$ are straightforward. For that case, we use the
+following lemmas.\\
+
+\noindent \textbf{Lemma} (Distributivity for $\mathcal{F}$)\\
+If $\FSEM{(A \to B)\sqcap (C \to D)} e \Gamma$,
+$A \sim C$, and everything is well formed, then
+$\FSEM{(A\sqcap C) \to (B\sqcap D)} e \Gamma$.\\
+(The proof is direct, using the Meet Theorem
+and the Weakening Lemma.)\\
+
+\noindent \textbf{Lemma} ($\mathcal{F}$ and Intersections)\\
+Suppose $\Gamma_1$ is a non-empty sequence of well-formed and
+consistent function types.
+If $\FSEM{\bigsqcap \Gamma_1} e \Gamma_2$, then
+$\FSEM{\bigsqcap\mathrm{dom}(\Gamma_1) \to \bigsqcap \mathrm{cod}(\Gamma_1)}
+e \Gamma_2$. \\
+(The proof is by induction on $\Gamma_1$ and uses the previous lemma.)
+
+\section{Conclusion}
+
+This result can be viewed a couple ways. As discussed at the beginning
+of this post, establishing the Meet Theorem means that the this
+call-by-value denotational semantics respects $\beta$-equality for any
+terminating argument expression. This is useful in proving the
+correctness of a function inlining optimizer. Also, it would be
+straightforward to define a call-by-name (or need) version of the
+semantics that respects $\beta$-equality unconditionally.
+
+Secondly, from the viewpoint of intersection type systems, this result
+shows that, once we require types to be well formed (i.e. self
+consistent), we no longer need the intersection introduction rule
+because it is a consequence of having the subtyping rule for
+distributing intersections through function types.
 

### 2018-08-09 DenotSem/determ.tex
check

@@
 \newcommand{\ATOMS}[1]{\mathit{atoms}(#1)}
+\newcommand{\sqinter}[0]{\bigsqcap}
 
@@
 hopefully remembering the challenges and motivations.
+Here is a tar ball of the mechanization in Isabelle and in pdf form.
 
@@
 The combined rule for function types says that the intersection of a
-sequence of function types $\bigsqcap_{i=1\ldots n} (C_i\to D_i)$ is a
+sequence of function types $\sqinter_{i=1\ldots n} (C_i\to D_i)$ is a
 subtype of $A \to B$ if
 \[
-  A <: \bigsqcap_{i\in\{1\ldots n\}} C_i
+  A <: \sqinter_{i\in\{1\ldots n\}} C_i
   \qquad \text{and}\qquad
-  \bigsqcap_{i\in\{1\ldots n\}} D_i <: B
+  \sqinter_{i\in\{1\ldots n\}} D_i <: B
 \]
@@
 \item For every $D\to E \in \Gamma$, we have $A <: D$, and
-\item $\bigsqcap \mathrm{cod}(\Gamma) <: B$.
+\item $\sqinter \mathrm{cod}(\Gamma) <: B$.
 \end{enumerate}
-Note that item 2 above implies that $A <: \bigsqcap \mathrm{dom}(\Gamma)$.
+Note that item 2 above implies that $A <: \sqinter \mathrm{dom}(\Gamma)$.
 
@@
   every element of $\Gamma_1$ is an atom of $A$,
-  $C_1 <: \bigsqcap \mathrm{dom}(\Gamma_1)$,
-  and $\bigsqcap \mathrm{cod}(\Gamma_1) <: C_2$.
+  $C_1 <: \sqinter \mathrm{dom}(\Gamma_1)$,
+  and $\sqinter \mathrm{cod}(\Gamma_1) <: C_2$.
 
@@
   every element of $\Gamma_2$ is an atom of $B$,
-  $D_1 <: \bigsqcap \mathrm{dom}(\Gamma_2)$,
-  and $\bigsqcap \mathrm{cod}(\Gamma_2) <: D_2$.
+  $D_1 <: \sqinter \mathrm{dom}(\Gamma_2)$,
+  and $\sqinter \mathrm{cod}(\Gamma_2) <: D_2$.
 
@@
     By the induction hypothesis, we have
-    $\bigsqcap \mathrm{dom}(\Gamma_1) \sim \bigsqcap \mathrm{dom}(\Gamma_2)$.
+    $\sqinter \mathrm{dom}(\Gamma_1) \sim \sqinter \mathrm{dom}(\Gamma_2)$.
 
     As an intermediate step, we shall prove that
-    $\bigsqcap \mathrm{cod}(\Gamma_1) \sim \bigsqcap \mathrm{cod}(\Gamma_2)$,
+    $\sqinter \mathrm{cod}(\Gamma_1) \sim \sqinter \mathrm{cod}(\Gamma_2)$,
     which we shall do by showing that all their atoms are consistent.
-    Suppose $A' \in \ATOMS{\bigsqcap \mathrm{cod}(\Gamma_1)}$
-    and $B' \in \ATOMS{\bigsqcap \mathrm{cod}(\Gamma_2)}$.
+    Suppose $A' \in \ATOMS{\sqinter \mathrm{cod}(\Gamma_1)}$
+    and $B' \in \ATOMS{\sqinter \mathrm{cod}(\Gamma_2)}$.
     There is some $A_1\to A_2 \in \Gamma_1$ where $A' \in \ATOMS{A_2}$.
@@
     Furthermore, we have $A_1 \sim B_1$ because
-    $\bigsqcap \mathrm{dom}(\Gamma_1) \sim \bigsqcap \mathrm{dom}(\Gamma_2)$,
+    $\sqinter \mathrm{dom}(\Gamma_1) \sim \sqinter \mathrm{dom}(\Gamma_2)$,
     so it must be the case that $A_2 \sim B_2$.
@@
     \\[2ex]
-    \intertext{where}
    \FSEM{n}e\Gamma &= \mathit{false} \\
@@
 consistent function types.
-If $\FSEM{\bigsqcap \Gamma_1} e \Gamma_2$, then
-$\FSEM{\bigsqcap\mathrm{dom}(\Gamma_1) \to \bigsqcap \mathrm{cod}(\Gamma_1)}
+If $\FSEM{\sqinter \Gamma_1} e \Gamma_2$, then
+$\FSEM{\sqinter\mathrm{dom}(\Gamma_1) \to \sqinter \mathrm{cod}(\Gamma_1)}
 e \Gamma_2$. \\

### 2018-08-09 DenotSem/determ.tex
check

@@
 But here we have a problem. Given our definition of $\sqcup$ in terms
-of set union, there won't necissarily be a single entry in $f \sqcup
+of set union, there won't necessarily be a single entry in $f \sqcup
 f'$ that combines the information from both $v_3 \mapsto v_4$ and
@@
 systems primarily differ in their definitions of subtyping.  Given the
-above proof attempt, I figured that I would need usual
+above proof attempt, I figured that I would need the usual
 co/contra-variant rule for function types and also the following rule
@@
 \]
-and defined subtyping according to the BCD intersection type system.
+and defined subtyping according to the BCD intersection type system
+(\emph{Lambda Calculus with Types}, Barendregt et al. 2013).
 \begin{gather*}
@@
 Turning back to the Join Theorem, I restated it in terms of the
-intersection type system and rebranded it the Meet Theorem. Instead of
+intersection type system and re-branded it the Meet Theorem. Instead of
 using the letter $\rho$ for environments, we shall switch to $\Gamma$
@@
 directed), reasoning by induction or cases (inversion) is more
-difficult, and it is redundant. Furthermore, the presense of the
-transitivity rule means that subtyping does not satisfy the subformula
-property. This term subformula property comes from logic, and means
+difficult, and it is redundant. Furthermore, the presence of the
+transitivity rule means that subtyping does not satisfy the sub-formula
+property. This term sub-formula property comes from logic, and means
 that a derivation (proof) of a formula only mentions propositions that
@@
 reading the literature. Unfortunately, it turns out that none of the
-published intersection type systems satisfied the subformula property
+published intersection type systems satisfied the sub-formula property
 and vast majority of them included the transitivity rule.  However,
@@
   equivalent to BCD subtyping and satisfy the sub-formula property.  I
-  am currently writting a paper on this but it is not ready yet.''
+  am currently writing a paper on this but it is not ready yet.''
 \end{quote}
@@
 %%  LocalWords:  denotationally ctx subexpression eq denot lrcl env
+%%  LocalWords:  iff Hah pdf subtyping contra distributivity BCD et
+%%  LocalWords:  Barendregt al Fundamenta Informaticae sequent

### 2018-08-09 DenotSem/determ.tex
spell check

@@
 Federated Logic Conference (FLOC). I was planning to attend FLOC
-anyways, for the DOMAINS workshop to celabrate Dana Scott's 85th
+anyways, for the DOMAINS workshop to celebrate Dana Scott's 85th
 birthday.
@@
   We already know that $C'$ is an atom, so we have a contradiction
-  and this case is vacously true.
+  and this case is vacuously true.
 \end{itemize}
@@
 We shall represent variables with de Bruijn indices, so an environment
-$\Gamma$ is a sequence of types.  The denotational semantics of the CBV
-$\lambda$-calculus is defined as follows.
+$\Gamma$ is a sequence of types.  The denotational semantics of the
+call-by-value $\lambda$-calculus is defined as follows.
 \begin{align*}
@@
 %%  LocalWords:  iff Hah pdf subtyping contra distributivity BCD et
-%%  LocalWords:  Barendregt al Fundamenta Informaticae sequent
+%%  LocalWords:  Barendregt al Fundamenta Informaticae sequent Coq th
+%%  LocalWords:  FLOC Scott's Olivier's disjunction reflexivity dom
+%%  LocalWords:  subtype Denotational denotational de Bruijn indices
+%%  LocalWords:  Subsumption inlining

### 2018-08-10 DenotSem/determ.tex
check

@@
 
-For a long time I've disliked definition of subtyping in which
+For a long time I've disliked definitions of subtyping in which
 transitivity is given as a rule instead of proved as a theorem.  There
@@
 For most type systems, proving the transitivity of subtyping is
-straightforward. But what I soon realized that the addition of the
+straightforward. But I soon realized that the addition of the
 distributivity rule makes it significantly more difficult.  After
@@
   \ESEM{\lambda e }\Gamma &= \{ A \mid \WF{A} \text{ and } \FSEM{A}e\Gamma \} \\
-  \ESEM{e_1\app e_2}\Gamma &= \left\{ C\, \middle| \exists A,B.\;
+  \ESEM{e_1\app e_2}\Gamma &= \left\{ C\, \middle| 
       \begin{array}{l}
-      A \in \ESEM{e_1}\Gamma,
+      \exists A,B.\; A \in \ESEM{e_1}\Gamma,
       B \in \ESEM{e_2}\Gamma,\\
@@
        \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\Gamma, B \in \ESEM{e_2}\Gamma,\\
-      \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
+      A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
       \end{array} \right\} \\
   \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\Gamma &=
-    \left\{ B\, \middle| \exists A.\;
-    \begin{array}{l}A \in \ESEM{e_1}\Gamma, A <: n,\\
+    \left\{ B\, \middle| 
+    \begin{array}{l}\exists A, n.\; A \in \ESEM{e_1}\Gamma, A <: n,\\
            n = 0 \Rightarrow B \in \ESEM{e_3}\Gamma,\\
@@
 \item If $\FSEM{A}e \Gamma_1$, $\Gamma_1 <: \Gamma_2$ and
-  $(\forall B \Gamma_1 \Gamma_2,\; B \in \ESEM{e}\Gamma_1, \Gamma_2 <: \Gamma_1
+  $(\forall B, \Gamma_1, \Gamma_2.\;
+     B \in \ESEM{e}\Gamma_1, \Gamma_2 <: \Gamma_1
     \Rightarrow B \in \ESEM{e}\Gamma_2)$, then

### 2018-08-10 DenotSem/determ.tex
check

@@
 without transitivity and instead proved it, but his system still did
-not satisfy the sub-formula property because of an addition rule for
+not satisfy the sub-formula property because of an additional rule for
 function types. Nevertheless, Olivier indicated that he was interested

### 2019-06-20 papers/intersection-transitivity.tex
check

\documentclass{article}
\usepackage{natbib}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{mathabx}
\usepackage{stmaryrd}
\usepackage{semantic}
%\usepackage{fullpage}
\usepackage{fontspec,unicode-math}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{constraint}[theorem]{Constraint}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\title{Transitivity of Subtyping for Intersection Types, Directly}
\author{Jeremy G. Siek}

\begin{document}
\maketitle

\newcommand{\TOP}{\ensuremath{\mathtt{U}}}
\newcommand{\dom}[1]{\mathrm{dom}(#1)}
\newcommand{\cod}[1]{\mathrm{cod}(#1)}

\citet{Barendregt:1983aa}
\citet{Barendregt:2013aa}

\citet{Laurent:2012aa}
\citet{Laurent:2018aa}

$λ^{\mathrm{BCD}}_∩$

Figure~\ref{fig:types-subtyping}

\begin{figure}[tbp]
  \[
  \begin{array}{lclr}
    \alpha,\beta & ::= & \TOP \mid c_0 \mid c_1 \mid c_2 \mid \cdots & \text{atoms}\\
    A,B,C,D & ::= & \alpha \mid A → B \mid A ∩ B & \text{types}
  \end{array}
  \]
  \fbox{$A ≤ B$} \\[1ex]
  \begin{tabular}{ll}
    (refl)  & $A ≤ A$ \\[1ex]
    (incl$_L$) & $A ∩ B ≤ A$ \\[1ex]
    (incl$_R$) & $A ∩ B ≤ B$ \\[1ex]
    (glb) & $\inference{A ≤ B & B ≤ C}{A ≤ C}$ \\[1ex]
    ($\TOP_{\mathrm{top}}$) & $A ≤ \TOP$ \\[1ex]
    ($\TOP→$) & $\TOP ≤ A → \TOP$ \\[1ex]
    ($→∩$) & $(A → B) ∩ (A → C) A → (B ∩ C)$ \\[1ex]
    ($→$) & $\inference{A' ≤ A & B ≤ B'}{A → B ≤ A' → B'}$
  \end{tabular}
  \caption{Intersection Types and BCD Subtyping}
  \label{fig:types-subtyping}
\end{figure}

\begin{align*}
  \dom{A → B} &= A \\
  \dom{A ∩ B} &= \dom{A} ∩ \dom {B} \\
  \\
  \cod{A → B} &= B \\
  \cod{A ∩ B} &= \cod{A} ∩ \cod {B}
\end{align*}

\begin{figure}
  \begin{tabular}{ll}
    (refl$_α$) & $α <: α$ \\[1ex]
    (incl$_L$) & $A ∩ B <: A$ \\[1ex]
    (incl$_R$) & $A ∩ B <: B$ \\[1ex]
    (glb) & $\inference{A <: B & B <: C}{A <: C}$ \\[1ex]
    ($\TOP_{\mathrm{top}}$) & $A <: \TOP$ \\[1ex]
    ($\TOP→$) & $\TOP <: A → \TOP$ \\[1ex]
    ($→$) &$\inference{D ⊆ C & A' <: \dom{D} & \cod{D} <: B' }{C <: A' → B'}$
  \end{tabular}
  \caption{New Subtyping Relation}
  \label{fig:new-subtyping}
\end{figure}

\clearpage
\pagebreak

\bibliographystyle{abbrvnat}
\bibliography{all}

\end{document}

### 2019-06-20 papers/intersection-transitivity.tex
check

@@
   \begin{tabular}{ll}
-    (refl)  & $A ≤ A$ \\[1ex]
-    (incl$_L$) & $A ∩ B ≤ A$ \\[1ex]
-    (incl$_R$) & $A ∩ B ≤ B$ \\[1ex]
-    (glb) & $\inference{A ≤ B & B ≤ C}{A ≤ C}$ \\[1ex]
-    ($\TOP_{\mathrm{top}}$) & $A ≤ \TOP$ \\[1ex]
-    ($\TOP→$) & $\TOP ≤ A → \TOP$ \\[1ex]
-    ($→∩$) & $(A → B) ∩ (A → C) A → (B ∩ C)$ \\[1ex]
+    (refl)  & $A ≤ A$ \\[2ex]
+    (incl$_L$) & $A ∩ B ≤ A$ \\[2ex]
+    (incl$_R$) & $A ∩ B ≤ B$ \\[2ex]
+    (glb) & $\inference{C ≤ A & C ≤ B}{C ≤ A ∩ B}$ \\[2ex]
+    (trans) & $\inference{A ≤ B & B ≤ C}{A ≤ C}$ \\[2ex]
+    ($\TOP_{\mathrm{top}}$) & $A ≤ \TOP$ \\[2ex]
+    ($\TOP→$) & $\TOP ≤ A → \TOP$ \\[2ex]
+    ($→∩$) & $(A → B) ∩ (A → C) A → (B ∩ C)$ \\[2ex]
     ($→$) & $\inference{A' ≤ A & B ≤ B'}{A → B ≤ A' → B'}$
@@
   \begin{tabular}{ll}
-    (refl$_α$) & $α <: α$ \\[1ex]
-    (incl$_L$) & $A ∩ B <: A$ \\[1ex]
-    (incl$_R$) & $A ∩ B <: B$ \\[1ex]
-    (glb) & $\inference{A <: B & B <: C}{A <: C}$ \\[1ex]
-    ($\TOP_{\mathrm{top}}$) & $A <: \TOP$ \\[1ex]
-    ($\TOP→$) & $\TOP <: A → \TOP$ \\[1ex]
+    (refl$_α$) & $α <: α$ \\[2ex]
+    (incl$_L$) & $\inference{A <: C}{A ∩ B <: C}$ \\[2ex]
+    (incl$_R$) & $\inference{B <: C} A ∩ B <: C$ \\[2ex]
+    (glb) & $\inference{C ≤ A & C ≤ B}{C ≤ A ∩ B}$ \\[2ex]
+    ($\TOP_{\mathrm{top}}$) & $A <: \TOP$ \\[2ex]
+    ($\TOP→$) & $\TOP <: A → \TOP$ \\[2ex]
     ($→$) &$\inference{D ⊆ C & A' <: \dom{D} & \cod{D} <: B' }{C <: A' → B'}$

### 2019-06-20 papers/intersection-transitivity.tex
check

@@
     (incl$_L$) & $\inference{A <: C}{A ∩ B <: C}$ \\[2ex]
-    (incl$_R$) & $\inference{B <: C} A ∩ B <: C$ \\[2ex]
+    (incl$_R$) & $\inference{B <: C}{A ∩ B <: C}$ \\[2ex]
     (glb) & $\inference{C ≤ A & C ≤ B}{C ≤ A ∩ B}$ \\[2ex]
@@
     ($\TOP→$) & $\TOP <: A → \TOP$ \\[2ex]
-    ($→$) &$\inference{D ⊆ C & A' <: \dom{D} & \cod{D} <: B' }{C <: A' → B'}$
+    ($→'$) &$\inference{D ⊆ C & A' <: \dom{D} & \cod{D} <: B' }{C <: A' → B'}$
   \end{tabular}

### 2019-06-21 papers/intersection-transitivity.tex
minor improvement

--- 2019-06-20 papers/intersection-transitivity.tex
@@
 
-\title{Transitivity of Subtyping for Intersection Types, Directly}
+\title{Transitivity of Subtyping for Intersection Types}
 \author{Jeremy G. Siek}
@@
 
-Figure~\ref{fig:types-subtyping}
+Figure~\ref{fig:types}
 
@@
   \]
+  \caption{Intersection Types}
+  \label{fig:types}
+\end{figure}
+
+Figure~\ref{fig:BCD-subtyping}
+
+\begin{figure}[tbp]
   \fbox{$A ≤ B$} \\[1ex]
-  \begin{tabular}{ll}
-    (refl)  & $A ≤ A$ \\[1ex]
-    (incl$_L$) & $A ∩ B ≤ A$ \\[1ex]
-    (incl$_R$) & $A ∩ B ≤ B$ \\[1ex]
-    (glb) & $\inference{A ≤ B & B ≤ C}{A ≤ C}$ \\[1ex]
-    ($\TOP_{\mathrm{top}}$) & $A ≤ \TOP$ \\[1ex]
-    ($\TOP→$) & $\TOP ≤ A → \TOP$ \\[1ex]
-    ($→∩$) & $(A → B) ∩ (A → C) A → (B ∩ C)$ \\[1ex]
+  \begin{tabular}{p{1in}l}
+    (refl)  & $A ≤ A$ \\[3ex]
+    (incl$_L$) & $A ∩ B ≤ A$ \\[3ex]
+    (incl$_R$) & $A ∩ B ≤ B$ \\[3ex]
+    (glb) & $\inference{C ≤ A & C ≤ B}{C ≤ A ∩ B}$ \\[3ex]
+    (trans) & $\inference{A ≤ B & B ≤ C}{A ≤ C}$ \\[3ex]
+    ($\TOP_{\mathrm{top}}$) & $A ≤ \TOP$ \\[3ex]
+    ($\TOP{→}$) & $\TOP ≤ A → \TOP$ \\[3ex]
+    (${→}{∩}$) & $(A → B) ∩ (A → C) ≤ A → (B ∩ C)$ \\[3ex]
     ($→$) & $\inference{A' ≤ A & B ≤ B'}{A → B ≤ A' → B'}$
   \end{tabular}
-  \caption{Intersection Types and BCD Subtyping}
-  \label{fig:types-subtyping}
+  \caption{BCD Subtyping}
+  \label{fig:BCD-subtyping}
 \end{figure}
 
+Figure~\ref{fig:aux}
+
+\begin{figure}[tbp]
+
+  \fbox{$\dom{A}, \cod{A}$}
   \begin{align*}
@@
 
-\begin{figure}
+  \fbox{$\mathsf{top}(A)$}
+  \begin{gather*}
+    \mathsf{top}(\TOP)
+    \quad
+    \inference{\mathsf{top}(B)}{\mathsf{top}(A → B)}
+    \quad
+    \inference{\mathsf{top}(A) & \mathsf{top}(B)}{\mathsf{top}(A ∩ B)}
+  \end{gather*}
+
+  \fbox{$A ∈ B$}
+  \begin{gather*}
+    \inference{}{\alpha ∈ \alpha}  \quad
+    \inference{}{A → B ∈ A → B} \\[3ex]
+    \inference{A ∈ B}{A ∈ B ∩ C} \quad
+    \inference{A ∈ C}{A ∈ B ∩ C}
+  \end{gather*}
+
+  \fbox{$A ⊆ B$}
+  \[
+     A ⊆ B = ∀ C.\, C ∈ A \text{ implies } C ∈ B
+  \]
+
+  \fbox{$\mathsf{topCod}(D)$}
+  \[
+  \mathsf{topCod}(D) =
+     \exists A B.\, A → B ∈ D \text{ and } \mathsf{top}(B)  
+  \]
+
+  \caption{Auxilliary definitions}
+  \label{fig:aux}
+\end{figure}
+
+\begin{figure}[tbp]
   \begin{tabular}{ll}
-    (refl$_α$) & $α <: α$ \\[1ex]
-    (incl$_L$) & $A ∩ B <: A$ \\[1ex]
-    (incl$_R$) & $A ∩ B <: B$ \\[1ex]
-    (glb) & $\inference{A <: B & B <: C}{A <: C}$ \\[1ex]
-    ($\TOP_{\mathrm{top}}$) & $A <: \TOP$ \\[1ex]
-    ($\TOP→$) & $\TOP <: A → \TOP$ \\[1ex]
-    ($→$) &$\inference{D ⊆ C & A' <: \dom{D} & \cod{D} <: B' }{C <: A' → B'}$
+    (refl$_α$) & $\inference{}{α <: α}$ \\[3ex]
+    (lb$_L$) & $\inference{A <: C}{A ∩ B <: C}$ \\[3ex]
+    (lb$_R$) & $\inference{B <: C}{A ∩ B <: C}$ \\[3ex]
+    (glb) & $\inference{C ≤ A & C ≤ B}{C ≤ A ∩ B}$ \\[3ex]
+    ($\TOP_{\mathrm{top}}$) & $\inference{}{A <: \TOP}$ \\[3ex]
+    ($\TOP{→}'$) & $\inference{}{C <: A → B}\;\mathsf{top}(B)$ \\[3ex]
+    ($→'$) &$\inference{A <: \dom{D} & \cod{D} <: B }{C <: A → B}\begin{array}{l} \neg\, \mathsf{top}(B) \\ \neg \, \mathsf{topCod}(D) \\ D ⊆ C \end{array}$
   \end{tabular}
@@
 \end{figure}
+
+\begin{definition}[factor]
+
+\end{definition}
+
+\begin{lemma}
+  If
+\begin{itemize}
+\item $\neg\, \mathsf{topCod}(D)$,
+\item $D ⊆ C$, and
+\item for any $A,B$, if $\neg\,\mathsf{top}(B)$ and $A → B ∈ D$,
+     then $A → B$ factors $u_2$ into some type $u_3$,
+\end{itemize}
+then $\dom{D} → \cod{D}$ factors $u_2$ into some type $u_4$.
+\end{lemma}
+
+\begin{align*}
+  \mathsf{size}(\alpha) &= 0 \\
+  \mathsf{size}(A → B) &= 1 + \mathsf{size}(A) + \mathsf{size}(B) \\
+  \mathsf{size}(A ∩ B) &= 1 + \mathsf{size}(A) + \mathsf{size}(B)
+\end{align*}
+
+\begin{align*}
+  \mathsf{depth}(\alpha) &= 0 \\
+  \mathsf{depth}(A → B) &= 1 + \max(\mathsf{depth}(A), \mathsf{depth}(B)) \\
+  \mathsf{depth}(A ∩ B) &= \max(\mathsf{depth}(A), \mathsf{depth}(B))
+\end{align*}
+
+\begin{lemma}[Transitivity of $<:$]
+    If $A <: B$ and $B <: C$, then $A <: C$.
+\end{lemma}
+
+\begin{theorem}[Equivalence of the subtyping relations]\ \\
+  $A <: B$ if and only if $A ≤ B$.
+\end{theorem}
 

### 2019-06-21 papers/intersection-transitivity.tex
check

@@
 
-\begin{definition}[factor]
+\begin{definition}[factors]
+  We say $A → B$ \emph{factors} $C$
+  if there exists some type $C'$ such that
+  $C' ⊆ C$, $\neg\,\mathsf{topCod}(C')$, $A <: \dom{C'}$, and $\cod{C'} <: B$.
+\end{definition}
 
-\end{definition}
+\begin{lemma}
+  If $\neg\,\mathsf{top}(w)$
+  and $u_1$
+\end{lemma}
 
@@
   \item $\neg\, \mathsf{topCod}(D)$,
-\item $D ⊆ C$, and
   \item for any $A,B$, if $\neg\,\mathsf{top}(B)$ and $A → B ∈ D$,
-     then $A → B$ factors $u_2$ into some type $u_3$,
+    then $A → B$ factors $C$,
   \end{itemize}
-then $\dom{D} → \cod{D}$ factors $u_2$ into some type $u_4$.
+  then $\dom{D} → \cod{D}$ factors $C$.
 \end{lemma}

### 2019-06-22 papers/intersection-transitivity.tex
check

@@
   \fbox{$A ≤ B$} \\[1ex]
+
+  \centering
   \begin{tabular}{p{1in}l}
@@
 
-  \fbox{$\mathsf{topCod}(D)$}
+  \fbox{$\mathsf{topInCod}(D)$}
   \[
-  \mathsf{topCod}(D) =
+  \mathsf{topInCod}(D) =
      \exists A B.\, A → B ∈ D \text{ and } \mathsf{top}(B)  
@@
 \begin{figure}[tbp]
-  \begin{tabular}{ll}
+  \fbox{$A <: B$} \\[1ex]
+
+  \centering
+  \begin{tabular}{p{1in}l}
     (refl$_α$) & $\inference{}{α <: α}$ \\[3ex]
@@
     ($\TOP{→}'$) & $\inference{}{C <: A → B}\;\mathsf{top}(B)$ \\[3ex]
-    ($→'$) &$\inference{A <: \dom{D} & \cod{D} <: B }{C <: A → B}\begin{array}{l} \neg\, \mathsf{top}(B) \\ \neg \, \mathsf{topCod}(D) \\ D ⊆ C \end{array}$
+    ($→'$) &$\inference{A <: \dom{D} & \cod{D} <: B }{C <: A → B}\begin{array}{l} \neg\, \mathsf{top}(B) \\ \neg \, \mathsf{topInCod}(D) \\ D ⊆ C \end{array}$
   \end{tabular}
@@
   if there exists some type $C'$ such that
-  $C' ⊆ C$, $\neg\,\mathsf{topCod}(C')$, $A <: \dom{C'}$, and $\cod{C'} <: B$.
+  $C' ⊆ C$, $\neg\,\mathsf{topInCod}(C')$, $A <: \dom{C'}$, and $\cod{C'} <: B$.
 \end{definition}
 
-\begin{lemma}
-  If $\neg\,\mathsf{top}(w)$
-  and $u_1$
+\begin{lemma}\label{lem:sub-fun-inv}
+  If $\neg\,\mathsf{top}(B)$,
+  $D <: C$, and
+  $A → B ∈ C$, then
+  $A → B$ factors $D$.
 \end{lemma}
 
-\begin{lemma}
+\begin{lemma}\label{lem:sub-inv-trans}
   If
   \begin{itemize}
-  \item $\neg\, \mathsf{topCod}(D)$,
+  \item $\neg\, \mathsf{topInCod}(D)$,
   \item for any $A,B$, if $\neg\,\mathsf{top}(B)$ and $A → B ∈ D$,
@@
 
-\begin{lemma}[Transitivity of $<:$]
+\begin{lemma}[Transitivity of $<:$]\label{lem:sub-trans}
     If $A <: B$ and $B <: C$, then $A <: C$.

### 2019-06-22 papers/intersection-transitivity.tex
check

@@
 
-\citet{Barendregt:1983aa}
+\begin{abstract}
+  The subtyping relation for intersection type systems traditionally
+  employs a transitivity rule (Barendregt et al. 1983), which means
+  that the subtyping judgment does not enjoy the subformula property.
+  Laurent develops a sequent-style subtyping judgment, without
+  transitivity, and proves transitivity via a sequence of six lemmas
+  that culminate in cut-elimination (2018). This article presents a
+  subtyping judgment, in regular style, that satisfies the subformula
+  property, and presents a direct proof of transitivity. Borrowing
+  from Laurent's system, the rule for function types is essentially
+  the $\beta$-soundness property.  The main lemma required for the
+  transitivity proof happens to be one that is used to prove the
+  inversion principle for subtyping of function types (Barendregt et
+  al. 2013). However, the choice of induction principle for the proof
+  of transitivity requires some care: we use strong induction on the
+  lexicographical ordering of the sum of the depths of the first and
+  last type followed by the sume of the sizes of the first and second
+  type.
+\end{abstract}
+
+\section{Introduction}
+
+Intersection types were invented by Coppo, Dezani-Ciancaglini, and
+Salle, as a tool for studying normalization in the lambda
+calculus~\citet{Coppo:1979aa}. By varying the subtyping rules and atom
+types, researchers have used intersection type systems to model many
+different
+calculi~\citep{Coppo:1980ab,Coppo:1981aa,Engeler:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Plotkin:1993ab,Honsell:1999aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa}.
+Perhaps the best-known of them is the BCD intersection type system of
+\citet{Barendregt:1983aa}.
+
 \citet{Barendregt:2013aa}
@@
 
-$λ^{\mathrm{BCD}}_∩$
+%$λ^{\mathrm{BCD}}_∩$
 
@@
   \end{tabular}
-  \caption{BCD Subtyping}
+  \caption{The Subtyping Judgement of Barendregt, Coppo, and Dezani (BCD).}
   \label{fig:BCD-subtyping}
@@
 
-  \caption{Auxilliary definitions}
+  \caption{Auxiliary definitions}
   \label{fig:aux}
@@
 \end{document}
+
+% LocalWords:  dom subtyping Barendregt et al subformula sequent aa
+% LocalWords:  BCD lclr refl glb topInCod

### 2019-06-22 papers/intersection-transitivity.tex
check

@@
 Salle, as a tool for studying normalization in the lambda
-calculus~\citet{Coppo:1979aa}. By varying the subtyping rules and atom
+calculus~\citep{Coppo:1979aa}. By varying the subtyping rules and atom
 types, researchers have used intersection type systems to model many

### 2019-06-22 papers/intersection-transitivity.tex
check

@@
   subtyping judgment, in regular style, that satisfies the subformula
-  property, and presents a direct proof of transitivity. Borrowing
-  from Laurent's system, the rule for function types is essentially
-  the $\beta$-soundness property.  The main lemma required for the
-  transitivity proof happens to be one that is used to prove the
-  inversion principle for subtyping of function types (Barendregt et
-  al. 2013). However, the choice of induction principle for the proof
-  of transitivity requires some care: we use strong induction on the
+  property, and gives a direct proof of transitivity. Borrowing from
+  Laurent's system, the rule for function types is essentially the
+  $\beta$-soundness property.  The main lemma required for the
+  transitivity proof is one that has been used to prove the inversion
+  principle for subtyping of function types (Barendregt et
+  al. 2013). The choice of induction principle for the proof of
+  transitivity requires some care: we use strong induction on the
   lexicographical ordering of the sum of the depths of the first and
@@
 Perhaps the best-known of them is the BCD intersection type system of
-\citet{Barendregt:1983aa}.
+\citet{Barendregt:1983aa}. For this article we focus on the BCD
+system, following the presentation of \citet{Barendregt:2013aa}.  We
+conjecture that our results apply to other intersection type systems.
 
-\citet{Barendregt:2013aa}
-
-\citet{Laurent:2012aa}
-\citet{Laurent:2018aa}
-
-%$λ^{\mathrm{BCD}}_∩$
-
-Figure~\ref{fig:types}
+The BCD intersection type systems extends the simply-typed lambda
+calculus with the addition of intersection types, written $A ∩ B$, a
+top type $\TOP$, and an infinite collection of type
+constants. Figure~\ref{fig:types} defines the grammar of types.
 
@@
 
-Figure~\ref{fig:BCD-subtyping}
+The BCD intersection type system includes a subsumption rule which
+states that a term $M$ in environment $\Gamma$ can be given type $B$
+if it has type $A$ and $A$ is a subtype of $B$, written $A ≤ B$.
+\[
+\inference{\Gamma \vdash M : A & A ≤ B}
+          {\Gamma \vdash M : B}
+\]
+Figure~\ref{fig:BCD-subtyping} reviews the BCD rules for subtyping.
+Note that in the (trans) rule, the type $B$ that appears in the
+premises does not appear in the conclusion. Thus, the BCD subtyping
+judgment does not enjoy the subformula property. The subformula
+property is a useful one. For example, the author ...
 
@@
 \end{figure}
+
+\citet{Laurent:2012aa}
+\citet{Laurent:2018aa}
+
+%$λ^{\mathrm{BCD}}_∩$
 

### 2019-06-22 papers/intersection-transitivity.tex
check

@@
   lexicographical ordering of the sum of the depths of the first and
-  last type followed by the sume of the sizes of the first and second
+  last type followed by the sum of the sizes of the first and second
   type.
@@
 premises does not appear in the conclusion. Thus, the BCD subtyping
-judgment does not enjoy the subformula property. The subformula
-property is a useful one. For example, the author ...
+judgment does not enjoy the subformula property.  For other systems,
+it is straightforward to remove the (trans) rule, modify the other
+rules, and then then prove transitivity~\citep{Muehlboeck:2018aa}.
+Unfortunately, the ${→}{∩}$ rule significantly complicates the
+situation for the BCD system.
 
@@
   \end{tabular}
-  \caption{The Subtyping Judgement of Barendregt, Coppo, and Dezani (BCD).}
+  \caption{The Subtyping Judgment of Barendregt, Coppo, and Dezani (BCD).}
   \label{fig:BCD-subtyping}
@@
 
-\citet{Laurent:2012aa}
-\citet{Laurent:2018aa}
+The subformula property is a useful one. For example, the author is
+using intersection types to create a denotational semantics for the
+ISWIM language, which includes constants and primitive
+operations~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa}.
+It seems that doing so requires placing extra conditions on types and
+that is much easier when subtyping satisfies the subformula property.
+
+%\citet{Laurent:2012aa}
+
+\citet{Laurent:2018aa} introduces the ISC sequent-style system,
+written $\Gamma \vdash B$, where $\Gamma$ is a sequence of types
+$A_1,\ldots,A_n$. The intuition is that $A_1,\ldots,A_n \vdash B$
+corresponds to $A_1 ∩ \cdots ∩ A_n ≤ B$. The ISC system satisfies the
+subformula property and is equivalent to the BCD system. To prove
+this, Laurent establishes six lemmas that culminate in
+cut-elimination, from which transitivity follows dirctly.
+
+This article presents a more direct route to the subformula property
+and transitivity. We present a subtyping relation $A <: B$ and
+directly prove transitivity, without using an auxilliary sequent-style
+system. Nevertheless, the intuitions are based on those of
+Laurent. The key to $A <: B$ is a rule for function types based on the
+$\beta$-soundness property~\citep{Barendregt:2013aa}, just as in ISC.
+
+ROAD MAP
 
@@
 
-Figure~\ref{fig:aux}
+\section{Intersection Subtyping, with Subformula}
+
+Our new subtyping judgment relies on several auxilliary notions, which
+we define in Figure~\ref{fig:aux}. These include the $\dom{}$ and
+$\cod{}$ functions. For function types they simply return the domain
+and codomain, respectively. For intersections, $\dom{A ∩ B}$ is the
+intersection of the domain of $A$ and $B$. Similarly for $\cod{}$.
+Both $\dom{}$ and $\cod{}$ are undefined on atoms.
 
@@
   \end{tabular}
-  \caption{New Subtyping Relation}
+  \caption{The New Subtyping Judgment}
   \label{fig:new-subtyping}
 \end{figure}
+
+\section{Transitivity}
 
@@
 
-\begin{lemma}[Transitivity of $<:$]\label{lem:sub-trans}
+\begin{theorem}[Transitivity of $<:$]\label{lem:sub-trans}
     If $A <: B$ and $B <: C$, then $A <: C$.
-\end{lemma}
+\end{theorem}
 
@@
 % LocalWords:  dom subtyping Barendregt et al subformula sequent aa
-% LocalWords:  BCD lclr refl glb topInCod
+% LocalWords:  BCD lclr refl glb topInCod Coppo Dezani Ciancaglini fk
+% LocalWords:  Salle Engeler Honsell Abramsky Plotkin Rocca Alessi
+% LocalWords:  subsumption subtype denotational ISWIM Landin
+% LocalWords:  Felleisen

### 2019-06-23 papers/intersection-transitivity.tex
check

@@
 \newcommand{\cod}[1]{\mathrm{cod}(#1)}
+\newcommand{\topP}[1]{\mathrm{top}(#1)}
+\newcommand{\topInCod}[1]{\mathrm{topInCod}(#1)}
 
@@
 different
-calculi~\citep{Coppo:1980ab,Coppo:1981aa,Engeler:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Plotkin:1993ab,Honsell:1999aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa}.
+calculi~\citep{Coppo:1980ab,Coppo:1981aa,Engeler:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Plotkin:1993ab,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa}.
 Perhaps the best-known of them is the BCD intersection type system of
@@
 \begin{figure}[tbp]
-  \fbox{$A ≤ B$} \\[1ex]
-
-  \centering
-  \begin{tabular}{p{1in}l}
-    (refl)  & $A ≤ A$ \\[3ex]
-    (incl$_L$) & $A ∩ B ≤ A$ \\[3ex]
-    (incl$_R$) & $A ∩ B ≤ B$ \\[3ex]
-    (glb) & $\inference{C ≤ A & C ≤ B}{C ≤ A ∩ B}$ \\[3ex]
-    (trans) & $\inference{A ≤ B & B ≤ C}{A ≤ C}$ \\[3ex]
-    ($\TOP_{\mathrm{top}}$) & $A ≤ \TOP$ \\[3ex]
-    ($\TOP{→}$) & $\TOP ≤ A → \TOP$ \\[3ex]
-    (${→}{∩}$) & $(A → B) ∩ (A → C) ≤ A → (B ∩ C)$ \\[3ex]
-    ($→$) & $\inference{A' ≤ A & B ≤ B'}{A → B ≤ A' → B'}$
-  \end{tabular}
-  \caption{The Subtyping Judgment of Barendregt, Coppo, and Dezani (BCD).}
+  \fbox{$A ≤ B$}
+  \begin{gather*}
+    \text{(refl)} \; \inference{}{A ≤ A} \quad
+    \text{(trans)} \; \inference{A ≤ B & B ≤ C}{A ≤ C} \\[3ex]
+    \text{(incl$_L$)} \; \inference{}{A ∩ B ≤ A} \quad
+    \text{(incl$_R$)} \; \inference{}{A ∩ B ≤ B} \quad
+    \text{(glb)} \; \inference{C ≤ A & C ≤ B}{C ≤ A ∩ B} \\[3ex]
+    (→) \; \inference{A' ≤ A & B ≤ B'}{A → B ≤ A' → B'} \quad
+    ({→}{∩}) \; \inference{}{(A → B) ∩ (A → C) ≤ A → (B ∩ C)} \\[3ex]
+    (\TOP_{\mathrm{top}}) \; \inference{}{A ≤ \TOP} \quad
+    (\TOP{→}) \; \inference{}{\TOP ≤ A → \TOP}
+  \end{gather*}
+  \caption{Subtyping of Barendregt, Coppo, and
+    Dezani-Ciancaglini (BCD).}
   \label{fig:BCD-subtyping}
@@
 that is much easier when subtyping satisfies the subformula property.
-
-%\citet{Laurent:2012aa}
 
@@
 $\beta$-soundness property~\citep{Barendregt:2013aa}, just as in ISC.
+The definitions and results in this article have been machine checked
+in Agda.
 
@@
 
-Our new subtyping judgment relies on several auxilliary notions, which
-we define in Figure~\ref{fig:aux}. These include the $\dom{}$ and
-$\cod{}$ functions. For function types they simply return the domain
-and codomain, respectively. For intersections, $\dom{A ∩ B}$ is the
-intersection of the domain of $A$ and $B$. Similarly for $\cod{}$.
-Both $\dom{}$ and $\cod{}$ are undefined on atoms.
+Our new subtyping judgment relies on several auxilliary notions that
+help us avoid the use of ellipses, which we define in
+Figure~\ref{fig:aux}. These include the $\dom{A}$ and $\cod{A}$
+functions, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
+relations $A ∈ B$ and $A ⊆ B$.
+%
+The $\dom{A}$ and $\cod{A}$ functions return the domain or codomain if
+$A$ is a function type, respectively. If $A$ is an intersection $A_1 ∩
+A_2$, then $\dom{A}$ is the intersection of the domain of $A_1$ and
+$A_2$.  If $A$ is an atom, $\dom{A}$ is undefined. Likewise for
+$\cod{A}$. For example, if $A = (A_1 → B_1) ∩ \cdots ∩ (A_n → B_n)$,
+then $\dom{A} = A_1 ∩ \cdots ∩ A_n$ and $\cod{A} = B_1 ∩ \cdots ∩ B_n$.
+%
+The $\topP{A}$ predicate identifies types that are equivalent to
+$\TOP$. The $\topInCod{A}$ predicate identifies types that have $\TOP$
+in their codomain.
+%
+The relation $A ∈ B$ indicates whether $A$ is syntactically a part of $B$.
+The relation $A ⊆ B$ holds when every part of $A$ is a part of $B$.
+We say that $B$ \emph{contains} $A$ if $A ⊆ B$.
+
+\begin{proposition}
+\item If $A ∩ B ⊆ C$, then $A ⊆ C$ and $B ⊆ C$. \label{prop:⊔⊆-inv}
+\end{proposition}
 
@@
 
-  \fbox{$\mathsf{top}(A)$}
+  \fbox{$\topP{A}$}
   \begin{gather*}
-    \mathsf{top}(\TOP)
+    \inference{}{\topP{\TOP}}
     \quad
-    \inference{\mathsf{top}(B)}{\mathsf{top}(A → B)}
+    \inference{\topP{B}}{\topP{A → B}}
     \quad
-    \inference{\mathsf{top}(A) & \mathsf{top}(B)}{\mathsf{top}(A ∩ B)}
+    \inference{\topP{A} & \topP{B}}{\topP{A ∩ B}}
   \end{gather*}
@@
     \inference{}{\alpha ∈ \alpha}  \quad
-    \inference{}{A → B ∈ A → B} \\[3ex]
+    \inference{}{A → B ∈ A → B} \quad
     \inference{A ∈ B}{A ∈ B ∩ C} \quad
@@
 
-  \caption{Auxiliary definitions}
+  \caption{Auxiliary Definitions}
   \label{fig:aux}
@@
 \begin{figure}[tbp]
-  \fbox{$A <: B$} \\[1ex]
-
-  \centering
-  \begin{tabular}{p{1in}l}
-    (refl$_α$) & $\inference{}{α <: α}$ \\[3ex]
-    (lb$_L$) & $\inference{A <: C}{A ∩ B <: C}$ \\[3ex]
-    (lb$_R$) & $\inference{B <: C}{A ∩ B <: C}$ \\[3ex]
-    (glb) & $\inference{C ≤ A & C ≤ B}{C ≤ A ∩ B}$ \\[3ex]
-    ($\TOP_{\mathrm{top}}$) & $\inference{}{A <: \TOP}$ \\[3ex]
-    ($\TOP{→}'$) & $\inference{}{C <: A → B}\;\mathsf{top}(B)$ \\[3ex]
-    ($→'$) &$\inference{A <: \dom{D} & \cod{D} <: B }{C <: A → B}\begin{array}{l} \neg\, \mathsf{top}(B) \\ \neg \, \mathsf{topInCod}(D) \\ D ⊆ C \end{array}$
-  \end{tabular}
+  \fbox{$A <: B$}
+  \begin{gather*}
+    \text{(refl$_α$)} \; \inference{}{α <: α} \\[3ex]
+    \text{(lb$_L$)} \; \inference{A <: C}{A ∩ B <: C} \quad
+    \text{(lb$_R$)} \; \inference{B <: C}{A ∩ B <: C} \quad
+    \text{(glb)} \; \inference{C ≤ A & C ≤ B}{C ≤ A ∩ B} \\[3ex]
+    (→') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C → D}
+    \begin{array}{l} \neg\, \mathsf{top}(D) \\
+      \neg \, \mathsf{topInCod}(B) \\
+      B ⊆ A \end{array}\\[3ex]
+    (\TOP_{\mathrm{top}}) \; \inference{}{A <: \TOP} \quad
+    (\TOP{→}') \; \inference{}{C <: A → B}\;\mathsf{top}(B)
+  \end{gather*}
   \caption{The New Subtyping Judgment}
@@
 
+The new intersection subtyping judgment, $A <: B$, is defined in
+Figure~\ref{fig:new-subtyping}. Of course, it omits the (trans) rule.
+We also omit the (refl) rule, replacing it with reflexivity for atoms
+(refl$_\alpha$). The most important rule is the one for function types
+$(→')$, which subsumes the rules $(→)$ and $({→}{∩})$ in BCD
+subtyping. It essentially turns the $\beta$-soundness property into a
+subtyping rule. It says that a type $A$ is a subtype of a function
+type $C → D$ if $A$ contains a type $B$ whose domain and codomain are
+larger and smaller than $C$ and $D$, respectively. The use of $B$
+instead of $A$ in the premise, with the condition $B ⊆ A$, allows this
+rule to absorb uses of (incl$_L$) and (incl$_R$) on the left.  The
+side conditions $\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed
+because of the (\TOP{→}') rule, which in turn is needed to preserve
+types under $\eta$-reduction.  In a system that does not involve
+$\eta$-reduction, the (\TOP{→}') rule can be ommitted, as well as
+those side conditions. The rules (lb$_L$) and (lb$_R$) adapt
+(incl$_L$) and (incl$_R$) to a system without transitivity, and have
+appeared many times in the literature~\citep{Bakel:1995aa}.
+
+\begin{proposition}\ 
+  \begin{enumerate}
+  \item (reflexivity) $A <: A$ \label{prop:refl}
+  \item If $A <: B ∩ C$, then $A <: B$ and $A <: C$. \label{prop:⊔⊑-inv}
+  \end{enumerate}
+\end{proposition}
+\begin{proof}
+  The proof of reflexivity is by induction on $A$. In the case
+  $A = A_1 → A_2$, we proceed by cases on whether $\topP{A_2}$.
+  If it is, deduce $A_1 → A_2 <: A_1 → A_2$ by rule $(\TOP{→}')$.
+  Otherwise, we use rule $(→')$
+\end{proof}
+
+\begin{proposition}\label{prop:u⊆v⊑w→u⊑w}
+ If $A <: B$ and $C ⊆ B$, then $A <: C$.
+\end{proposition}
+
+\clearpage
+\pagebreak
 \section{Transitivity}
@@
 
-\begin{lemma}\label{lem:sub-fun-inv}
+\begin{lemma}\label{lem:⊑-fun-inv}
   If $\neg\,\mathsf{top}(B)$,
@@
 
+\section{Equivalence of BCD Subtyping and the new Subtyping}
+
 \begin{theorem}[Equivalence of the subtyping relations]\ \\

### 2019-06-23 papers/intersection-transitivity.tex
check

@@
     \text{(incl$_R$)} \; \inference{}{A ∩ B ≤ B} \quad
-    \text{(glb)} \; \inference{C ≤ A & C ≤ B}{C ≤ A ∩ B} \\[3ex]
-    (→) \; \inference{A' ≤ A & B ≤ B'}{A → B ≤ A' → B'} \quad
+    \text{(glb)} \; \inference{A ≤ C & A ≤ D}{A ≤ C ∩ D} \\[3ex]
+    (→) \; \inference{C ≤ A & B ≤ D}{A → B ≤ C → D} \quad
     ({→}{∩}) \; \inference{}{(A → B) ∩ (A → C) ≤ A → (B ∩ C)} \\[3ex]
     (\TOP_{\mathrm{top}}) \; \inference{}{A ≤ \TOP} \quad
-    (\TOP{→}) \; \inference{}{\TOP ≤ A → \TOP}
+    (\TOP{→}) \; \inference{}{\TOP ≤ C → \TOP}
   \end{gather*}
@@
 $\cod{A}$. For example, if $A = (A_1 → B_1) ∩ \cdots ∩ (A_n → B_n)$,
-then $\dom{A} = A_1 ∩ \cdots ∩ A_n$ and $\cod{A} = B_1 ∩ \cdots ∩ B_n$.
+then $\dom{A} = A_1 ∩ \cdots ∩ A_n$ and $\cod{A} = B_1 ∩ \cdots ∩
+B_n$.  When $\dom{A}$ or $\cod{A}$ appears in lemma or theorem
+statement, we implicitly assume that $A$ is a type such that $\dom{A}$
+and $\cod{A}$ are defined.
 %
@@
     \text{(lb$_R$)} \; \inference{B <: C}{A ∩ B <: C} \quad
-    \text{(glb)} \; \inference{C ≤ A & C ≤ B}{C ≤ A ∩ B} \\[3ex]
+    \text{(glb)} \; \inference{A ≤ C & A ≤ D}{A ≤ C ∩ D} \\[3ex]
     (→') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C → D}
@@
     (\TOP_{\mathrm{top}}) \; \inference{}{A <: \TOP} \quad
-    (\TOP{→}') \; \inference{}{C <: A → B}\;\mathsf{top}(B)
+    (\TOP{→}') \; \inference{}{A <: C → D}\;\mathsf{top}(D)
   \end{gather*}
@@
 The new intersection subtyping judgment, $A <: B$, is defined in
-Figure~\ref{fig:new-subtyping}. Of course, it omits the (trans) rule.
-We also omit the (refl) rule, replacing it with reflexivity for atoms
+Figure~\ref{fig:new-subtyping}. First, it does not include the (trans)
+rule.  It also replaces the (refl) rule with reflexivity for atoms
 (refl$_\alpha$). The most important rule is the one for function types
-$(→')$, which subsumes the rules $(→)$ and $({→}{∩})$ in BCD
-subtyping. It essentially turns the $\beta$-soundness property into a
+$(→')$, which subsumes $(→)$ and $({→}{∩})$ in BCD subtyping.  The
+$(→')$ rule essentially turns the $\beta$-soundness property into a
 subtyping rule. It says that a type $A$ is a subtype of a function
@@
 (incl$_L$) and (incl$_R$) to a system without transitivity, and have
-appeared many times in the literature~\citep{Bakel:1995aa}.
+appeared many times in the literature~\citep{Bakel:1995aa}.  The
+$(\TOP{→}')$ rule generalizes the (\TOP{→}) rule, replacing the $\TOP$
+on the left with any type $A$, because for transitivity, any type is
+below $\TOP$. The $(\TOP{→}')$ rule also replaces the $\TOP$ in the
+codomain on the right with any type $D$ that is equivalent to $\TOP$.
 
@@
   \item If $A <: B ∩ C$, then $A <: B$ and $A <: C$. \label{prop:⊔⊑-inv}
+  \item If $A <: B$ and $C ∈ B$, then $A <: C$.\label{prop:u∈v⊑w→u⊑w}
+  \item If $A <: B$ and $C ⊆ B$, then $A <: C$.\label{prop:u⊆v⊑w→u⊑w}
   \end{enumerate}
 \end{proposition}
-\begin{proof}
-  The proof of reflexivity is by induction on $A$. In the case
+\begin{proof}\ 
+  \begin{enumerate}
+    \item The proof of reflexivity is by induction on $A$. In the case
       $A = A_1 → A_2$, we proceed by cases on whether $\topP{A_2}$.
@@
       Otherwise, we use rule $(→')$
+    \item The proof is by induction on the derivation of $A <: B ∩ C$.
+    \item The proof is by induction on $B$. In the case where $B = B_1 ∩ B_2$,
+      we have that either $C ∈ B_1$ or $C ∈ B_2$, but in
+      either case we use part \ref{prop:⊔⊑-inv} of this proposition to
+      fulfill the premise of the induction hypothesis.
+    \item The proof  is by induction on $C$, using part \ref{prop:u∈v⊑w→u⊑w}
+      of this proposition in the cases for atoms and function types. 
+  \end{enumerate}
 \end{proof}
 
-\begin{proposition}\label{prop:u⊆v⊑w→u⊑w}
- If $A <: B$ and $C ⊆ B$, then $A <: C$.
+Next we turn to the subtyping inversion principle for function types.
+The idea is to generalize the rule $(→')$ with respect to the type on
+the right, allowing any type that contains a function type.  The
+premises of $(→')$ are somewhat complex, so we package most of them
+into the following definition.
+
+\begin{definition}[factors]
+  We say $C → D$ \emph{factors} $A$
+  if there exists some type $B$ such that
+  $B ⊆ A$, $\neg\,\mathsf{topInCod}(B)$, $C <: \dom{B}$, and $\cod{B} <: D$.
+\end{definition}
+
+\begin{proposition}[Inversion Principle for Function Types]\label{lem:⊑-fun-inv}
+  If $A <: C'$, $C → D ∈ C'$, and $\neg\,\mathsf{top}(D)$, then
+  $C → D$ factors $A$.
 \end{proposition}
+\begin{proof}
+  The proof is a straightforward induction on $A <: C'$.
+\end{proof}
 
-\clearpage
-\pagebreak
 \section{Transitivity}
 
-\begin{definition}[factors]
-  We say $A → B$ \emph{factors} $C$
-  if there exists some type $C'$ such that
-  $C' ⊆ C$, $\neg\,\mathsf{topInCod}(C')$, $A <: \dom{C'}$, and $\cod{C'} <: B$.
-\end{definition}
-
-\begin{lemma}\label{lem:⊑-fun-inv}
-  If $\neg\,\mathsf{top}(B)$,
-  $D <: C$, and
-  $A → B ∈ C$, then
-  $A → B$ factors $D$.
-\end{lemma}
+The proof of transitivity relies on the following lemma, which is
+traditionally needed to prove the inversion principle for function
+types. However, it was not needed for our system because the rule
+$(→')$ is already quite close to the inversion principle.
+The lemma states that if every function type $C → D$ in A
+factors $B$, then $\dom{A} → \cod{A}$ also factors $B$.
 
@@
   \begin{itemize}
-  \item $\neg\, \mathsf{topInCod}(D)$,
-  \item for any $A,B$, if $\neg\,\mathsf{top}(B)$ and $A → B ∈ D$,
-    then $A → B$ factors $C$,
+  \item for any $C$ and $D$, if $C → D ∈ A$ and $\neg\,\mathsf{top}(D)$,
+    then $C → D$ factors $B$, and
+  \item $\neg\, \mathsf{topInCod}(A)$,
   \end{itemize}
-  then $\dom{D} → \cod{D}$ factors $C$.
+  then $\dom{A} → \cod{A}$ factors $B$.
 \end{lemma}
+\begin{proof}
+  The proof is by induction on $A$.
+  \begin{itemize}
+  \item Case $A$ is an atom. The statement is vacuously true.
+  \item Case $A = A_1 → A_2$ is a function type. Then we conclude by applying
+    the premise with $C$ and $D$ instantiated to $A_1$ and $A_2$ respectively.
+  \item Case $A = A_1 ∩ A_2$.
+
+    UNDER CONSTRUCTION
+  \end{itemize}
+\end{proof}
 

### 2019-06-23 papers/intersection-transitivity.tex
check

@@
     the premise with $C$ and $D$ instantiated to $A_1$ and $A_2$ respectively.
-  \item Case $A = A_1 ∩ A_2$.
-
-    UNDER CONSTRUCTION
+  \item Case $A = A_1 ∩ A_2$.  By the induction hypothesis for $A_1$
+    and for $A_2$, we have that $\dom{A_1} → \cod{A_1}$ factors $B$
+    and so does $\dom{A_2} → \cod{A_2}$.  So there exists
+    $B_1$ and $B_2$ such that $B_1 ⊆ B$, $\neg\,\topInCod{B_1}$,
+    $\dom{A_1} <: \dom{B_1}$, $\cod{B_1} <: \cod{A_2}$ and similarly
+    for $B_2$. We need to show that $\dom{A} → \cod{A}$ factors
+    $B$. We choose the witness $B_1 ∩ B_2$.  Clearly we have $B_1 ∩
+    B_2 ⊆ B$ and $\neg \topInCod{B_1 ∩ B_2}$.  Also, we have
+    \[
+    \dom{B_1} ∩ \dom{B_2} <: \dom{A_1} ∩ \dom{A_2}
+    \]
+    and
+    \[
+    \cod{A_1} ∩ \cod{A_2} <: \cod{B_1} ∩ \cod{B_2}
+    \]
+    Thus, we have that $\dom{B_1 ∩ B_2} <: \dom{A}$
+    and $\cod{A} <: \cod{B_1 ∩ B_2}$, and this case is complete.
   \end{itemize}
 \end{proof}
+
+We now turn to the proof of transitivity, that
+if $A <: B$ and $B <: C$, then $A <: C$.
 

### 2019-06-24 papers/intersection-transitivity.tex
check

@@
 \usepackage{fontspec,unicode-math}
+\usepackage{hyperref}
 
@@
 \newcommand{\TOP}{\ensuremath{\mathtt{U}}}
-\newcommand{\dom}[1]{\mathrm{dom}(#1)}
-\newcommand{\cod}[1]{\mathrm{cod}(#1)}
-\newcommand{\topP}[1]{\mathrm{top}(#1)}
-\newcommand{\topInCod}[1]{\mathrm{topInCod}(#1)}
+\newcommand{\dom}[1]{\mathsf{dom}(#1)}
+\newcommand{\cod}[1]{\mathsf{cod}(#1)}
+\newcommand{\topP}[1]{\mathsf{top}(#1)}
+\newcommand{\topInCod}[1]{\mathsf{topInCod}(#1)}
+\newcommand{\depth}[1]{\mathsf{depth}(#1)}
+\newcommand{\size}[1]{\mathsf{size}(#1)}
 
@@
   subtyping judgment, in regular style, that satisfies the subformula
-  property, and gives a direct proof of transitivity. Borrowing from
-  Laurent's system, the rule for function types is essentially the
-  $\beta$-soundness property.  The main lemma required for the
+  property, and presents a direct proof of transitivity. Borrowing
+  from Laurent's system, the rule for function types is essentially
+  the $\beta$-soundness property.  The main lemma required for the
   transitivity proof is one that has been used to prove the inversion
-  principle for subtyping of function types (Barendregt et
-  al. 2013). The choice of induction principle for the proof of
-  transitivity requires some care: we use strong induction on the
-  lexicographical ordering of the sum of the depths of the first and
-  last type followed by the sum of the sizes of the first and second
-  type.
+  principle for subtyping of function types. The choice of induction
+  principle for the proof of transitivity is subtle: we use
+  well-founded induction on the lexicographical ordering of the sum of
+  the depths of the first and last type followed by the sum of the
+  sizes of the middle and last type. The article concludes with a
+  proof that the new subtyping judgment is equivalent to that of
+  Barendregt, Coppo, and Dezani-Ciancaglini.
 \end{abstract}
@@
 calculus~\citep{Coppo:1979aa}. By varying the subtyping rules and atom
-types, researchers have used intersection type systems to model many
+types, researchers use intersection type systems to model many
 different
@@
 system, following the presentation of \citet{Barendregt:2013aa}.  We
-conjecture that our results apply to other intersection type systems.
+conjecture that our results apply to other intersection type systems
+as well.
 
@@
   \begin{array}{lclr}
-    \alpha,\beta & ::= & \TOP \mid c_0 \mid c_1 \mid c_2 \mid \cdots & \text{atoms}\\
-    A,B,C,D & ::= & \alpha \mid A → B \mid A ∩ B & \text{types}
+    α,\beta & ::= & \TOP \mid c_0 \mid c_1 \mid c_2 \mid \cdots & \text{atoms}\\
+    A,B,C,D & ::= & α \mid A → B \mid A ∩ B & \text{types}
   \end{array}
@@
 rules, and then then prove transitivity~\citep{Muehlboeck:2018aa}.
-Unfortunately, the ${→}{∩}$ rule significantly complicates the
-situation for the BCD system.
+Unfortunately, the $({→}{∩})$ rule of the BCD system significantly
+complicates the situation.
 
@@
 It seems that doing so requires placing extra conditions on types and
-that is much easier when subtyping satisfies the subformula property.
+it is much easier to do so when subtyping satisfies the subformula
+property.
 
@@
 this, Laurent establishes six lemmas that culminate in
-cut-elimination, from which transitivity follows dirctly.
+cut-elimination, from which transitivity follows.
 
@@
 and transitivity. We present a subtyping relation $A <: B$ and
-directly prove transitivity, without using an auxilliary sequent-style
+directly prove transitivity without using an auxiliary sequent-style
 system. Nevertheless, the intuitions are based on those of
@@
 
-ROAD MAP
+We describe the new subtyping relation in
+Section~\ref{sec:new-subtyping}, prove transitivity in
+Section~\ref{sec:trans}, and prove its equivalence to BCD subtyping in
+Section~\ref{sec:equiv}. We make some concluding remarks in
+Section~\ref{sec:conclude}.
 
@@
 
-\section{Intersection Subtyping, with Subformula}
+\section{A New Subtyping Judgment}
+\label{sec:new-subtyping}
 
-Our new subtyping judgment relies on several auxilliary notions that
+Our new subtyping judgment relies on several auxiliary notions that
 help us avoid the use of ellipses, which we define in
@@
 
-\begin{proposition}
-\item If $A ∩ B ⊆ C$, then $A ⊆ C$ and $B ⊆ C$. \label{prop:⊔⊆-inv}
+\begin{proposition}\label{prop:⊔⊆-inv}
+ \item If $A ∩ B ⊆ C$, then $A ⊆ C$ and $B ⊆ C$. 
 \end{proposition}
@@
   \begin{gather*}
-    \inference{}{\alpha ∈ \alpha}  \quad
+    \inference{}{α ∈ α}  \quad
     \inference{}{A → B ∈ A → B} \quad
@@
     \text{(lb$_R$)} \; \inference{B <: C}{A ∩ B <: C} \quad
-    \text{(glb)} \; \inference{A ≤ C & A ≤ D}{A ≤ C ∩ D} \\[3ex]
+    \text{(glb)} \; \inference{A <: C & A <: D}{A <: C ∩ D} \\[3ex]
     (→') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C → D}
@@
 rule.  It also replaces the (refl) rule with reflexivity for atoms
-(refl$_\alpha$). The most important rule is the one for function types
+(refl$_α$). The most important rule is the one for function types
 $(→')$, which subsumes $(→)$ and $({→}{∩})$ in BCD subtyping.  The
 $(→')$ rule essentially turns the $\beta$-soundness property into a
-subtyping rule. It says that a type $A$ is a subtype of a function
-type $C → D$ if $A$ contains a type $B$ whose domain and codomain are
-larger and smaller than $C$ and $D$, respectively. The use of $B$
-instead of $A$ in the premise, with the condition $B ⊆ A$, allows this
-rule to absorb uses of (incl$_L$) and (incl$_R$) on the left.  The
-side conditions $\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed
-because of the (\TOP{→}') rule, which in turn is needed to preserve
-types under $\eta$-reduction.  In a system that does not involve
-$\eta$-reduction, the (\TOP{→}') rule can be ommitted, as well as
-those side conditions. The rules (lb$_L$) and (lb$_R$) adapt
-(incl$_L$) and (incl$_R$) to a system without transitivity, and have
-appeared many times in the literature~\citep{Bakel:1995aa}.  The
-$(\TOP{→}')$ rule generalizes the (\TOP{→}) rule, replacing the $\TOP$
-on the left with any type $A$, because for transitivity, any type is
-below $\TOP$. The $(\TOP{→}')$ rule also replaces the $\TOP$ in the
-codomain on the right with any type $D$ that is equivalent to $\TOP$.
+subtyping rule. The $(→')$ rule says that a type $A$ is a subtype of a
+function type $C → D$ if a subset of $A$, call it $B$, has domain and
+codomain that are larger and smaller than $C$ and $D$,
+respectively. The use of a subset of $A$ enables this rule to absorb
+uses of (incl$_L$) and (incl$_R$) on the left.  The side conditions
+$\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed because of the
+(\TOP{→}') rule, which in turn is needed to preserve types under
+$\eta$-reduction.  In a system that does not involve $\eta$-reduction,
+the (\TOP{→}') rule can be omitted, as well as those side
+conditions. The rules (lb$_L$) and (lb$_R$) adapt (incl$_L$) and
+(incl$_R$) to a system without transitivity, and have appeared many
+times in the literature~\citep{Bakel:1995aa}.  The $(\TOP{→}')$ rule
+generalizes the (\TOP{→}) rule, replacing the $\TOP$ on the left with
+any type $A$, because for transitivity, any type is below $\TOP$. The
+$(\TOP{→}')$ rule also replaces the $\TOP$ in the codomain on the
+right with any type $D$ that is equivalent to $\TOP$.
 
-\begin{proposition}\ 
+Before moving on, we make note of some basic facts regarding the $<:$
+relation and the $\topP{A}$ predicate.
+
+\begin{proposition}[Basic Properties of $<:$]\ \label{prop:subtyping}
   \begin{enumerate}
-  \item (reflexivity) $A <: A$ \label{prop:refl}
+  \item (reflexivity) $A <: A$ \label{prop:⊑-refl}
   \item If $A <: B ∩ C$, then $A <: B$ and $A <: C$. \label{prop:⊔⊑-inv}
@@
     If it is, deduce $A_1 → A_2 <: A_1 → A_2$ by rule $(\TOP{→}')$.
-      Otherwise, we use rule $(→')$
+    Otherwise, apply rule $(→')$
   \item The proof is by induction on the derivation of $A <: B ∩ C$.
-    \item The proof is by induction on $B$. In the case where $B = B_1 ∩ B_2$,
-      we have that either $C ∈ B_1$ or $C ∈ B_2$, but in
-      either case we use part \ref{prop:⊔⊑-inv} of this proposition to
-      fulfill the premise of the induction hypothesis.
+  \item The proof is by induction on $B$. In the case where $B = B_1
+    ∩ B_2$, either $C ∈ B_1$ or $C ∈ B_2$, but in either case part
+    \ref{prop:⊔⊑-inv} of this proposition fulfills the premise of
+    the induction hypothesis, from which the conclusion follows.
   \item The proof  is by induction on $C$, using part \ref{prop:u∈v⊑w→u⊑w}
     of this proposition in the cases for atoms and function types.
+  \end{enumerate}
+\end{proof}
+
+\begin{proposition}[Properties of $\topP{A}$]\ \label{prop:top}
+  \begin{enumerate}
+  \item If $\topP{A}$ then $\topP{\cod{A}}$.\label{prop:AllBot-cod}
+  \item If $\topP{A}$ and $B ∈ A$, then $\topP{B}$.\label{prop:AllBot-∈}
+  \item If $\topP{A}$ and $B ⊆ A$, then $\topP{B}$.\label{prop:AllBot-⊆}
+  \item If $\topP{A}$ and $A <: B$, then $\topP{B}$.\label{prop:AllBot-⊑}
+  \item If $\topP{A}$, then $B <: A$.\label{prop:AllBot-⊑-any}
+  \end{enumerate}
+\end{proposition}
+\begin{proof}\
+  \begin{enumerate}
+  \item The proof is a straightforward induction on $A$.
+  \item The proof is also a straightforward induction on $A$.
+  \item The proof is by induction on $B$. The cases for atoms and
+    function types are proved by part \ref{prop:AllBot-∈} of
+    this proposition. In the case for
+    $B = B_1 ∩ B_2$, from $B_1 ∩ B_2 <: A$, we have
+    $B_1 <: A$ and $B_2 <: A$ (Proposition~\ref{prop:⊔⊆-inv}).
+    Then by the induction hypotheses for $B_1$ and $B_2$ we have
+    $\topP{B_1}$ and $\topP{B_2}$, from which we conclude
+    that $\topP{B_1 ∩ B_2}$.
+  \item The proof is by induction on the derivation of $A <: B$.
+    All of the cases are straightforward except for rule $(→')$.
+    In that case we have $B = B_1 → B_2$ and some $A'$ such that
+    $A' ⊆ A$, $B_1 <: \dom{A'}$, $\cod{A'} <: B_2$, $\neg\,\topP{B_2}$,
+    and $\neg\,\topInCod{A'}$. From the premise $\topP{A}$
+    and part \ref{prop:AllBot-⊆} of this proposition, we
+    have $\topP{A'}$. Then by part \ref{prop:AllBot-cod}
+    we have $\topP{\cod{A'}}$. By the induction hypothesis
+    for $\cod{A'} <: B$ we conclude that $\topP{B}$.
+  \item The proof is a straightforward induction on $A$.
   \end{enumerate}
@@
   if there exists some type $B$ such that
-  $B ⊆ A$, $\neg\,\mathsf{topInCod}(B)$, $C <: \dom{B}$, and $\cod{B} <: D$.
+  $B ⊆ A$, $C <: \dom{B}$, $\cod{B} <: D$, and
+  $\neg\,\mathsf{topInCod}(B)$.
 \end{definition}
 
-\begin{proposition}[Inversion Principle for Function Types]\label{lem:⊑-fun-inv}
+\begin{proposition}[Inversion Principle for Function Types]
+  \label{prop:⊑-fun-inv}
   If $A <: C'$, $C → D ∈ C'$, and $\neg\,\mathsf{top}(D)$, then
@@
 \section{Transitivity}
+\label{sec:trans}
 
@@
   \begin{itemize}
-  \item for any $C$ and $D$, if $C → D ∈ A$ and $\neg\,\mathsf{top}(D)$,
+  \item for any $C$ $D$, if $C → D ∈ A$ and $\neg\,\mathsf{top}(D)$,
     then $C → D$ factors $B$, and
@@
 
-We now turn to the proof of transitivity, that
-if $A <: B$ and $B <: C$, then $A <: C$.
-
+We now turn to the proof of transitivity, that if $A <: B$ and $B <:
+C$, then $A <: C$. The proof is by well-founded induction on the
+lexicographical ordering of the sum of the depths of $A$ and $C$
+followed by the sum of the sizes of $B$ and $C$. To be precise, we
+define this ordering as follows.
+\[
+\langle A, B, C \rangle \ll \langle A', B', C' \rangle
+=
+\begin{array}{l}
+\depth{A} + \depth{C} < \depth{A'} + \depth{C'}
+\text{ or } \\
+(\depth{A} + \depth{C}
+\leq
+\depth{A'} + \depth{C'} \\
+\text{ and }
+\size{B} + \size{C}
+<
+\size{B'} + \size{C'})
+\end{array}
+\]
+where $\size{A}$ is 
 \begin{align*}
-  \mathsf{size}(\alpha) &= 0 \\
-  \mathsf{size}(A → B) &= 1 + \mathsf{size}(A) + \mathsf{size}(B) \\
-  \mathsf{size}(A ∩ B) &= 1 + \mathsf{size}(A) + \mathsf{size}(B)
+  \size{α} &= 0 \\
+  \size{A → B} &= 1 + \size{A} + \size{B} \\
+  \size{A ∩ B} &= 1 + \size{A} + \size{B}
+\end{align*}
+and $\depth{A}$ is 
+\begin{align*}
+  \depth{α} &= 0 \\
+  \depth{A → B} &= 1 + \max(\depth{A}, \depth{B}) \\
+  \depth{A ∩ B} &= \max(\depth{A}, \depth{B})
 \end{align*}
 
-\begin{align*}
-  \mathsf{depth}(\alpha) &= 0 \\
-  \mathsf{depth}(A → B) &= 1 + \max(\mathsf{depth}(A), \mathsf{depth}(B)) \\
-  \mathsf{depth}(A ∩ B) &= \max(\mathsf{depth}(A), \mathsf{depth}(B))
-\end{align*}
-
-\begin{theorem}[Transitivity of $<:$]\label{lem:sub-trans}
+\begin{theorem}[Transitivity of $<:$]\label{thm:⊑-trans}
     If $A <: B$ and $B <: C$, then $A <: C$.
 \end{theorem}
+\begin{proof}
+  The proof is by well-founded induction on the relation $\ll$.
+  We proceed by cases on the last rule applied in the
+  derivation of $B <: C$.
+  \begin{description}
+  \item[Case (refl$_α$)] We have $B = C = α$.  From the premise $A <:
+    B$ we immediately conclude that $A <: α$.
+  \item[Case (lb$_L$)] So $B = B_1 ∩ B_2$, $B_1 <: C$, and $A <: B_1 ∩
+    B_2$.  We have $A <: B_1$ (Proposition~\ref{prop:subtyping} part
+    \ref{prop:⊔⊑-inv}), so we conclude that $A <: C$ by the induction
+    hypothesis, noting that $\langle A, B_1, C \rangle \ll \langle A,
+    B, C \rangle$ because $\size{B_1} < \size{B}$.
+  \item[Case (lb$_R$)] So $B = B_1 ∩ B_2$, $B_2 <: C$, and $A <: B_1 ∩
+    B_2$.  We have $A <: B_2$ (Proposition~\ref{prop:subtyping} part
+    \ref{prop:⊔⊑-inv}), so we conclude that $A <: C$ by the induction
+    hypothesis, noting that $\langle A, B_2, C \rangle \ll \langle A,
+    B, C \rangle$ because $\depth{B_2} \leq \depth{B}$ and $\size{B_2}
+    < \size{B}$.
+  \item[Case (glb)] We have $C = C_1 ∩ C_2$, $B <: C_1$, and $B <:
+    C_2$.  By the induction hypothesis, we have $A <: C_1$ and $A <:
+    C_2$, noting that $\langle A, B, C_1 \rangle \ll \langle A, B, C
+    \rangle$ $\langle A, B, C_2 \rangle \ll \langle A, B, C\rangle$
+    because $\depth{C_1} \leq \depth{C}$, $\depth{C_2} \leq
+    \depth{C}$, $\size{C_1} < C$, and $\size{C_2} < C$.
+    We conclude $A <: C_1 ∩ C_2$ by rule (glb).
+  \item[Case $(→')$] So $C = C_1 → C_2$, $\neg \topP{C_2}$, and there
+    exists $B'$ such that $C_1 <: \dom{B'}$, $\cod{B'} <: C_2$, $B' ⊆
+    B$, and $\neg \topInCod{B'}$. From $A <: B$ and $B' ⊆ B$, we have
+    $A <: B'$ (Proposition~\ref{prop:subtyping} part
+    \ref{prop:u⊆v⊑w→u⊑w}). Thus, for any $B_1 → B_2 ∈ B'$, $B_1 → B_2$
+    factors $A$ (Proposition~\ref{prop:⊑-fun-inv}). We have
+    satisfied the premises of Lemma~\ref{lem:sub-inv-trans},
+    so $\dom{B'} → \cod{B'}$ factors $A$. That means there exists
+    $A'$ such that $A' ⊆ A$, $\neg\,\topInCod{A'}$, $\dom{B'} <: \dom{A'}$,
+    and $\cod{A'} <: \cod{B'}$. Then by the induction hypothesis,
+    we have
+    \[
+    C_1 <: \dom{A'} \quad\text{and}\quad \cod{A'} <: C_2
+    \]
+    noting that
+    $\langle C_1, \dom{B'}, \dom{A'} \rangle \ll \langle A, B, C \rangle$
+    because $\depth{C_1} + \depth{\dom{A'}} < \depth{A} + \depth{C}$
+    and
+    $\langle \cod{A'}, \cod{B'}, C_2 \rangle \ll \langle A, B, C \rangle$
+    because $\depth{\cod{A'}} + \depth{C_2} < \depth{A} + \depth{C}$.
+    We conclude that $A <: C_1 → C_2$ by rule $(→')$ witnessed by $A'$.
+  \item[Case $(\TOP_{\mathrm{top}})$] We have $C = \TOP$ and
+    conclude $A <: \TOP$ by rule $(\TOP_{\mathrm{top}})$.
+  \item[Case $(\TOP{→}')$] We have $C = C_1 → C_2$ and $\topP{C_2}$.
+    We conclude $A <: C_1 → C_2$ by rule $(\TOP{→}')$.
+  \end{description}
+\end{proof}
 
-\section{Equivalence of BCD Subtyping and the new Subtyping}
+\section{Equivalence with BCD Subtyping}
+\label{sec:equiv}
+
+Having proved (trans), we next prove $(→)$ and $({→}{∩})$
+and then show that $A <: B$ is equivalent to $A ≤ B$.
+
+\begin{lemma}[$→$]\label{lem:⊑-fun′}
+  If $C <: A$ and $B <: D$, then
+  $A → B <: C → D$.
+\end{lemma}
+\begin{proof}
+  Consider whether $\topP{D}$ or not.
+  \begin{description}
+  \item[Case $\topP{D}$] We conclude $A → B <: C → D$
+    by rule $(\TOP{→}')$.
+  \item[Case $\neg \,\topP{D}$]
+    Consider whether $\topP{B}$ or not.
+    \begin{description}
+    \item[Case $\topP{B}$] So $\topP{D}$ (Prop. \ref{prop:top}
+      part \ref{prop:AllBot-⊑}), but that is a contradiction.
+    \item[Case $\neg\,\topP{B}$]
+      We conclude that $A → B <: C → D$ by rule $(→')$.
+    \end{description}
+  \end{description}
+\end{proof}
+
+\begin{lemma}[${→}{∩}$]\label{lem:⊑-dist}
+  $(A → B) ∩ (A → C) <: A → (B ∩ C)$
+\end{lemma}
+\begin{proof}
+   We consider the cases for whether $\topP{B}$ or $\topP{C}$.
+   \begin{description}
+   \item[Case $\topP{B}$ and $\topP{C}$]
+     Then $\topP{B ∩ C}$ and we conclude that
+     $(A → B) ∩ (A → C) <: A → (B ∩ C)$
+     by rule $(\TOP{→}')$.
+   \item[Case $\topP{B}$ and $\neg\,\topP{C}$]
+     We conclude that 
+     $(A → B) ∩ (A → C) <: A → (B ∩ C)$
+     by rule $(→')$, choosing the witness $A → C$
+     and noting that $C <: B$ 
+     by way of Proposition~\ref{prop:top}
+     part \ref{prop:AllBot-⊑-any}
+     and $C <: C$ by Proposition~\ref{prop:subtyping}
+     part \ref{prop:⊑-refl}.
+   \item[Case $\neg\,\topP{B}$ and $\topP{C}$]
+     We conclude that 
+     $(A → B) ∩ (A → C) <: A → (B ∩ C)$
+     by rule $(→')$, this time with witness $A → B$
+     and noting that
+     $B <: B$ by Proposition~\ref{prop:subtyping}
+     part \ref{prop:⊑-refl}
+     and $B <: C$ 
+     by way of Proposition~\ref{prop:top}
+     part \ref{prop:AllBot-⊑-any}.
+   \item[Case $\neg,\topP{B}$ and $\neg\,\topP{C}$]
+     Again we apply rule $(→')$, but with witness
+     $(A → B) ∩ (A → C)$.
+   \end{description}
+\end{proof}
+
+\noindent We require one more lemma.
+
+\begin{lemma}
+  \label{lem:dv↦cv<:v}
+  $A ≤ \dom{A} → \cod{A}$.
+\end{lemma}
+\begin{proof}
+  The proof is by induction on $A$.
+\end{proof}
+
+\noindent Now for the proof of equivalence
 
@@
 \end{theorem}
+\begin{proof}
+  We prove each direction of the if-and-only-if separately.
+  \begin{description}
+  \item[$A <: B$ implies $A ≤ B$]
+    We proceed by induction on the derivation of $A <: B$.
+    \begin{description}
+    \item[Case (refl$_α$)] We conclude $α ≤ α$ by (refl).      
+    \item[Case (lb$_L$)] By the induction hypothesis we have $A ≤ C$.
+      By (incl$_L$) we have $A ∩ B ≤ A$. We conclude
+      that $A ∩ B ≤ C$ by (trans).
+    \item[Case (lb$_R$)] By the induction hypothesis we have $B ≤ C$.
+      By (incl$_R$) we have $A ∩ B ≤ B$. We conclude
+      that $A ∩ B ≤ C$ by (trans).
+    \item[Case (glb)] By the induction hypothesis we have
+      $A ≤ C$ and $A ≤ D$, so we conclude that
+      $A ≤ C ∩ D$ by (glb).
+    \item[Case $(→')$] By the induction hypothesis we have
+      $C ≤ \dom{B}$ and also $\cod{B} ≤ D$.
+      From $B ⊆ A$ we have $A ≤ B$.      
+      Then by Lemma~\ref{lem:dv↦cv<:v} we have
+      $B ≤ \dom{B} → \cod{B}$.
+      Also, we have $\dom{B} → \cod{B} ≤ C → D$ by rule $(→)$.
+      We conclude that $A ≤ C → D$ by chaining the three prior
+      facts using (trans).
+    \item[Case $(\TOP_{\mathrm{top}})$]
+      We conclude that $A ≤ \TOP$ by $(\TOP_{\mathrm{top}})$.
+    \item[Case $(\TOP{→}')$] We have $A ≤ \TOP$ and $\TOP <: C →
+      \TOP$.  Also, $C → \TOP ≤ C → D$ because $\TOP <: D$ follows
+      from $\topP{D}$.  Thus, applying (trans) we conclude $A ≤ C →
+      D$.
+    \end{description}
 
-\clearpage
+  \item[$A ≤ B$ implies $A <: B$]
+    We proceed by induction on the derivation of $A ≤ B$.
+    \begin{description}
+    \item[Case (refl)] We conclude $A <: A$ by Prop.~\ref{prop:subtyping}
+      part \ref{prop:⊑-refl}.
+    \item[Case (trans)]
+      By the induction hypothesis, we have $A <: B$ and $B <: C$.
+      We conclude that $A <: C$ by Theorem~\ref{thm:⊑-trans}.
+    \item[Case (incl$_L$)] We have $A <: A$
+      (Prop.~\ref{prop:subtyping} part \ref{prop:⊑-refl}),
+      and therefore $A ∩ B <: A$ by rule (lb$_L$).
+    \item[Case (incl$_R$)] We have $B <: B$
+      (Prop.~\ref{prop:subtyping} part \ref{prop:⊑-refl}),
+      and therefore $A ∩ B <: B$ by rule (lb$_R$).
+    \item[Case (glb)] By the induction hypothesis, we have
+      $A <: C$ and $A <: D$, so we conclude that $A <: C ∩ D$ by (glb).
+    \item[Case $(→)$] By the induction hypothesis, we have $C <: A$
+      and $B <: D$. We conclude that $A → B <: C → D$
+      by Lemma~\ref{lem:⊑-fun′}.
+    \item[Case $({→}{∩})$] We conclude that
+      $(A → B) ∩ (A → C) <: A → (B ∩ C)$
+      by Lemma~\ref{lem:⊑-dist}.
+    \item[Case $(\TOP_{\mathrm{top}})$]
+      We conclude that $A <: \TOP$ by rule $(\TOP_{\mathrm{top}})$.
+    \item[Case $(\TOP{→})$]
+      We have $\topP{\TOP}$, so $\TOP <: C → \TOP$ by rule $(\TOP{→}')$.
+    \end{description}
+
+  \end{description}
+\end{proof}
+
+\section{Conclusion}
+\label{sec:conclude}
+
+In this article we present a new subtyping relation $A <: B$ for
+intersection types that enjoys the subformula property.  None of the
+rules of the new subtyping relation are particularly novel, but the
+fact that we can prove transitivity directly from them is!  We prove
+that the new relation is equivalent to the subtyping relation $A ≤ B$
+of Barendregt, Coppo, and Dezani-Ciancaglini.
+
 \pagebreak
@@
 % LocalWords:  Salle Engeler Honsell Abramsky Plotkin Rocca Alessi
-% LocalWords:  subsumption subtype denotational ISWIM Landin
-% LocalWords:  Felleisen
+% LocalWords:  subsumption subtype denotational ISWIM Landin Ishihara
+% LocalWords:  Felleisen Muehlboeck ISC Agda codomain reflexivity
+% LocalWords:  Bakel

### 2019-06-24 papers/intersection-transitivity.tex
check

@@
 \documentclass{article}
+%\usepackage{cmbright}
 \usepackage{natbib}
@@
 \usepackage{amssymb}
-\usepackage{mathabx}
+%\usepackage{mathabx}
 \usepackage{stmaryrd}
@@
 %\usepackage{fullpage}
-\usepackage{fontspec,unicode-math}
+%\usepackage{fontspec,unicode-math}
 \usepackage{hyperref}
@@
 The BCD intersection type systems extends the simply-typed lambda
-calculus with the addition of intersection types, written $A ∩ B$, a
+calculus with the addition of intersection types, written $A \cap B$, a
 top type $\TOP$, and an infinite collection of type
@@
   \begin{array}{lclr}
-    α,\beta & ::= & \TOP \mid c_0 \mid c_1 \mid c_2 \mid \cdots & \text{atoms}\\
-    A,B,C,D & ::= & α \mid A → B \mid A ∩ B & \text{types}
+    \alpha,\beta & ::= & \TOP \mid c_0 \mid c_1 \mid c_2 \mid \cdots & \text{atoms}\\
+    A,B,C,D & ::= & \alpha \mid A \to B \mid A \cap B & \text{types}
   \end{array}
@@
 states that a term $M$ in environment $\Gamma$ can be given type $B$
-if it has type $A$ and $A$ is a subtype of $B$, written $A ≤ B$.
+if it has type $A$ and $A$ is a subtype of $B$, written $A \leq B$.
 \[
-\inference{\Gamma \vdash M : A & A ≤ B}
+\inference{\Gamma \vdash M : A & A \leq B}
           {\Gamma \vdash M : B}
@@
 rules, and then then prove transitivity~\citep{Muehlboeck:2018aa}.
-Unfortunately, the $({→}{∩})$ rule of the BCD system significantly
+Unfortunately, the $({\to}{\cap})$ rule of the BCD system significantly
 complicates the situation.
@@
 \begin{figure}[tbp]
-  \fbox{$A ≤ B$}
+  \fbox{$A \leq B$}
   \begin{gather*}
-    \text{(refl)} \; \inference{}{A ≤ A} \quad
-    \text{(trans)} \; \inference{A ≤ B & B ≤ C}{A ≤ C} \\[3ex]
-    \text{(incl$_L$)} \; \inference{}{A ∩ B ≤ A} \quad
-    \text{(incl$_R$)} \; \inference{}{A ∩ B ≤ B} \quad
-    \text{(glb)} \; \inference{A ≤ C & A ≤ D}{A ≤ C ∩ D} \\[3ex]
-    (→) \; \inference{C ≤ A & B ≤ D}{A → B ≤ C → D} \quad
-    ({→}{∩}) \; \inference{}{(A → B) ∩ (A → C) ≤ A → (B ∩ C)} \\[3ex]
-    (\TOP_{\mathrm{top}}) \; \inference{}{A ≤ \TOP} \quad
-    (\TOP{→}) \; \inference{}{\TOP ≤ C → \TOP}
+    \text{(refl)} \; \inference{}{A \leq A} \quad
+    \text{(trans)} \; \inference{A \leq B & B \leq C}{A \leq C} \\[3ex]
+    \text{(incl$_L$)} \; \inference{}{A \cap B \leq A} \quad
+    \text{(incl$_R$)} \; \inference{}{A \cap B \leq B} \quad
+    \text{(glb)} \; \inference{A \leq C & A \leq D}{A \leq C \cap D} \\[3ex]
+    (\to) \; \inference{C \leq A & B \leq D}{A \to B \leq C \to D} \quad
+    ({\to}{\cap}) \; \inference{}{(A \to B) \cap (A \to C) \leq A \to (B \cap C)} \\[3ex]
+    (\TOP_{\mathrm{top}}) \; \inference{}{A \leq \TOP} \quad
+    (\TOP{\to}) \; \inference{}{\TOP \leq C \to \TOP}
   \end{gather*}
@@
 $A_1,\ldots,A_n$. The intuition is that $A_1,\ldots,A_n \vdash B$
-corresponds to $A_1 ∩ \cdots ∩ A_n ≤ B$. The ISC system satisfies the
+corresponds to $A_1 \cap \cdots \cap A_n \leq B$. The ISC system satisfies the
 subformula property and is equivalent to the BCD system. To prove
@@
 cut-elimination, from which transitivity follows.
+
+\citet{Bi:2018aa} also present a subtyping system that satisfies the
+subformula property and prove that is equivalent to BCD
+subtyping. Their system is based on the decision procedure
+of~\citet{Pierce:1989aa} and takes the form $\mathcal{L} \vdash A
+\prec: B$, where $\mathcal{L}$ is a queue of types that are peeled off
+from the domain of the type on the right. So the judgement
+$\mathcal{L} \vdash A \prec: B$ is equivalent to $A \leq \mathcal{L}
+\to B$ where $\mathcal{L} \to B$ is defined as
+\begin{align*}
+  [] \to B &= B\\
+  (\mathcal{L},A) \to B &= \mathcal{L} \to (A \to B)
+\end{align*}
+The proof of transitivity for this system is an adaptation of Pierce's
+but it corrects some errors and introduces more lemmas concerning the
+auxilliary notion of reflexive supertypes.
 
@@
 
-%$λ^{\mathrm{BCD}}_∩$
+%$λ^{\mathrm{BCD}}_\cap$
 
@@
 functions, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
-relations $A ∈ B$ and $A ⊆ B$.
+relations $A \in B$ and $A \subseteq B$.
 %
 The $\dom{A}$ and $\cod{A}$ functions return the domain or codomain if
-$A$ is a function type, respectively. If $A$ is an intersection $A_1 ∩
+$A$ is a function type, respectively. If $A$ is an intersection $A_1 \cap
 A_2$, then $\dom{A}$ is the intersection of the domain of $A_1$ and
 $A_2$.  If $A$ is an atom, $\dom{A}$ is undefined. Likewise for
-$\cod{A}$. For example, if $A = (A_1 → B_1) ∩ \cdots ∩ (A_n → B_n)$,
-then $\dom{A} = A_1 ∩ \cdots ∩ A_n$ and $\cod{A} = B_1 ∩ \cdots ∩
+$\cod{A}$. For example, if $A = (A_1 \to B_1) \cap \cdots \cap (A_n \to B_n)$,
+then $\dom{A} = A_1 \cap \cdots \cap A_n$ and $\cod{A} = B_1 \cap \cdots \cap
 B_n$.  When $\dom{A}$ or $\cod{A}$ appears in lemma or theorem
@@
 %
-The relation $A ∈ B$ indicates whether $A$ is syntactically a part of $B$.
-The relation $A ⊆ B$ holds when every part of $A$ is a part of $B$.
-We say that $B$ \emph{contains} $A$ if $A ⊆ B$.
+The relation $A \in B$ indicates whether $A$ is syntactically a part of $B$.
+The relation $A \subseteq B$ holds when every part of $A$ is a part of $B$.
+We say that $B$ \emph{contains} $A$ if $A \subseteq B$.
 
-\begin{proposition}\label{prop:⊔⊆-inv}
- \item If $A ∩ B ⊆ C$, then $A ⊆ C$ and $B ⊆ C$. 
+\begin{proposition}\label{prop:union-subset-inv}
+ \item If $A \cap B \subseteq C$, then $A \subseteq C$ and $B \subseteq C$. 
 \end{proposition}
@@
   \begin{align*}
-  \dom{A → B} &= A \\
-  \dom{A ∩ B} &= \dom{A} ∩ \dom {B} \\
+  \dom{A \to B} &= A \\
+  \dom{A \cap B} &= \dom{A} \cap \dom {B} \\
   \\
-  \cod{A → B} &= B \\
-  \cod{A ∩ B} &= \cod{A} ∩ \cod {B}
+  \cod{A \to B} &= B \\
+  \cod{A \cap B} &= \cod{A} \cap \cod {B}
   \end{align*}
@@
     \quad
-    \inference{\topP{B}}{\topP{A → B}}
+    \inference{\topP{B}}{\topP{A \to B}}
     \quad
-    \inference{\topP{A} & \topP{B}}{\topP{A ∩ B}}
+    \inference{\topP{A} & \topP{B}}{\topP{A \cap B}}
   \end{gather*}
 
-  \fbox{$A ∈ B$}
+  \fbox{$A \in B$}
   \begin{gather*}
-    \inference{}{α ∈ α}  \quad
-    \inference{}{A → B ∈ A → B} \quad
-    \inference{A ∈ B}{A ∈ B ∩ C} \quad
-    \inference{A ∈ C}{A ∈ B ∩ C}
+    \inference{}{\alpha \in \alpha}  \quad
+    \inference{}{A \to B \in A \to B} \quad
+    \inference{A \in B}{A \in B \cap C} \quad
+    \inference{A \in C}{A \in B \cap C}
   \end{gather*}
 
-  \fbox{$A ⊆ B$}
+  \fbox{$A \subseteq B$}
   \[
-     A ⊆ B = ∀ C.\, C ∈ A \text{ implies } C ∈ B
+     A \subseteq B = ∀ C.\, C \in A \text{ implies } C \in B
   \]
@@
   \mathsf{topInCod}(D) =
-     \exists A B.\, A → B ∈ D \text{ and } \mathsf{top}(B)  
+     \exists A B.\, A \to B \in D \text{ and } \mathsf{top}(B)  
   \]
@@
   \begin{gather*}
-    \text{(refl$_α$)} \; \inference{}{α <: α} \\[3ex]
-    \text{(lb$_L$)} \; \inference{A <: C}{A ∩ B <: C} \quad
-    \text{(lb$_R$)} \; \inference{B <: C}{A ∩ B <: C} \quad
-    \text{(glb)} \; \inference{A <: C & A <: D}{A <: C ∩ D} \\[3ex]
-    (→') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C → D}
+    \text{(refl$_\alpha$)} \; \inference{}{\alpha <: \alpha} \\[3ex]
+    \text{(lb$_L$)} \; \inference{A <: C}{A \cap B <: C} \quad
+    \text{(lb$_R$)} \; \inference{B <: C}{A \cap B <: C} \quad
+    \text{(glb)} \; \inference{A <: C & A <: D}{A <: C \cap D} \\[3ex]
+    (\to') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C \to D}
     \begin{array}{l} \neg\, \mathsf{top}(D) \\
       \neg \, \mathsf{topInCod}(B) \\
-      B ⊆ A \end{array}\\[3ex]
+      B \subseteq A \end{array}\\[3ex]
     (\TOP_{\mathrm{top}}) \; \inference{}{A <: \TOP} \quad
-    (\TOP{→}') \; \inference{}{A <: C → D}\;\mathsf{top}(D)
+    (\TOP{\to}') \; \inference{}{A <: C \to D}\;\mathsf{top}(D)
   \end{gather*}
@@
 rule.  It also replaces the (refl) rule with reflexivity for atoms
-(refl$_α$). The most important rule is the one for function types
-$(→')$, which subsumes $(→)$ and $({→}{∩})$ in BCD subtyping.  The
-$(→')$ rule essentially turns the $\beta$-soundness property into a
-subtyping rule. The $(→')$ rule says that a type $A$ is a subtype of a
-function type $C → D$ if a subset of $A$, call it $B$, has domain and
+(refl$_\alpha$). The most important rule is the one for function types
+$(\to')$, which subsumes $(\to)$ and $({\to}{\cap})$ in BCD subtyping.  The
+$(\to')$ rule essentially turns the $\beta$-soundness property into a
+subtyping rule. The $(\to')$ rule says that a type $A$ is a subtype of a
+function type $C \to D$ if a subset of $A$, call it $B$, has domain and
 codomain that are larger and smaller than $C$ and $D$,
@@
 $\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed because of the
-(\TOP{→}') rule, which in turn is needed to preserve types under
+$(\TOP{\to}')$ rule, which in turn is needed to preserve types under
 $\eta$-reduction.  In a system that does not involve $\eta$-reduction,
-the (\TOP{→}') rule can be omitted, as well as those side
+the $(\TOP{\to}')$ rule can be omitted, as well as those side
 conditions. The rules (lb$_L$) and (lb$_R$) adapt (incl$_L$) and
 (incl$_R$) to a system without transitivity, and have appeared many
-times in the literature~\citep{Bakel:1995aa}.  The $(\TOP{→}')$ rule
-generalizes the (\TOP{→}) rule, replacing the $\TOP$ on the left with
+times in the literature~\citep{Bakel:1995aa}.  The $(\TOP{\to}')$ rule
+generalizes the $(\TOP{\to})$ rule, replacing the $\TOP$ on the left with
 any type $A$, because for transitivity, any type is below $\TOP$. The
-$(\TOP{→}')$ rule also replaces the $\TOP$ in the codomain on the
+$(\TOP{\to}')$ rule also replaces the $\TOP$ in the codomain on the
 right with any type $D$ that is equivalent to $\TOP$.
@@
   \item (reflexivity) $A <: A$ \label{prop:⊑-refl}
-  \item If $A <: B ∩ C$, then $A <: B$ and $A <: C$. \label{prop:⊔⊑-inv}
-  \item If $A <: B$ and $C ∈ B$, then $A <: C$.\label{prop:u∈v⊑w→u⊑w}
-  \item If $A <: B$ and $C ⊆ B$, then $A <: C$.\label{prop:u⊆v⊑w→u⊑w}
+  \item If $A <: B \cap C$, then $A <: B$ and $A <: C$. \label{prop:⊔⊑-inv}
+  \item If $A <: B$ and $C \in B$, then $A <: C$.\label{prop:in-sub-sub}
+  \item If $A <: B$ and $C \subseteq B$, then $A <: C$.\label{prop:subset-sub-sub}
   \end{enumerate}
@@
   \item The proof of reflexivity is by induction on $A$. In the case
-    $A = A_1 → A_2$, we proceed by cases on whether $\topP{A_2}$.
-    If it is, deduce $A_1 → A_2 <: A_1 → A_2$ by rule $(\TOP{→}')$.
-    Otherwise, apply rule $(→')$
-  \item The proof is by induction on the derivation of $A <: B ∩ C$.
+    $A = A_1 \to A_2$, we proceed by cases on whether $\topP{A_2}$.
+    If it is, deduce $A_1 \to A_2 <: A_1 \to A_2$ by rule $(\TOP{\to}')$.
+    Otherwise, apply rule $(\to')$
+  \item The proof is by induction on the derivation of $A <: B \cap C$.
   \item The proof is by induction on $B$. In the case where $B = B_1
-    ∩ B_2$, either $C ∈ B_1$ or $C ∈ B_2$, but in either case part
+    \cap B_2$, either $C \in B_1$ or $C \in B_2$, but in either case part
     \ref{prop:⊔⊑-inv} of this proposition fulfills the premise of
     the induction hypothesis, from which the conclusion follows.
-  \item The proof  is by induction on $C$, using part \ref{prop:u∈v⊑w→u⊑w}
+  \item The proof  is by induction on $C$, using part \ref{prop:in-sub-sub}
     of this proposition in the cases for atoms and function types.
@@
   \item If $\topP{A}$ then $\topP{\cod{A}}$.\label{prop:AllBot-cod}
-  \item If $\topP{A}$ and $B ∈ A$, then $\topP{B}$.\label{prop:AllBot-∈}
-  \item If $\topP{A}$ and $B ⊆ A$, then $\topP{B}$.\label{prop:AllBot-⊆}
+  \item If $\topP{A}$ and $B \in A$, then $\topP{B}$.\label{prop:AllBot-in}
+  \item If $\topP{A}$ and $B \subseteq A$, then $\topP{B}$.\label{prop:AllBot-subset}
   \item If $\topP{A}$ and $A <: B$, then $\topP{B}$.\label{prop:AllBot-⊑}
@@
   \item The proof is by induction on $B$. The cases for atoms and
-    function types are proved by part \ref{prop:AllBot-∈} of
+    function types are proved by part \ref{prop:AllBot-in} of
     this proposition. In the case for
-    $B = B_1 ∩ B_2$, from $B_1 ∩ B_2 <: A$, we have
-    $B_1 <: A$ and $B_2 <: A$ (Proposition~\ref{prop:⊔⊆-inv}).
+    $B = B_1 \cap B_2$, from $B_1 \cap B_2 <: A$, we have
+    $B_1 <: A$ and $B_2 <: A$ (Proposition~\ref{prop:union-subset-inv}).
     Then by the induction hypotheses for $B_1$ and $B_2$ we have
     $\topP{B_1}$ and $\topP{B_2}$, from which we conclude
-    that $\topP{B_1 ∩ B_2}$.
+    that $\topP{B_1 \cap B_2}$.
   \item The proof is by induction on the derivation of $A <: B$.
-    All of the cases are straightforward except for rule $(→')$.
-    In that case we have $B = B_1 → B_2$ and some $A'$ such that
-    $A' ⊆ A$, $B_1 <: \dom{A'}$, $\cod{A'} <: B_2$, $\neg\,\topP{B_2}$,
+    All of the cases are straightforward except for rule $(\to')$.
+    In that case we have $B = B_1 \to B_2$ and some $A'$ such that
+    $A' \subseteq A$, $B_1 <: \dom{A'}$, $\cod{A'} <: B_2$, $\neg\,\topP{B_2}$,
     and $\neg\,\topInCod{A'}$. From the premise $\topP{A}$
-    and part \ref{prop:AllBot-⊆} of this proposition, we
+    and part \ref{prop:AllBot-subset} of this proposition, we
     have $\topP{A'}$. Then by part \ref{prop:AllBot-cod}
@@
 Next we turn to the subtyping inversion principle for function types.
-The idea is to generalize the rule $(→')$ with respect to the type on
+The idea is to generalize the rule $(\to')$ with respect to the type on
 the right, allowing any type that contains a function type.  The
-premises of $(→')$ are somewhat complex, so we package most of them
+premises of $(\to')$ are somewhat complex, so we package most of them
 into the following definition.
@@
 \begin{definition}[factors]
-  We say $C → D$ \emph{factors} $A$
+  We say $C \to D$ \emph{factors} $A$
   if there exists some type $B$ such that
-  $B ⊆ A$, $C <: \dom{B}$, $\cod{B} <: D$, and
+  $B \subseteq A$, $C <: \dom{B}$, $\cod{B} <: D$, and
   $\neg\,\mathsf{topInCod}(B)$.
@@
   \label{prop:⊑-fun-inv}
-  If $A <: C'$, $C → D ∈ C'$, and $\neg\,\mathsf{top}(D)$, then
-  $C → D$ factors $A$.
+  If $A <: C'$, $C \to D \in C'$, and $\neg\,\mathsf{top}(D)$, then
+  $C \to D$ factors $A$.
 \end{proposition}
@@
 types. However, it was not needed for our system because the rule
-$(→')$ is already quite close to the inversion principle.
-The lemma states that if every function type $C → D$ in A
-factors $B$, then $\dom{A} → \cod{A}$ also factors $B$.
+$(\to')$ is already quite close to the inversion principle.
+The lemma states that if every function type $C \to D$ in A
+factors $B$, then $\dom{A} \to \cod{A}$ also factors $B$.
 
@@
   \begin{itemize}
-  \item for any $C$ $D$, if $C → D ∈ A$ and $\neg\,\mathsf{top}(D)$,
-    then $C → D$ factors $B$, and
+  \item for any $C$ $D$, if $C \to D \in A$ and $\neg\,\mathsf{top}(D)$,
+    then $C \to D$ factors $B$, and
   \item $\neg\, \mathsf{topInCod}(A)$,
   \end{itemize}
-  then $\dom{A} → \cod{A}$ factors $B$.
+  then $\dom{A} \to \cod{A}$ factors $B$.
 \end{lemma}
@@
   \item Case $A$ is an atom. The statement is vacuously true.
-  \item Case $A = A_1 → A_2$ is a function type. Then we conclude by applying
+  \item Case $A = A_1 \to A_2$ is a function type. Then we conclude by applying
     the premise with $C$ and $D$ instantiated to $A_1$ and $A_2$ respectively.
-  \item Case $A = A_1 ∩ A_2$.  By the induction hypothesis for $A_1$
-    and for $A_2$, we have that $\dom{A_1} → \cod{A_1}$ factors $B$
-    and so does $\dom{A_2} → \cod{A_2}$.  So there exists
-    $B_1$ and $B_2$ such that $B_1 ⊆ B$, $\neg\,\topInCod{B_1}$,
+  \item Case $A = A_1 \cap A_2$.  By the induction hypothesis for $A_1$
+    and for $A_2$, we have that $\dom{A_1} \to \cod{A_1}$ factors $B$
+    and so does $\dom{A_2} \to \cod{A_2}$.  So there exists
+    $B_1$ and $B_2$ such that $B_1 \subseteq B$, $\neg\,\topInCod{B_1}$,
     $\dom{A_1} <: \dom{B_1}$, $\cod{B_1} <: \cod{A_2}$ and similarly
-    for $B_2$. We need to show that $\dom{A} → \cod{A}$ factors
-    $B$. We choose the witness $B_1 ∩ B_2$.  Clearly we have $B_1 ∩
-    B_2 ⊆ B$ and $\neg \topInCod{B_1 ∩ B_2}$.  Also, we have
+    for $B_2$. We need to show that $\dom{A} \to \cod{A}$ factors
+    $B$. We choose the witness $B_1 \cap B_2$.  Clearly we have $B_1 \cap
+    B_2 \subseteq B$ and $\neg \topInCod{B_1 \cap B_2}$.  Also, we have
     \[
-    \dom{B_1} ∩ \dom{B_2} <: \dom{A_1} ∩ \dom{A_2}
+    \dom{B_1} \cap \dom{B_2} <: \dom{A_1} \cap \dom{A_2}
     \]
@@
     \[
-    \cod{A_1} ∩ \cod{A_2} <: \cod{B_1} ∩ \cod{B_2}
+    \cod{A_1} \cap \cod{A_2} <: \cod{B_1} \cap \cod{B_2}
     \]
-    Thus, we have that $\dom{B_1 ∩ B_2} <: \dom{A}$
-    and $\cod{A} <: \cod{B_1 ∩ B_2}$, and this case is complete.
+    Thus, we have that $\dom{B_1 \cap B_2} <: \dom{A}$
+    and $\cod{A} <: \cod{B_1 \cap B_2}$, and this case is complete.
   \end{itemize}
@@
 \begin{align*}
-  \size{α} &= 0 \\
-  \size{A → B} &= 1 + \size{A} + \size{B} \\
-  \size{A ∩ B} &= 1 + \size{A} + \size{B}
+  \size{\alpha} &= 0 \\
+  \size{A \to B} &= 1 + \size{A} + \size{B} \\
+  \size{A \cap B} &= 1 + \size{A} + \size{B}
 \end{align*}
@@
 \begin{align*}
-  \depth{α} &= 0 \\
-  \depth{A → B} &= 1 + \max(\depth{A}, \depth{B}) \\
-  \depth{A ∩ B} &= \max(\depth{A}, \depth{B})
+  \depth{\alpha} &= 0 \\
+  \depth{A \to B} &= 1 + \max(\depth{A}, \depth{B}) \\
+  \depth{A \cap B} &= \max(\depth{A}, \depth{B})
 \end{align*}
@@
   \begin{description}
-  \item[Case (refl$_α$)] We have $B = C = α$.  From the premise $A <:
-    B$ we immediately conclude that $A <: α$.
-  \item[Case (lb$_L$)] So $B = B_1 ∩ B_2$, $B_1 <: C$, and $A <: B_1 ∩
+  \item[Case (refl$_\alpha$)] We have $B = C = \alpha$.  From the premise $A <:
+    B$ we immediately conclude that $A <: \alpha$.
+  \item[Case (lb$_L$)] So $B = B_1 \cap B_2$, $B_1 <: C$, and $A <: B_1 \cap
     B_2$.  We have $A <: B_1$ (Proposition~\ref{prop:subtyping} part
@@
     B, C \rangle$ because $\size{B_1} < \size{B}$.
-  \item[Case (lb$_R$)] So $B = B_1 ∩ B_2$, $B_2 <: C$, and $A <: B_1 ∩
+  \item[Case (lb$_R$)] So $B = B_1 \cap B_2$, $B_2 <: C$, and $A <: B_1 \cap
     B_2$.  We have $A <: B_2$ (Proposition~\ref{prop:subtyping} part
@@
     < \size{B}$.
-  \item[Case (glb)] We have $C = C_1 ∩ C_2$, $B <: C_1$, and $B <:
+  \item[Case (glb)] We have $C = C_1 \cap C_2$, $B <: C_1$, and $B <:
     C_2$.  By the induction hypothesis, we have $A <: C_1$ and $A <:
@@
     \depth{C}$, $\size{C_1} < C$, and $\size{C_2} < C$.
-    We conclude $A <: C_1 ∩ C_2$ by rule (glb).
-  \item[Case $(→')$] So $C = C_1 → C_2$, $\neg \topP{C_2}$, and there
-    exists $B'$ such that $C_1 <: \dom{B'}$, $\cod{B'} <: C_2$, $B' ⊆
-    B$, and $\neg \topInCod{B'}$. From $A <: B$ and $B' ⊆ B$, we have
-    $A <: B'$ (Proposition~\ref{prop:subtyping} part
-    \ref{prop:u⊆v⊑w→u⊑w}). Thus, for any $B_1 → B_2 ∈ B'$, $B_1 → B_2$
-    factors $A$ (Proposition~\ref{prop:⊑-fun-inv}). We have
-    satisfied the premises of Lemma~\ref{lem:sub-inv-trans},
-    so $\dom{B'} → \cod{B'}$ factors $A$. That means there exists
-    $A'$ such that $A' ⊆ A$, $\neg\,\topInCod{A'}$, $\dom{B'} <: \dom{A'}$,
-    and $\cod{A'} <: \cod{B'}$. Then by the induction hypothesis,
-    we have
+    We conclude $A <: C_1 \cap C_2$ by rule (glb).
+  \item[Case $(\to')$] So $C = C_1 \to C_2$, $\neg \topP{C_2}$, and
+    there exists $B'$ such that $C_1 <: \dom{B'}$, $\cod{B'} <: C_2$,
+    $B' \subseteq B$, and $\neg \topInCod{B'}$. From $A <: B$ and $B' \subseteq B$, we
+    have $A <: B'$ (Proposition~\ref{prop:subtyping} part
+    \ref{prop:subset-sub-sub}). Thus, for any $B_1 \to B_2 \in B'$,
+    $B_1 \to B_2$ factors $A$ (Proposition~\ref{prop:⊑-fun-inv}). We
+    have satisfied the premises of Lemma~\ref{lem:sub-inv-trans}, so
+    $\dom{B'} \to \cod{B'}$ factors $A$. That means there exists $A'$
+    such that $A' \subseteq A$, $\neg\,\topInCod{A'}$, $\dom{B'} <: \dom{A'}$,
+    and $\cod{A'} <: \cod{B'}$. Then by the induction hypothesis, we
+    have
     \[
@@
     because $\depth{\cod{A'}} + \depth{C_2} < \depth{A} + \depth{C}$.
-    We conclude that $A <: C_1 → C_2$ by rule $(→')$ witnessed by $A'$.
+    We conclude that $A <: C_1 \to C_2$ by rule $(\to')$ witnessed by $A'$.
   \item[Case $(\TOP_{\mathrm{top}})$] We have $C = \TOP$ and
     conclude $A <: \TOP$ by rule $(\TOP_{\mathrm{top}})$.
-  \item[Case $(\TOP{→}')$] We have $C = C_1 → C_2$ and $\topP{C_2}$.
-    We conclude $A <: C_1 → C_2$ by rule $(\TOP{→}')$.
+  \item[Case $(\TOP{\to}')$] We have $C = C_1 \to C_2$ and $\topP{C_2}$.
+    We conclude $A <: C_1 \to C_2$ by rule $(\TOP{\to}')$.
   \end{description}
@@
 
-Having proved (trans), we next prove $(→)$ and $({→}{∩})$
-and then show that $A <: B$ is equivalent to $A ≤ B$.
+Having proved (trans), we next prove $(\to)$ and $({\to}{\cap})$
+and then show that $A <: B$ is equivalent to $A \leq B$.
 
-\begin{lemma}[$→$]\label{lem:⊑-fun′}
+\begin{lemma}[$\to$]\label{lem:⊑-fun′}
   If $C <: A$ and $B <: D$, then
-  $A → B <: C → D$.
+  $A \to B <: C \to D$.
 \end{lemma}
@@
   \begin{description}
-  \item[Case $\topP{D}$] We conclude $A → B <: C → D$
-    by rule $(\TOP{→}')$.
+  \item[Case $\topP{D}$] We conclude $A \to B <: C \to D$
+    by rule $(\TOP{\to}')$.
   \item[Case $\neg \,\topP{D}$]
@@
     \item[Case $\neg\,\topP{B}$]
-      We conclude that $A → B <: C → D$ by rule $(→')$.
+      We conclude that $A \to B <: C \to D$ by rule $(\to')$.
     \end{description}
@@
 
-\begin{lemma}[${→}{∩}$]\label{lem:⊑-dist}
-  $(A → B) ∩ (A → C) <: A → (B ∩ C)$
+\begin{lemma}[${\to}{\cap}$]\label{lem:⊑-dist}
+  $(A \to B) \cap (A \to C) <: A \to (B \cap C)$
 \end{lemma}
@@
    \item[Case $\topP{B}$ and $\topP{C}$]
-     Then $\topP{B ∩ C}$ and we conclude that
-     $(A → B) ∩ (A → C) <: A → (B ∩ C)$
-     by rule $(\TOP{→}')$.
+     Then $\topP{B \cap C}$ and we conclude that
+     $(A \to B) \cap (A \to C) <: A \to (B \cap C)$
+     by rule $(\TOP{\to}')$.
    \item[Case $\topP{B}$ and $\neg\,\topP{C}$]
      We conclude that 
-     $(A → B) ∩ (A → C) <: A → (B ∩ C)$
-     by rule $(→')$, choosing the witness $A → C$
+     $(A \to B) \cap (A \to C) <: A \to (B \cap C)$
+     by rule $(\to')$, choosing the witness $A \to C$
      and noting that $C <: B$ 
@@
      We conclude that 
-     $(A → B) ∩ (A → C) <: A → (B ∩ C)$
-     by rule $(→')$, this time with witness $A → B$
+     $(A \to B) \cap (A \to C) <: A \to (B \cap C)$
+     by rule $(\to')$, this time with witness $A \to B$
      and noting that
@@
    \item[Case $\neg,\topP{B}$ and $\neg\,\topP{C}$]
-     Again we apply rule $(→')$, but with witness
-     $(A → B) ∩ (A → C)$.
+     Again we apply rule $(\to')$, but with witness
+     $(A \to B) \cap (A \to C)$.
    \end{description}
@@
   \label{lem:dv↦cv<:v}
-  $A ≤ \dom{A} → \cod{A}$.
+  $A \leq \dom{A} \to \cod{A}$.
 \end{lemma}
@@
 \begin{theorem}[Equivalence of the subtyping relations]\ \\
-  $A <: B$ if and only if $A ≤ B$.
+  $A <: B$ if and only if $A \leq B$.
 \end{theorem}
@@
   \begin{description}
-  \item[$A <: B$ implies $A ≤ B$]
+  \item[$A <: B$ implies $A \leq B$]
     We proceed by induction on the derivation of $A <: B$.
     \begin{description}
-    \item[Case (refl$_α$)] We conclude $α ≤ α$ by (refl).      
-    \item[Case (lb$_L$)] By the induction hypothesis we have $A ≤ C$.
-      By (incl$_L$) we have $A ∩ B ≤ A$. We conclude
-      that $A ∩ B ≤ C$ by (trans).
-    \item[Case (lb$_R$)] By the induction hypothesis we have $B ≤ C$.
-      By (incl$_R$) we have $A ∩ B ≤ B$. We conclude
-      that $A ∩ B ≤ C$ by (trans).
+    \item[Case (refl$_\alpha$)] We conclude $\alpha \leq \alpha$ by (refl).      
+    \item[Case (lb$_L$)] By the induction hypothesis we have $A \leq C$.
+      By (incl$_L$) we have $A \cap B \leq A$. We conclude
+      that $A \cap B \leq C$ by (trans).
+    \item[Case (lb$_R$)] By the induction hypothesis we have $B \leq C$.
+      By (incl$_R$) we have $A \cap B \leq B$. We conclude
+      that $A \cap B \leq C$ by (trans).
     \item[Case (glb)] By the induction hypothesis we have
-      $A ≤ C$ and $A ≤ D$, so we conclude that
-      $A ≤ C ∩ D$ by (glb).
-    \item[Case $(→')$] By the induction hypothesis we have
-      $C ≤ \dom{B}$ and also $\cod{B} ≤ D$.
-      From $B ⊆ A$ we have $A ≤ B$.      
+      $A \leq C$ and $A \leq D$, so we conclude that
+      $A \leq C \cap D$ by (glb).
+    \item[Case $(\to')$] By the induction hypothesis we have
+      $C \leq \dom{B}$ and also $\cod{B} \leq D$.
+      From $B \subseteq A$ we have $A \leq B$.      
       Then by Lemma~\ref{lem:dv↦cv<:v} we have
-      $B ≤ \dom{B} → \cod{B}$.
-      Also, we have $\dom{B} → \cod{B} ≤ C → D$ by rule $(→)$.
-      We conclude that $A ≤ C → D$ by chaining the three prior
+      $B \leq \dom{B} \to \cod{B}$.
+      Also, we have $\dom{B} \to \cod{B} \leq C \to D$ by rule $(\to)$.
+      We conclude that $A \leq C \to D$ by chaining the three prior
       facts using (trans).
     \item[Case $(\TOP_{\mathrm{top}})$]
-      We conclude that $A ≤ \TOP$ by $(\TOP_{\mathrm{top}})$.
-    \item[Case $(\TOP{→}')$] We have $A ≤ \TOP$ and $\TOP <: C →
-      \TOP$.  Also, $C → \TOP ≤ C → D$ because $\TOP <: D$ follows
-      from $\topP{D}$.  Thus, applying (trans) we conclude $A ≤ C →
+      We conclude that $A \leq \TOP$ by $(\TOP_{\mathrm{top}})$.
+    \item[Case $(\TOP{\to}')$] We have $A \leq \TOP$ and $\TOP <: C \to
+      \TOP$.  Also, $C \to \TOP \leq C \to D$ because $\TOP <: D$ follows
+      from $\topP{D}$.  Thus, applying (trans) we conclude $A \leq C \to
       D$.
@@
 
-  \item[$A ≤ B$ implies $A <: B$]
-    We proceed by induction on the derivation of $A ≤ B$.
+  \item[$A \leq B$ implies $A <: B$]
+    We proceed by induction on the derivation of $A \leq B$.
     \begin{description}
@@
       (Prop.~\ref{prop:subtyping} part \ref{prop:⊑-refl}),
-      and therefore $A ∩ B <: A$ by rule (lb$_L$).
+      and therefore $A \cap B <: A$ by rule (lb$_L$).
     \item[Case (incl$_R$)] We have $B <: B$
       (Prop.~\ref{prop:subtyping} part \ref{prop:⊑-refl}),
-      and therefore $A ∩ B <: B$ by rule (lb$_R$).
+      and therefore $A \cap B <: B$ by rule (lb$_R$).
     \item[Case (glb)] By the induction hypothesis, we have
-      $A <: C$ and $A <: D$, so we conclude that $A <: C ∩ D$ by (glb).
-    \item[Case $(→)$] By the induction hypothesis, we have $C <: A$
-      and $B <: D$. We conclude that $A → B <: C → D$
+      $A <: C$ and $A <: D$, so we conclude that $A <: C \cap D$ by (glb).
+    \item[Case $(\to)$] By the induction hypothesis, we have $C <: A$
+      and $B <: D$. We conclude that $A \to B <: C \to D$
       by Lemma~\ref{lem:⊑-fun′}.
-    \item[Case $({→}{∩})$] We conclude that
-      $(A → B) ∩ (A → C) <: A → (B ∩ C)$
+    \item[Case $({\to}{\cap})$] We conclude that
+      $(A \to B) \cap (A \to C) <: A \to (B \cap C)$
       by Lemma~\ref{lem:⊑-dist}.
@@
       We conclude that $A <: \TOP$ by rule $(\TOP_{\mathrm{top}})$.
-    \item[Case $(\TOP{→})$]
-      We have $\topP{\TOP}$, so $\TOP <: C → \TOP$ by rule $(\TOP{→}')$.
+    \item[Case $(\TOP{\to})$]
+      We have $\topP{\TOP}$, so $\TOP <: C \to \TOP$ by rule $(\TOP{\to}')$.
     \end{description}
@@
 fact that we can prove transitivity directly from them is!  We prove
-that the new relation is equivalent to the subtyping relation $A ≤ B$
+that the new relation is equivalent to the subtyping relation $A \leq B$
 of Barendregt, Coppo, and Dezani-Ciancaglini.
+
+\section*{Acknowledgments}
+
+This material is based upon work supported by the National Science
+Foundation under Grant No. 1814460.
 

### 2019-06-25 papers/denot-ISWIM.tex
start of denot ISWIM paper

\documentclass{article}
%\usepackage{cmbright}
\usepackage{natbib}
\usepackage{amsthm}
\usepackage{amsmath}
\usepackage{amssymb}
%\usepackage{mathabx}
\usepackage{stmaryrd}
\usepackage{semantic}
%\usepackage{fullpage}
%\usepackage{fontspec,unicode-math}
\usepackage{hyperref}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{corollary}[theorem]{Corollary}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{constraint}[theorem]{Constraint}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\newcommand{\true}[0]{\mathsf{true}}
\newcommand{\curry}[1]{\mathsf{curry}(#1)}
\newcommand{\apply}[2]{\mathsf{apply}(#1, #2)}

\title{An Elementary Denotational \\ Semantics for ISWIM}
\author{Jeremy G. Siek}

\begin{document}
\maketitle

\begin{abstract}
  This article takes a baby step toward developing denotational
  semantics that are suitable for widespread use by programming
  language practitioners and researchers who work outside of
  semantics, e.g, in compiler correctness, static analysis, language
  standards, etc. The current practice in such areas is to use
  operational semantics, and it is quite successful. However, such
  semantics are not compositional, which complicates reasoning about
  them.  Denotational semantics are compositional, but their
  mathematical sophistication has made them too time consuming for
  practitioners to understand and customize to their uses. One purpose
  of this article is to see how little mathematics is necessary to
  develop compositional semantics.

  As a case study, this article develops a denotational semantics for
  the ISWIM programming language, that is, for a call-by-value lambda
  calculus extended with arbitrary primitive operations and literals.
  It models lambda expressions with lookup tables, a simple idea that
  goes back to the graph and filter models for the lambda
  calculus. Unfortunately, those models cannot be used as-is for
  ISWIM: lookup tables are too loose in that they can represent
  arbitrary relations, not just functions. So the article presents a
  way to restrict lookup tables to just functions. The article proves
  the soundness and adequacy of the resulting denotational semantics,
  proves that denotational equality implies contextual equivalence,
  and demostrates uses of the semantics in compiler correctness proofs
  (inlining and closure conversion).
\end{abstract}

\section{Introduction}

\citet{Barendregt:2013aa}

\section{Semantics of ISWIM}

Values
\[
\begin{array}{lcll}
  \mathbb{A} & ::= & \mathbb{Z} \mid \mathbb{B} \mid \cdots & \text{atomic types}\\
  \mathbb{P} & ::= & \mathbb{A} \mid \mathbb{A} \to \mathbb{P} & \text{primitive types}\\
  c & \in & \mathbb{P} & \text{constants} \\
  u,v,w & ::= & c \mid \bot \mid v \mapsto w \mid u \sqcup v & \text{values}
\end{array}
\]

\begin{align*}
  \curry{D}\, \gamma\, (v \mapsto w) &= D \,(\gamma,v)\, w\\
  \curry{D}\, \gamma\, (u \sqcup v) &= D \,\gamma\, u \text{ and } D\, \gamma\, v \\
  \curry{D}\, \gamma\, \bot &= \true \\
\end{align*}

\begin{align*}
  \apply{D}{E}\,\gamma\,w &=
    \exists v.\, D\,\gamma\,(v \mapsto w) \text{ and } E \, \gamma \, v
\end{align*}

\section{Conclusion}

\end{document}

### 2019-06-27 papers/denot-ISWIM.tex
check

@@
 \begin{abstract}
-  This article takes a baby step toward developing denotational
+  This article takes a small step toward developing denotational
   semantics that are suitable for widespread use by programming
@@
   practitioners to understand and customize to their uses. One purpose
-  of this article is to see how little mathematics is necessary to
-  develop compositional semantics.
+  of this article is to see how little sophistication and abstraction
+  is necessary to define a compositional semantics.
 
@@
   calculus extended with arbitrary primitive operations and literals.
-  It models lambda expressions with lookup tables, a simple idea that
-  goes back to the graph and filter models for the lambda
-  calculus. Unfortunately, those models cannot be used as-is for
-  ISWIM: lookup tables are too loose in that they can represent
-  arbitrary relations, not just functions. So the article presents a
-  way to restrict lookup tables to just functions. The article proves
-  the soundness and adequacy of the resulting denotational semantics,
-  proves that denotational equality implies contextual equivalence,
-  and demostrates uses of the semantics in compiler correctness proofs
+  We choose a denotational model based on the graph and filter models
+  for the lambda calculus because of their extreme simplicity; they
+  use lookup tables to represent functions. Unfortunately, these
+  models are not adequate as-is for ISWIM because lookup tables
+  represent arbitrary relations, not just functions. This article
+  presents a simple way to restrict lookup tables so that they only
+  approximate functions. The article proves the soundness and adequacy
+  of the resulting denotational semantics for ISWIM, proves that
+  denotational equality implies contextual equivalence, and
+  demostrates uses of the semantics in compiler correctness proofs
   (inlining and closure conversion).
@@
 
+\bibliographystyle{abbrvnat}
+\bibliography{all}
+
 \end{document}

### 2019-06-27 papers/intersection-transitivity.tex
check

@@
 conjecture that our results apply to other intersection type systems
-as well.
+that include the $({\to}{\cap})$ rule for distributing intersection
+and function types.
 
@@
 Figure~\ref{fig:BCD-subtyping} reviews the BCD rules for subtyping.
-Note that in the (trans) rule, the type $B$ that appears in the
-premises does not appear in the conclusion. Thus, the BCD subtyping
-judgment does not enjoy the subformula property.  For other systems,
-it is straightforward to remove the (trans) rule, modify the other
-rules, and then then prove transitivity~\citep{Muehlboeck:2018aa}.
-Unfortunately, the $({\to}{\cap})$ rule of the BCD system significantly
-complicates the situation.
+Note that in the (trans) rule, the type $B$ appears in the premises
+but not in the conclusion. Thus, the BCD subtyping judgment does not
+enjoy the subformula property.  For many other subtyping systems, it
+is straightforward to remove the (trans) rule, modify the other rules,
+and then then prove transitivity.  Unfortunately, the $({\to}{\cap})$
+rule of the BCD system significantly complicates the situation.
 
@@
 
-The subformula property is a useful one. For example, the author is
-using intersection types to create a denotational semantics for the
-ISWIM language, which includes constants and primitive
+The subformula property is a useful one. For example, the author uses
+intersection types to create denotational semantics (a filter model)
+for the ISWIM language, which includes constants and primitive
 operations~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa}.
-It seems that doing so requires placing extra conditions on types and
-it is much easier to do so when subtyping satisfies the subformula
-property.
+This requires placing extra conditions on types and it is much easier
+to do so when subtyping satisfies the subformula property.
+
+There has been a recent flurry of interest in subtyping for
+intersection types that has led to several systems that satisfy the
+subformula property. We discuss each of them in the next few
+paragraphs.
 
@@
 
-\citet{Bi:2018aa} also present a subtyping system that satisfies the
+\citet{Bi:2018aa} present a subtyping algorithm that satisfies the
 subformula property and prove that is equivalent to BCD
@@
 The proof of transitivity for this system is an adaptation of Pierce's
-but it corrects some errors and introduces more lemmas concerning the
-auxilliary notion of reflexive supertypes.
+but it corrects some errors and introduces more lemmas concerning an
+auxilliary notion called reflexive supertypes.
 
-This article presents a more direct route to the subformula property
-and transitivity. We present a subtyping relation $A <: B$ and
-directly prove transitivity without using an auxiliary sequent-style
-system. Nevertheless, the intuitions are based on those of
-Laurent. The key to $A <: B$ is a rule for function types based on the
-$\beta$-soundness property~\citep{Barendregt:2013aa}, just as in ISC.
-The definitions and results in this article have been machine checked
-in Agda.
+\citet{Muehlboeck:2018aa} develop a framework for obtaining subtyping
+algorithms for systems that include intersection and union types and
+that are extensible to other types. To decide $A \leq B$ their
+algorithm converts $A$ to disjunctive normal form and then applies a
+client-supplied function to each collection of literals. To obtain a
+system equivalent to BCD, including the ${\to}{\cap}$ rule, the client
+side function saturates the collection of literals by applying
+${\to}{\cap}$ (left to right) as much as possible.
+[todo: say something about the proof of transitivity]
+
+This article presents a direct route to the subformula property and
+proof of transitivity for BCD subtyping. We present straightforward
+rules for a new subtyping relation $A <: B$ and directly prove
+transitivity. The key to $A <: B$ is a rule for function types similar
+to the one in Laurent's ISC, which effectively turns the
+$\beta$-soundness property~\citep{Barendregt:2013aa} into a subtyping
+rule. While the rules that define $A <: B$ are not particularly novel,
+the fact that transivity can be proved directly from them is
+suprising.
 
@@
 
+The definitions and results in this article have been machine checked
+in Agda.
+
 %$λ^{\mathrm{BCD}}_\cap$
@@
 
-Our new subtyping judgment relies on several auxiliary notions that
-help us avoid the use of ellipses, which we define in
-Figure~\ref{fig:aux}. These include the $\dom{A}$ and $\cod{A}$
-functions, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
+Our new subtyping judgment relies on a few definitions that are given
+in Figure~\ref{fig:aux}. These include the partial functions $\dom{A}$
+and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
 relations $A \in B$ and $A \subseteq B$.
 %
-The $\dom{A}$ and $\cod{A}$ functions return the domain or codomain if
-$A$ is a function type, respectively. If $A$ is an intersection $A_1 \cap
-A_2$, then $\dom{A}$ is the intersection of the domain of $A_1$ and
-$A_2$.  If $A$ is an atom, $\dom{A}$ is undefined. Likewise for
-$\cod{A}$. For example, if $A = (A_1 \to B_1) \cap \cdots \cap (A_n \to B_n)$,
-then $\dom{A} = A_1 \cap \cdots \cap A_n$ and $\cod{A} = B_1 \cap \cdots \cap
-B_n$.  When $\dom{A}$ or $\cod{A}$ appears in lemma or theorem
-statement, we implicitly assume that $A$ is a type such that $\dom{A}$
-and $\cod{A}$ are defined.
+The $\dom{A}$ and $\cod{A}$ partial functions return the domain or
+codomain if $A$ is a function type, respectively. If $A$ is an
+intersection $A_1 \cap A_2$, then $\dom{A}$ is the intersection of the
+domain of $A_1$ and $A_2$.  If $A$ is an atom, $\dom{A}$ is
+undefined. Likewise for $\cod{A}$. For example, if $A = (A_1 \to B_1)
+\cap \cdots \cap (A_n \to B_n)$, then $\dom{A} = A_1 \cap \cdots \cap
+A_n$ and $\cod{A} = B_1 \cap \cdots \cap B_n$.  When $\dom{A}$ or
+$\cod{A}$ appears in lemma or theorem statement, we implicitly assume
+that $A$ is a type such that $\dom{A}$ and $\cod{A}$ are defined.
 %

### 2020-05-13 papers/intersection-transitivity.tex
starting to revise

@@
 \begin{abstract}
-  The subtyping relation for intersection type systems traditionally
-  employs a transitivity rule (Barendregt et al. 1983), which means
-  that the subtyping judgment does not enjoy the subformula property.
-  Laurent develops a sequent-style subtyping judgment, without
-  transitivity, and proves transitivity via a sequence of six lemmas
-  that culminate in cut-elimination (2018). This article presents a
-  subtyping judgment, in regular style, that satisfies the subformula
-  property, and presents a direct proof of transitivity. Borrowing
-  from Laurent's system, the rule for function types is essentially
-  the $\beta$-soundness property.  The main lemma required for the
-  transitivity proof is one that has been used to prove the inversion
-  principle for subtyping of function types. The choice of induction
-  principle for the proof of transitivity is subtle: we use
-  well-founded induction on the lexicographical ordering of the sum of
-  the depths of the first and last type followed by the sum of the
-  sizes of the middle and last type. The article concludes with a
-  proof that the new subtyping judgment is equivalent to that of
-  Barendregt, Coppo, and Dezani-Ciancaglini.
+  The subtyping rules for intersection type systems traditionally
+  employ a transitivity rule (Barendregt et al. 1983), which means
+  that subtyping does not enjoy the subformula property.  Laurent
+  develops a sequent-style subtyping system, without transitivity, and
+  proves transitivity via a sequence of six lemmas that culminate in
+  cut-elimination (2018). This article develops a subtyping system in
+  regular style that omits transitivity and provides a direct proof of
+  transitivity. The new system satisfies something like the subformula
+  property: every type occuring in the subtyping derivation of $A <:
+  B$ is a subformula of $A$ or $B$, or an intersection of such
+  subformulas.  Borrowing from Laurent's system, the rule for function
+  types is essentially the $\beta$-soundness property.  The main lemma
+  required for the transitivity proof is one that has been used to
+  prove the inversion principle for subtyping of function types. The
+  choice of induction principle for the proof of transitivity is
+  subtle: we use well-founded induction on the lexicographical
+  ordering of the sum of the depths of the first and last type
+  followed by the sum of the sizes of the middle and last type. The
+  article concludes with a proof that the new subtyping judgment is
+  equivalent to that of Barendregt, Coppo, and Dezani-Ciancaglini.
 \end{abstract}
@@
 Salle, as a tool for studying normalization in the lambda
-calculus~\citep{Coppo:1979aa}. By varying the subtyping rules and atom
-types, researchers use intersection type systems to model many
-different
-calculi~\citep{Coppo:1980ab,Coppo:1981aa,Engeler:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Plotkin:1993ab,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa}.
-Perhaps the best-known of them is the BCD intersection type system of
+calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
+been used for at least three purposes: to enhance the expressiveness
+of type systems
+~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
+to increase the precision of static
+analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
+and to express filter models for the denotational semantics of a
+variety of lambda calculi
+~\citep{Coppo:1980ab,Coppo:1981aa,Engeler:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Plotkin:1993ab,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa}.
+
+UNDER CONSTRUCTION
+
+A filter model
+
+\[
+   \llbracket M \rrbracket = \{ A \mid \emptyset \vdash M : A \}
+\]
+
+Perhaps the best-known  them is the BCD intersection type system of
 \citet{Barendregt:1983aa}. For this article we focus on the BCD

### 2020-05-14 papers/intersection-transitivity.tex
check

@@
   regular style that omits transitivity and provides a direct proof of
-  transitivity. The new system satisfies something like the subformula
-  property: every type occuring in the subtyping derivation of $A <:
-  B$ is a subformula of $A$ or $B$, or an intersection of such
-  subformulas.  Borrowing from Laurent's system, the rule for function
-  types is essentially the $\beta$-soundness property.  The main lemma
-  required for the transitivity proof is one that has been used to
-  prove the inversion principle for subtyping of function types. The
-  choice of induction principle for the proof of transitivity is
-  subtle: we use well-founded induction on the lexicographical
-  ordering of the sum of the depths of the first and last type
-  followed by the sum of the sizes of the middle and last type. The
-  article concludes with a proof that the new subtyping judgment is
-  equivalent to that of Barendregt, Coppo, and Dezani-Ciancaglini.
+  transitivity, significantly reducing the length of the proof. The
+  new system satisfies something like the subformula property: every
+  type occuring in the derivation of $A <: B$ is a subformula of $A$
+  or $B$, or an intersection of such subformulas.  Borrowing from
+  Laurent's system, the rule for function types is essentially the
+  $\beta$-soundness property.  The main lemma required for the
+  transitivity proof is one that has been used to prove the inversion
+  principle for subtyping of function types. The article concludes
+  with a proof that the new subtyping system is equivalent to that of
+  Barendregt, Coppo, and Dezani-Ciancaglini.
 \end{abstract}
@@
 Intersection types were invented by Coppo, Dezani-Ciancaglini, and
-Salle, as a tool for studying normalization in the lambda
+Salle as a tool for studying normalization in the lambda
 calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
@@
 
-UNDER CONSTRUCTION
+The work in this
 

### 2020-05-14 papers/LMCS-subtyping-intersections/paper.tex
rename

--- 2020-05-14 papers/intersection-transitivity.tex
@@
 is straightforward to remove the (trans) rule, modify the other rules,
-and then then prove transitivity.  Unfortunately, the $({\to}{\cap})$
+and then prove transitivity.  Unfortunately, the $({\to}{\cap})$
 rule of the BCD system significantly complicates the situation.
@@
 
-Before moving on, we make note of some basic facts regarding the $<:$
+Before moving on, we take note of some basic facts regarding the $<:$
 relation and the $\topP{A}$ predicate.

### 2020-05-14 papers/LMCS-subtyping-intersections/paper.tex
finished draft of new motivation section

@@
   employ a transitivity rule (Barendregt et al. 1983), which means
-  that subtyping does not enjoy the subformula property.  Laurent
-  develops a sequent-style subtyping system, without transitivity, and
-  proves transitivity via a sequence of six lemmas that culminate in
-  cut-elimination (2018). This article develops a subtyping system in
-  regular style that omits transitivity and provides a direct proof of
-  transitivity, significantly reducing the length of the proof. The
-  new system satisfies something like the subformula property: every
-  type occuring in the derivation of $A <: B$ is a subformula of $A$
-  or $B$, or an intersection of such subformulas.  Borrowing from
-  Laurent's system, the rule for function types is essentially the
-  $\beta$-soundness property.  The main lemma required for the
-  transitivity proof is one that has been used to prove the inversion
-  principle for subtyping of function types. The article concludes
-  with a proof that the new subtyping system is equivalent to that of
-  Barendregt, Coppo, and Dezani-Ciancaglini.
+  that subtyping does not satisfy the subformula property, making it
+  more difficult to use in filter models for compiler verification.
+  Laurent develops a sequent-style subtyping system, without
+  transitivity, and proves transitivity via a sequence of six lemmas
+  that culminate in cut-elimination (2018). This article develops a
+  subtyping system in regular style that omits transitivity and
+  provides a direct proof of transitivity, significantly reducing the
+  length of the proof. The new system satisfies something like the
+  subformula property: every type occuring in the derivation of $A <:
+  B$ is a subformula of $A$ or $B$, or an intersection of such
+  subformulas.  Borrowing from Laurent's system, the rule for function
+  types is essentially the $\beta$-soundness property.  The main lemma
+  required for the transitivity proof is one that has been used to
+  prove the inversion principle for subtyping of function types. The
+  article concludes with a proof that the new subtyping system is
+  equivalent to that of Barendregt, Coppo, and Dezani-Ciancaglini.
 \end{abstract}
@@
 Intersection types were invented by Coppo, Dezani-Ciancaglini, and
-Salle as a tool for studying normalization in the lambda
-calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
-been used for at least three purposes: to enhance the expressiveness
-of type systems
-~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
-to increase the precision of static
+Salle to study normalization in the lambda
+calculus~\citep{Coppo:1979aa}. Subsequently intersection type systems
+have been used for at least three purposes:
+\begin{itemize}
+  \item as type systems for programming
+    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
+  \item to formulate precise static
     analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
-and to express filter models for the denotational semantics of a
-variety of lambda calculi
-~\citep{Coppo:1980ab,Coppo:1981aa,Engeler:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Plotkin:1993ab,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa}.
+    and
+  \item to express filter models for the denotational semantics of a
+    variety of lambda
+    calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
+\end{itemize}
 
-The work in this
+There are many intersection type systems; perhaps the best-known them
+is the BCD intersection type system of \citet{Barendregt:1983aa}. For
+this article we focus on the BCD system, following the presentation of
+\citet{Barendregt:2013aa}. We conjecture that our results apply to
+other intersection type systems that include the $({\to}{\cap})$ rule
+for distributing intersection and function types.
 
-A filter model
-
-\[
-   \llbracket M \rrbracket = \{ A \mid \emptyset \vdash M : A \}
-\]
-
-Perhaps the best-known  them is the BCD intersection type system of
-\citet{Barendregt:1983aa}. For this article we focus on the BCD
-system, following the presentation of \citet{Barendregt:2013aa}.  We
-conjecture that our results apply to other intersection type systems
-that include the $({\to}{\cap})$ rule for distributing intersection
-and function types.
-
-The BCD intersection type systems extends the simply-typed lambda
-calculus with the addition of intersection types, written $A \cap B$, a
-top type $\TOP$, and an infinite collection of type
-constants. Figure~\ref{fig:types} defines the grammar of types.
+The BCD intersection type system includes function types, $A \to B$,
+intersection types, $A \cap B$, a top type, $\TOP$, and an infinite
+collection of type constants. Figure~\ref{fig:types} defines the
+grammar of types.
 
@@
 \]
-Figure~\ref{fig:BCD-subtyping} reviews the BCD rules for subtyping.
-Note that in the (trans) rule, the type $B$ appears in the premises
-but not in the conclusion. Thus, the BCD subtyping judgment does not
-enjoy the subformula property.  For many other subtyping systems, it
-is straightforward to remove the (trans) rule, modify the other rules,
-and then prove transitivity.  Unfortunately, the $({\to}{\cap})$
-rule of the BCD system significantly complicates the situation.
+Figure~\ref{fig:BCD-subtyping} reviews the BCD subtyping system.  Note
+that in the (trans) rule, the type $B$ appears in the premises but not
+in the conclusion. Thus, the BCD subtyping system does not satisfy the
+subformula property.  For many other subtyping systems, it is
+straightforward to remove the (trans) rule, modify the other rules,
+and then prove transitivity.  Unfortunately, the $({\to}{\cap})$ rule
+of the BCD system significantly complicates the situation.
 
@@
 
-The subformula property is a useful one. For example, the author uses
-intersection types to create denotational semantics (a filter model)
-for the ISWIM language, which includes constants and primitive
-operations~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa}.
-This requires placing extra conditions on types and it is much easier
-to do so when subtyping satisfies the subformula property.
+\section{Motivations from Verified Compilation}
+
+Our interest in intersection type systems stems from our work on
+mechanized proofs of correctness for compiling functional programming
+languages. We are exploring whether the use of denotational semantics
+can streamline such proofs compared to the operational techniques,
+e.g., used to prove the correctness of
+CakeML~\citep{Kumar:2014aa,Owens:2017aa}. In particular, we are
+constructing filter models (based on intersection type systems)
+because they are relatively straightforward to mechanize in a proof
+assistant compared to traditional domain
+theory~\citep{Benton:2009ab,Dockins:2014aa}.  In a filter model, the
+meaning of a program $M$ is the set of all types that can be assigned
+to it by an intersection type system:
+\[
+   \llbracket M \rrbracket = \{ A \mid \emptyset \vdash M : A \}
+\]
+Most type systems are much too coarse of an abstraction to be used as
+filter models. Intersection type systems are special in that they can
+completely characterize the runtime behavior of a program.  To catch a
+glimpse of how filter models work, consider the meaning of the
+identity function in the $\lambda$-calculus.
+\[
+  \llbracket \lambda x.\, x \rrbracket =
+      \{ c_0 \to c_0, c_1 \to c_1, c_2 \to c_2, \ldots \}
+\]
+If one takes each type constant $c_i$ to be a singleton type for the
+number $i$, then the set of types assigned to the identity function
+starts to look like the graph of the function. Indeed, filter models
+are closely related to graph
+models~\citep{Scott:1976lq,Engeler:1981aa,Plotkin:1993ab}.
+
+One of the standard compiler transformations that we are interested in
+verifying is common subexpression elimination. For example, this
+transformation would replace the duplicated $(f \; x)$ terms in the
+following
+\[
+  M = \lambda f.\, \lambda h.\, \lambda x.\, h \; (f \; x) \; (f \; x)
+\]
+with a single instance of $(f\;x)$ that is let-bound to a variable $y$:
+\[
+M' = \lambda f.\, \lambda h.\, \lambda x.\,
+\mathsf{let}\; y = (f \; x) \;\mathsf{in}\;
+h \; y \; y
+\]
+Unfortunately, $M$ and $M'$ are not denotationally equal according to
+most filter models because filter models allow function graphs that
+represent arbitrary relations. The above transformation depends on $(f
+\; x)$ producing the same result every time, but a filter model
+associates arbitrary graphs with parameter $f$.
+
+This problem is not conceptually difficult to solve; one just needs to
+restrict the graphs to only allow approximations of functions. This
+can be accomplished by requiring every entry in a function's graph to
+be ``consistent''. Two function types, $A \to B$ and $A' \to B'$, are
+consistent, written $(A \to B) \sim (A' \to B')$, when either $A
+\not\sim A'$ or $A \sim A'$ and $B \sim B'$. Two type constants are
+consistent when they are equal. One would then hope to prove that for
+any term $M$, if $\Gamma \vdash M : C$ and $\Gamma \vdash M : C'$,
+then $C \sim C'$ as long as $C$, $C'$, and $\Gamma$ are self
+consistent. A key lemma needed for that proof is that consistency is
+upward closed with respect to subtyping: if $A \sim B$, $A <: C$, $B
+<: D$, then $C \sim D$.
+
+This brings us back to the transitivity rule and the subformula
+property. The proof of this lemma hits a snag in the cases where $A <:
+C$ and $B <: D$ are derived using the transitivity rule. The
+intermediate type is not guaranteed to be self consistent. We explored
+adding this guarantee to the transitivity rule, but it caused a
+significant increase in obligations elsewhere in the proof. By some
+serendipity, around that moment in time \citet{Laurent:2018aa}
+published the first subtyping system for intersection types that
+removes the transitivity rule and satisfies the subformula property.
+Laurent developed a sequent-style subtyping system, written $\Gamma
+\vdash B$, where $\Gamma$ is a sequence of types $A_1,\ldots,A_n$. The
+intuition is that $A_1,\ldots,A_n \vdash B$ corresponds to $A_1 \cap
+\cdots \cap A_n \leq B$. This system satisfies the subformula property
+and is equivalent to the BCD system. To prove this, Laurent
+establishes six lemmas that culminate in cut-elimination, from which
+transitivity follows. Laurent mechanized these proofs in Coq.
+
+We immediately took advantage of Laurent's result to complete a filter
+model for the ISWIM
+language~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa}.
+We mechanized the formal development and proofs in
+Isabelle~\citep{Siek:2018aa}.
+
+A year later we began to port this filter model to the Agda proof
+assistant, motivated by the need to use dependent types to make our
+development more reusable: we needed several variations on the filter
+model to give semantics to different languages, including ones that
+are call-by-value and call-by-name, and to the different intermediate
+languages used in our compiler. We found it tedious to repeat
+Laurent's proof of transitivity for each of these variations, and in
+the process developed an intuition that the sequent-style system was
+not necessary. The present article discusses how we were able to
+extract the key feature from Laurent's system, the rule for subtyping
+between function types (which effectively turns the $\beta$-soundness
+property~\citep{Barendregt:2013aa} into a subtyping rule) and splice
+it into the BCD subtyping system, replacing the function $(\to)$,
+distributivity $({\to}{\cap})$, and transitivity (trans) rules.  We
+then prove transitivity directly, without the sequence of six lemmas,
+instead using one lemma from the literature on inversion principles
+for subtyping of function types~\citep{Barendregt:2013aa}. The end
+product is a subtyping system that highly suitable for mechanizing
+filter models.
+
+We discuss two recent developments regarding subtyping for
+intersection types in Section~\ref{sec:other-developments}.  We
+present our new subtyping system in Section~\ref{sec:new-subtyping},
+prove transitivity in Section~\ref{sec:trans}, and prove its
+equivalence to BCD subtyping in Section~\ref{sec:equiv}. We make some
+concluding remarks in Section~\ref{sec:conclude}.
+
+The definitions and results in this article have been machine checked
+in Agda, in the file \texttt{agda/TransSubInter.agda} in the following
+repository.
+
+\begin{center}
+  \url{https://github.com/jsiek/denotational_semantics}
+\end{center}
+
+\section{Other Developments}
+\label{sec:other-developments}
 
 There has been a recent flurry of interest in subtyping for
-intersection types that has led to several systems that satisfy the
-subformula property. We discuss each of them in the next few
-paragraphs.
-
-\citet{Laurent:2018aa} introduces the ISC sequent-style system,
-written $\Gamma \vdash B$, where $\Gamma$ is a sequence of types
-$A_1,\ldots,A_n$. The intuition is that $A_1,\ldots,A_n \vdash B$
-corresponds to $A_1 \cap \cdots \cap A_n \leq B$. The ISC system satisfies the
-subformula property and is equivalent to the BCD system. To prove
-this, Laurent establishes six lemmas that culminate in
-cut-elimination, from which transitivity follows.
+intersection types that has led to two more systems that satisfy the
+subformula property, in addition to the one by \citet{Laurent:2018aa}
+discussed above. These two systems were developed for the purposes of
+type checking programming languages, and not for the construction of
+filter models, so the goals are somewhat different.
 
 \citet{Bi:2018aa} present a subtyping algorithm that satisfies the
-subformula property and prove that is equivalent to BCD
+subformula property and proves that it is equivalent to BCD
 subtyping. Their system is based on the decision procedure
@@
 ${\to}{\cap}$ (left to right) as much as possible.
-[todo: say something about the proof of transitivity]
-
-This article presents a direct route to the subformula property and
-proof of transitivity for BCD subtyping. We present straightforward
-rules for a new subtyping relation $A <: B$ and directly prove
-transitivity. The key to $A <: B$ is a rule for function types similar
-to the one in Laurent's ISC, which effectively turns the
-$\beta$-soundness property~\citep{Barendregt:2013aa} into a subtyping
-rule. While the rules that define $A <: B$ are not particularly novel,
-the fact that transivity can be proved directly from them is
-suprising.
-
-We describe the new subtyping relation in
-Section~\ref{sec:new-subtyping}, prove transitivity in
-Section~\ref{sec:trans}, and prove its equivalence to BCD subtyping in
-Section~\ref{sec:equiv}. We make some concluding remarks in
-Section~\ref{sec:conclude}.
-
-The definitions and results in this article have been machine checked
-in Agda.
+%[todo: say something about the proof of transitivity]
 

### 2020-05-14 papers/LMCS-subtyping-intersections/paper.tex
revisions

@@
 \newcommand{\TOP}{\ensuremath{\mathtt{U}}}
-\newcommand{\dom}[1]{\mathsf{dom}(#1)}
-\newcommand{\cod}[1]{\mathsf{cod}(#1)}
+\newcommand{\dom}[1]{\cap\mathsf{dom}(#1)}
+\newcommand{\cod}[1]{\cap\mathsf{cod}(#1)}
 \newcommand{\topP}[1]{\mathsf{top}(#1)}
@@
 \newcommand{\size}[1]{\mathsf{size}(#1)}
+\newcommand{\inside}[0]{\inplus}
+\newcommand{\contains}[0]{\subsetpluseq}
 
@@
 language~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa}.
-We mechanized the formal development and proofs in
-Isabelle~\citep{Siek:2018aa}.
+We mechanized the formal development and proofs using the Isabelle
+proof assistant~\citep{Siek:2018aa}.
 
 A year later we began to port this filter model to the Agda proof
-assistant, motivated by the need to use dependent types to make our
-development more reusable: we needed several variations on the filter
-model to give semantics to different languages, including ones that
-are call-by-value and call-by-name, and to the different intermediate
-languages used in our compiler. We found it tedious to repeat
-Laurent's proof of transitivity for each of these variations, and in
-the process developed an intuition that the sequent-style system was
-not necessary. The present article discusses how we were able to
-extract the key feature from Laurent's system, the rule for subtyping
-between function types (which effectively turns the $\beta$-soundness
-property~\citep{Barendregt:2013aa} into a subtyping rule) and splice
-it into the BCD subtyping system, replacing the function $(\to)$,
-distributivity $({\to}{\cap})$, and transitivity (trans) rules.  We
-then prove transitivity directly, without the sequence of six lemmas,
-instead using one lemma from the literature on inversion principles
-for subtyping of function types~\citep{Barendregt:2013aa}. The end
-product is a subtyping system that highly suitable for mechanizing
-filter models.
+assistant, motivated by the availability of dependent types to make
+our development more reusable: we needed several variations on the
+filter model to give semantics to different languages, including ones
+that are call-by-value and call-by-name, and to the different
+intermediate languages used in our compiler. We found it tedious to
+repeat Laurent's proof of transitivity for each of these variations,
+and in the process developed an intuition that the sequent-style
+system was not necessary. The present article discusses how we were
+able to extract the key feature from Laurent's system, the rule for
+subtyping between function types and adapt the BCD subtyping system,
+replacing the function $(\to)$, distributivity $({\to}{\cap})$, and
+transitivity (trans) rules.  We then prove transitivity directly,
+without the sequence of six lemmas, instead using one lemma that has
+been used to prove the inversion principle for subtyping of function
+types~\citep{Barendregt:2013aa}. The new subtyping system does not
+technically satisfy the subformula property but it satisfies a similar
+property: every type occuring in the derivation of $A <: B$ is a
+subformula of $A$ or $B$ or an intersection of such subformulas.
+This property is enough to ensure that if $A$ and $B$ are self
+consistent, then so are all the types that appear in the derivation of
+$A <: B$. As a result, the new subtyping system is highly suitable for
+mechanizing filter models.
 
@@
 and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
-relations $A \in B$ and $A \subseteq B$.
+relations $A \inside B$ and $A \contains B$.
 %
@@
 %
-The relation $A \in B$ indicates whether $A$ is syntactically a part of $B$.
-The relation $A \subseteq B$ holds when every part of $A$ is a part of $B$.
-We say that $B$ \emph{contains} $A$ if $A \subseteq B$.
+The relations $A \inside B$ and $A \contains B$ enable the treatment
+of a sequence of intersections $B_1 \cap \cdots \cap B_n$ (modulo
+associativity and commutativity) as if it was set of types $\{ B_1,
+\ldots, B_n \}$, where each $B_i$ is an atom or a function type.  We
+say that $A$ is a part of $B$ if $A \inside B$ and we say that $B$
+\emph{contains} $A$ if $A \contains B$.
 
 \begin{proposition}\label{prop:union-subset-inv}
- \item If $A \cap B \subseteq C$, then $A \subseteq C$ and $B \subseteq C$. 
+ \item If $A \cap B \contains C$, then $A \contains C$ and $B \contains C$. 
 \end{proposition}
@@
 
+  \fbox{$A \inside B$}
+  \begin{gather*}
+    \inference{}{\alpha \inside \alpha}  \quad
+    \inference{}{A \to B \inside A \to B} \quad
+    \inference{A \inside B}{A \inside B \cap C} \quad
+    \inference{A \inside C}{A \inside B \cap C}
+  \end{gather*}
+
+  \fbox{$A \contains B$}
+  \[
+     A \contains B = \forall C.\, C \inside A \text{ implies } C \inside B
+  \]
+
   \fbox{$\topP{A}$}
@@
 
-  \fbox{$A \in B$}
-  \begin{gather*}
-    \inference{}{\alpha \in \alpha}  \quad
-    \inference{}{A \to B \in A \to B} \quad
-    \inference{A \in B}{A \in B \cap C} \quad
-    \inference{A \in C}{A \in B \cap C}
-  \end{gather*}
-
-  \fbox{$A \subseteq B$}
-  \[
-     A \subseteq B = ∀ C.\, C \in A \text{ implies } C \in B
-  \]
-
   \fbox{$\mathsf{topInCod}(D)$}
@@
   \mathsf{topInCod}(D) =
-     \exists A B.\, A \to B \in D \text{ and } \mathsf{top}(B)  
+     \exists A B.\, A \to B \inside D \text{ and } \mathsf{top}(B)  
   \]
@@
     (\to') \; \inference{C <: \dom{B} & \cod{B} <: D }{A <: C \to D}
-    \begin{array}{l} \neg\, \mathsf{top}(D) \\
-      \neg \, \mathsf{topInCod}(B) \\
-      B \subseteq A \end{array}\\[3ex]
+    \begin{array}{l}
+      B \contains A\\
+      \neg\, \mathsf{top}(D) \\
+      \neg \, \mathsf{topInCod}(B)
+      \end{array}\\[3ex]
     (\TOP_{\mathrm{top}}) \; \inference{}{A <: \TOP} \quad
@@
 (refl$_\alpha$). The most important rule is the one for function types
-$(\to')$, which subsumes $(\to)$ and $({\to}{\cap})$ in BCD subtyping.  The
-$(\to')$ rule essentially turns the $\beta$-soundness property into a
-subtyping rule. The $(\to')$ rule says that a type $A$ is a subtype of a
-function type $C \to D$ if a subset of $A$, call it $B$, has domain and
-codomain that are larger and smaller than $C$ and $D$,
-respectively. The use of a subset of $A$ enables this rule to absorb
-uses of (incl$_L$) and (incl$_R$) on the left.  The side conditions
-$\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed because of the
-$(\TOP{\to}')$ rule, which in turn is needed to preserve types under
-$\eta$-reduction.  In a system that does not involve $\eta$-reduction,
-the $(\TOP{\to}')$ rule can be omitted, as well as those side
+$(\to')$, which subsumes $(\to)$ and $({\to}{\cap})$ in BCD subtyping.
+The $(\to')$ rule essentially turns the $\beta$-soundness property
+into a subtyping rule~\citep{Barendregt:2013aa}. The $(\to')$ rule
+says that a type $A$ is a subtype of a function type $C \to D$ if a
+type contained in $A$, call it $B$, has domain and codomain that are
+larger and smaller than $C$ and $D$, respectively. The use of
+containment enables this rule to absorb uses of (incl$_L$) and
+(incl$_R$) on the left.  The side conditions $\neg\;\topP{B}$ and
+$\neg\;\topInCod{D}$ are needed because of the $(\TOP{\to}')$ rule,
+which in turn is needed to preserve types under $\eta$-reduction.  In
+the many systems that do not involve $\eta$-reduction, the
+$(\TOP{\to}')$ rule can be omitted, as well as these side
 conditions. The rules (lb$_L$) and (lb$_R$) adapt (incl$_L$) and
@@
 times in the literature~\citep{Bakel:1995aa}.  The $(\TOP{\to}')$ rule
-generalizes the $(\TOP{\to})$ rule, replacing the $\TOP$ on the left with
-any type $A$, because for transitivity, any type is below $\TOP$. The
-$(\TOP{\to}')$ rule also replaces the $\TOP$ in the codomain on the
-right with any type $D$ that is equivalent to $\TOP$.
+generalizes the $(\TOP{\to})$ rule, replacing the $\TOP$ on the left
+with any type $A$, because for transitivity, any type is below
+$\TOP$. The $(\TOP{\to}')$ rule replaces the $\TOP$ in the
+codomain on the right with any type $D$ that is equivalent to $\TOP$.
 
@@
   \item If $A <: B \cap C$, then $A <: B$ and $A <: C$. \label{prop:⊔⊑-inv}
-  \item If $A <: B$ and $C \in B$, then $A <: C$.\label{prop:in-sub-sub}
-  \item If $A <: B$ and $C \subseteq B$, then $A <: C$.\label{prop:subset-sub-sub}
+  \item If $A <: B$ and $C \inside B$, then $A <: C$.\label{prop:in-sub-sub}
+  \item If $A <: B$ and $C \contains B$, then $A <: C$.\label{prop:subset-sub-sub}
   \end{enumerate}
@@
   \item The proof is by induction on $B$. In the case where $B = B_1
-    \cap B_2$, either $C \in B_1$ or $C \in B_2$, but in either case part
+    \cap B_2$, either $C \inside B_1$ or $C \inside B_2$, but in either case part
     \ref{prop:⊔⊑-inv} of this proposition fulfills the premise of
@@
   \item If $\topP{A}$ then $\topP{\cod{A}}$.\label{prop:AllBot-cod}
-  \item If $\topP{A}$ and $B \in A$, then $\topP{B}$.\label{prop:AllBot-in}
-  \item If $\topP{A}$ and $B \subseteq A$, then $\topP{B}$.\label{prop:AllBot-subset}
+  \item If $\topP{A}$ and $B \inside A$, then $\topP{B}$.\label{prop:AllBot-in}
+  \item If $\topP{A}$ and $B \contains A$, then $\topP{B}$.\label{prop:AllBot-subset}
   \item If $\topP{A}$ and $A <: B$, then $\topP{B}$.\label{prop:AllBot-⊑}
@@
     In that case we have $B = B_1 \to B_2$ and some $A'$ such that
-    $A' \subseteq A$, $B_1 <: \dom{A'}$, $\cod{A'} <: B_2$, $\neg\,\topP{B_2}$,
+    $A' \contains A$, $B_1 <: \dom{A'}$, $\cod{A'} <: B_2$, $\neg\,\topP{B_2}$,
     and $\neg\,\topInCod{A'}$. From the premise $\topP{A}$
@@
   if there exists some type $B$ such that
-  $B \subseteq A$, $C <: \dom{B}$, $\cod{B} <: D$, and
+  $B \contains A$, $C <: \dom{B}$, $\cod{B} <: D$, and
   $\neg\,\mathsf{topInCod}(B)$.
@@
 \begin{proposition}[Inversion Principle for Function Types]
-  \label{prop:⊑-fun-inv}
-  If $A <: C'$, $C \to D \in C'$, and $\neg\,\mathsf{top}(D)$, then
+  \label{prop:⊑-fun-inv}\ \\
+  If $A <: B$, $C \to D \inside B$, and $\neg\,\mathsf{top}(D)$, then
   $C \to D$ factors $A$.
@@
 \begin{proof}
-  The proof is a straightforward induction on $A <: C'$.
+  The proof is a straightforward induction on $A <: B$.
 \end{proof}
@@
   \begin{itemize}
-  \item for any $C$ $D$, if $C \to D \in A$ and $\neg\,\mathsf{top}(D)$,
+  \item for any $C$ $D$, if $C \to D \inside A$ and $\neg\,\mathsf{top}(D)$,
     then $C \to D$ factors $B$, and
@@
     and so does $\dom{A_2} \to \cod{A_2}$.  So there exists
-    $B_1$ and $B_2$ such that $B_1 \subseteq B$, $\neg\,\topInCod{B_1}$,
+    $B_1$ and $B_2$ such that $B_1 \contains B$, $\neg\,\topInCod{B_1}$,
     $\dom{A_1} <: \dom{B_1}$, $\cod{B_1} <: \cod{A_2}$ and similarly
@@
     $B$. We choose the witness $B_1 \cap B_2$.  Clearly we have $B_1 \cap
-    B_2 \subseteq B$ and $\neg \topInCod{B_1 \cap B_2}$.  Also, we have
+    B_2 \contains B$ and $\neg \topInCod{B_1 \cap B_2}$.  Also, we have
     \[
@@
     there exists $B'$ such that $C_1 <: \dom{B'}$, $\cod{B'} <: C_2$,
-    $B' \subseteq B$, and $\neg \topInCod{B'}$. From $A <: B$ and $B' \subseteq B$, we
+    $B' \contains B$, and $\neg \topInCod{B'}$. From $A <: B$ and $B' \contains B$, we
     have $A <: B'$ (Proposition~\ref{prop:subtyping} part
-    \ref{prop:subset-sub-sub}). Thus, for any $B_1 \to B_2 \in B'$,
+    \ref{prop:subset-sub-sub}). Thus, for any $B_1 \to B_2 \inside B'$,
     $B_1 \to B_2$ factors $A$ (Proposition~\ref{prop:⊑-fun-inv}). We
@@
     $\dom{B'} \to \cod{B'}$ factors $A$. That means there exists $A'$
-    such that $A' \subseteq A$, $\neg\,\topInCod{A'}$, $\dom{B'} <: \dom{A'}$,
+    such that $A' \contains A$, $\neg\,\topInCod{A'}$, $\dom{B'} <: \dom{A'}$,
     and $\cod{A'} <: \cod{B'}$. Then by the induction hypothesis, we
@@
       $C \leq \dom{B}$ and also $\cod{B} \leq D$.
-      From $B \subseteq A$ we have $A \leq B$.      
+      From $B \contains A$ we have $A \leq B$.      
       Then by Lemma~\ref{lem:dv↦cv<:v} we have
@@
 In this article we present a new subtyping relation $A <: B$ for
-intersection types that enjoys the subformula property.  None of the
-rules of the new subtyping relation are particularly novel, but the
-fact that we can prove transitivity directly from them is!  We prove
-that the new relation is equivalent to the subtyping relation $A \leq B$
-of Barendregt, Coppo, and Dezani-Ciancaglini.
+intersection types that satisfies a property similar to the subformula
+property.  None of the rules of the new subtyping relation are
+particularly novel, but the fact that we can prove transitivity
+directly from is surprising!  We prove that the new relation is
+equivalent to the subtyping relation $A \leq B$ of Barendregt, Coppo,
+and Dezani-Ciancaglini.
 

### 2020-05-14 papers/LMCS-subtyping-intersections/response.txt
progress on response

Reviewer 1
==========

>> The introduction is rather short and only gives one application for
>> which the subformula would be interesting. Moreover it is not
>> explained why the proposal from Laurent would not be appropriate for
>> this application.

The revision includes a new section "Motivations from Verified
Compilation" that goes into much more detail regarding the need for
what the revision now calls the "subformula conjunction property", and
explains how we initially used Laurent's proposal but then later
switched to the new system.

>> While the inference system defining <: is claimed to satisfy the
>> subformula property, it is not true strictly speaking. This is because
>> dom(A) and cod(A) are not subformulas of A. E.g. the derivation of
>> (A->B)/\(A->C)<:A->(B/\C) uses A/\A. Most of the consequences of the
>> subformula property hold anyway because dom(A) and cod(A) are not so
>> far from being subformulas of A (they are strictly smaller than A,
>> contain no new atom or new connective, etc).

True. The revision fixes this problem being more precise regarding the
terminology, defining the "subformula conjunction property".  Also the
revision explains why this property is enough for our purposes.

>> The abstract mentions that the transitivity proof requires a subtle
>> choice of the induction measure. Nevertheless it seems possible to
>> simplify it by using a lexicographic ordering on (depth B, size B,
>> size C) or on (depth B, size B + size C).

Excellent, we've switched to the lexicographic ordering on
(depth B, size B, size C).

>> page 1, line -5: type systems -> type system
>> page 2, line -15: then then -> then
>> page 3, line 4: provide a link to the Agda development
>> page 3, line 11: functions -> partial functions
>> page 3, lines 18,19: point to Proposition 3(5)
>> page 3, line -7: remarks about eta-reduction are not very clear
>>                  BCD typing is not preserved under eta-reduction
>> page 4, figure 3, A\subseteq B: missing \forall
>> page 5, line -3: <: -> \subseteq (3 times)
>> page 7, lines -2,-1: remove depth(B2)<=depth(B) and

The revision fixes all of the above.

Reviewer 2
==========

>> Despite these strong points, I am sorry to have to recommend rejection
>> of the paper in its current form.  There are two main reasons for
>> this:

>> 1. lack of details for the motivating connection to Siek's broader
>>    project(s), alluded to in the Introduction but too briefly to
>>    assess the value of this work on that basis.

The revision includes a new section "Motivations from Verified
Compilation" to address this problem.

>> 2. failure, as also in the Laurent paper that seems to have been the
>>    starting point for this work, to consider the deep line of work on
>>    logical treatments of subtyping.  I will mention papers in this
>>    line below.  The result is that Siek's formulation, while
>>    technically correct, is set up in a pretty delicate (and, in my
>>    view, not aesthetically or foundationally pleasing) way that I fear
>>    is unlikely to be robust to changes in the type system.

>> Main criticism 2:

>> The second criticism is about positioning the work with respect to an
>> important line of previous research concerned with the logical
>> underpinnings of subtyping. A very well written and informative
>> example is Tiuryn's "A Sequent Calculus for Subtyping Polymorphic
>> Types", which devises several subtyping systems as logics, and relates
>> to previous efforts in that direction. These works have a number of
>> interesting advantages over the proposals of Laurent and Siek.

>> First, they deal in a more compositional way with function types,
>> using ideas from substructural logic. The need for substructuralism is
>> due to the fact that usually one does not want to consider (a -> b ->
>> c) to be a subtype of (b -> a -> c), though as intuitionistic
>> formulas, these are equivalent.  In contrast, Siek's system handles
>> function types through some very particular meta-level operations on
>> type expressions and side conditions to the function subtyping rule
>> (->').  I worry that these conditions will be brittle, and not work
>> well in a setting where compositionality at the type level is
>> required.  Imagine, for example, allowing type-level functions as in
>> System F-omega.  Then particular checks on the forms of type
>> expressions will fail in the presence of applications of type
>> variables (i.e., abstracted type expressions), since we will not have
>> a concrete type expression to apply these various checks and meta-level
>> operations to.  So an approach that is based just on syntactic conditions
>> on type expressions seems fragile and not likely to be extensible to
>> richer type systems.

>> A point worth noting in passing is that Tiuryn's paper actually
>> presents a Curry-Howard view of subtyping and gives rules that derive
>> coercions. In fact, the line of work goes back to a paper of
>> Mitchell's which relates subtyping to typing by coercions which
>> eta-reduce to the identity. (One can find similar ideas in Miquel's
>> "Implicit Calculus of Constructions".)

>> @article{DBLP:journals/iandc/Tiuryn01, author = {Jerzy Tiuryn}, title = {A Sequent Calculus for Subtyping Polymorphic Types}, journal = {Inf. Comput.}, volume = {164}, number = {2}, pages = {345--369}, year = {2001}}

>> Other comments:

>> A. page 1: you say that BCD extends STLC with addition of intersection
>>    types, a top type, and "an infinite collection of type constants".
>>    The last of these surprised me as STLC usually includes base types
>>    (at least one, or else there are no types at all).  If you can
>>    clarify (very briefly -- do not want to distract from the main flow
>>    of your discourse) that would be great.

>> B. page 2, you write that the "BCD subtyping judgment does not enjoy
>>    the subformula property".  But proof-theoretically, the subformula
>>    property is a property of a proof system: a proof system has the
>>    subformula property if for every closed proof of that proof system,
>>    every formula occurring in the proof is a subformula of the theorem
>>    proved by the proof.  So it is a (meta-level) type error to refer
>>    to a judgment as having the subformula.  To be clear: it is not
>>    even correct to say that the set of valid subtyping judgments has
>>    the property, for different proof systems (some with, some without
>>    the property) could give rise to the same valid judgments.  We are
>>    speaking about whether some particular set of proof rules has the
>>    property.  Even there we could imagine a distinction (which I did
>>    not find in a brief literature search) between the *rules* having
>>    the property (we could say intrinsic subformula property?) and the
>>    proofs derivable from the rules having the property (extrinsic, the
>>    definition I use just above). For say that the proof system
>>    includes a spurious rule with some premise that is not a subformula
>>    of the conclusion, but where that rule can never actually be
>>    successfully used (due to other constraints in it) in a closed
>>    derivation.  Then the proof system would have the subformula
>>    property extrinsically but not intrinsically.

>> C. page 3: you say that A \in B "indicates whether A is syntactically
>>    a part of B".  This is not correct without some violence to the
>>    usual notion of syntactic part.  For example, alpha is not \in
>>    (alpha -> alpha) according to the rules of Figure 3.  This is again
>>    part of what I find displeasing about this approach: we have some
>>    very particular and not too intuitive (at least, not correctly
>>    intuitively described, in this case!) notions that yes, do lead to
>>    a correct system (this is proven in Agda).  But it would be much
>>    more satisfying, in the spirit of Tiuryn's work, for example, and
>>    the other works he cites and builds on, to see a compositional
>>    treatment of function types via substructural typing.  This would
>>    avoid these fiddly syntactic definitions.

Reviewer 3
==========

>> “Transitivity of Subtyping for Intersection Types”

>> The paper considers intersection type systems in the style of Coppo,
>> Dezani-Ciancaglini, and Salle, and more closely Barendregt, Dekkers,
>> and Statman (BCD). These systems typically include an explicit
>> transitivity rule in the Subtyping relation, which, among other
>> things, poses issues for inversion lemmas. The paper follows up on a
>> recent result by Laurent who demonstrated that equivalent systems can
>> be formulated without explicit transitivity. In comparison to Laurent,
>> the paper presents a much simplified proof strategy, reducing six
>> lemmas to one, and eliminating the need for a separate sequent-style
>> system.

>> With the adoption of intersection types into mainstream programming
>> languages like TypeScript and Scala, the paper is timely, and it
>> addresses an important question with a convincing solution, that may
>> be useful for future large-scale language formalization efforts.

>> The main technical innovation appears to be the design of the new
>> subtyping judgment "<:" in Figure 4 (later shown to be equivalent to
>> "≤" in BCD), and the induction metrics that enable the proof of
>> Theorem 7 (transitivity).

>> The paper is overall well written and clear, with a good level of
>> detail provided for the proofs. The formal development is mechanized
>> in Agda, which provides additional confidence in the results.

>> The only question I had is specifically related to the recent adoption
>> of intersection types in Scala and in formal work on the underlying
>> DOT calculus. The need for transitivity elimination is also described
>> in the OOPSLA'16 paper "Type Soundness for Dependent Object Types",
>> and the apparent solution adopted there was to "push back" top-level
>> uses of the transitivity rule further up into derivations, based on a
>> lemma "A <: B /\ B <! C ==> A <! C" where "A <! C" is a subtyping
>> derivation that does not end in the transitivity rule but may use it
>> internally. While it is not possible to get rid of transitivity
>> entirely in DOT, it seems possible to apply the "pushback" process to
>> exhaustion in the BCD setting considered in this paper, and in fact
>> I'm wondering if this is roughly equivalent to the induction metric
>> "<<". Perhaps the author can give this some thought, and maybe a
>> discussion of this piece of related work would be a useful addition
>> (if only to strengthen the motivation).

>> As far as I remember, DOT does not include the rule "(→∩)", however,
>> adapted in similar way as "(lbL)" and "(lbR)" to support internal
>> transitivity, it does not seem to pose any obvious major difficulties:

>>    A' ≤ A   (B ∩ C) ≤ D
>> -------------------------
>> (A → B) ∩ (A → C) ≤ A' → D

>> Minor notes:

>> The paper has a few formatting issues with text running into the
>> margin (p1, p8, some URLs in the bibliography).

>> Typos:

>> "and then then prove transitivity" (p2) double "then"

>> "Before moving on, we make note of some basic facts" perhaps "take
>> note" or "make a note of"?

>> Bibliography:

>> "M. Dezani-Ciancaglini, F. Honsell, and Y. Motohama. Compositional
>> characterisations of -terms" missing a lambda symbol

>> "E. Engeler. Algebras and combinators. algebra universalis" capitalize
>> journal title (?)

>> "H. Ishihara and T. Kurata. Completeness of intersection and union
>> type assignment systems for call-by-value -models." missing lambda

### 2020-05-14 papers/LMCS-subtyping-intersections/paper.tex
progress on response

@@
 \newcommand{\inside}[0]{\inplus}
-\newcommand{\contains}[0]{\subsetpluseq}
+\newcommand{\containedin}[0]{\subsetpluseq}
 
 \begin{abstract}
-  The subtyping rules for intersection type systems traditionally
-  employ a transitivity rule (Barendregt et al. 1983), which means
-  that subtyping does not satisfy the subformula property, making it
-  more difficult to use in filter models for compiler verification.
+  The subtyping rules for intersection types traditionally employ a
+  transitivity rule (Barendregt et al. 1983), which means that
+  subtyping does not satisfy the subformula property, making it more
+  difficult to use in filter models for compiler verification.
   Laurent develops a sequent-style subtyping system, without
@@
   subformulas.  Borrowing from Laurent's system, the rule for function
-  types is essentially the $\beta$-soundness property.  The main lemma
-  required for the transitivity proof is one that has been used to
-  prove the inversion principle for subtyping of function types. The
-  article concludes with a proof that the new subtyping system is
-  equivalent to that of Barendregt, Coppo, and Dezani-Ciancaglini.
+  types is essentially the $\beta$-soundness property. The article
+  proves that the new subtyping system is equivalent to that of
+  Barendregt, Coppo, and Dezani-Ciancaglini.
 \end{abstract}
@@
 types~\citep{Barendregt:2013aa}. The new subtyping system does not
-technically satisfy the subformula property but it satisfies a similar
-property: every type occuring in the derivation of $A <: B$ is a
-subformula of $A$ or $B$ or an intersection of such subformulas.
-This property is enough to ensure that if $A$ and $B$ are self
-consistent, then so are all the types that appear in the derivation of
-$A <: B$. As a result, the new subtyping system is highly suitable for
-mechanizing filter models.
+technically satisfy the subformula property but it satisfies what we
+call the \emph{subformula conjunction property}: every type occuring
+in the derivation of $A <: B$ is a subformula of $A$ or $B$ or an
+intersection of such subformulas.  This property is enough to ensure
+that if $A$ and $B$ are self consistent, then so are all the types
+that appear in the derivation of $A <: B$. As a result, the new
+subtyping system is highly suitable for mechanizing filter models.
 
@@
 and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
-relations $A \inside B$ and $A \contains B$.
+relations $A \inside B$ and $A \containedin B$.
 %
@@
 The $\topP{A}$ predicate identifies types that are equivalent to
-$\TOP$. The $\topInCod{A}$ predicate identifies types that have $\TOP$
-in their codomain.
+$\TOP$. (See Propostion~\ref{prop:top}(\ref{prop:AllBot-⊑-any}).) The
+$\topInCod{A}$ predicate identifies types that have $\TOP$ in their
+codomain.
 %
-The relations $A \inside B$ and $A \contains B$ enable the treatment
+The relations $A \inside B$ and $A \containedin B$ enable the treatment
 of a sequence of intersections $B_1 \cap \cdots \cap B_n$ (modulo
@@
 say that $A$ is a part of $B$ if $A \inside B$ and we say that $B$
-\emph{contains} $A$ if $A \contains B$.
+\emph{contains} $A$ if $A \containedin B$.
 
 \begin{proposition}\label{prop:union-subset-inv}
- \item If $A \cap B \contains C$, then $A \contains C$ and $B \contains C$. 
+ \item If $A \cap B \containedin C$, then $A \containedin C$ and $B \containedin C$. 
 \end{proposition}
@@
 
-  \fbox{$A \contains B$}
+  \fbox{$A \containedin B$}
   \[
-     A \contains B = \forall C.\, C \inside A \text{ implies } C \inside B
+     A \containedin B = \forall C.\, C \inside A \text{ implies } C \inside B
   \]
@@
     \begin{array}{l}
-      B \contains A\\
+      B \containedin A\\
       \neg\, \mathsf{top}(D) \\
@@
   \item If $A <: B$ and $C \inside B$, then $A <: C$.\label{prop:in-sub-sub}
-  \item If $A <: B$ and $C \contains B$, then $A <: C$.\label{prop:subset-sub-sub}
+  \item If $A <: B$ and $C \containedin B$, then $A <: C$.\label{prop:subset-sub-sub}
   \end{enumerate}
@@
   \item If $\topP{A}$ and $B \inside A$, then $\topP{B}$.\label{prop:AllBot-in}
-  \item If $\topP{A}$ and $B \contains A$, then $\topP{B}$.\label{prop:AllBot-subset}
+  \item If $\topP{A}$ and $B \containedin A$, then $\topP{B}$.\label{prop:AllBot-subset}
   \item If $\topP{A}$ and $A <: B$, then $\topP{B}$.\label{prop:AllBot-⊑}
@@
     this proposition. In the case for
-    $B = B_1 \cap B_2$, from $B_1 \cap B_2 <: A$, we have
-    $B_1 <: A$ and $B_2 <: A$ (Proposition~\ref{prop:union-subset-inv}).
+    $B = B_1 \cap B_2$, from $B_1 \cap B_2 \containedin A$, we have
+    $B_1 \containedin A$ and $B_2 \containedin A$
+    (Proposition~\ref{prop:union-subset-inv}).
     Then by the induction hypotheses for $B_1$ and $B_2$ we have
@@
     In that case we have $B = B_1 \to B_2$ and some $A'$ such that
-    $A' \contains A$, $B_1 <: \dom{A'}$, $\cod{A'} <: B_2$, $\neg\,\topP{B_2}$,
+    $A' \containedin A$, $B_1 <: \dom{A'}$, $\cod{A'} <: B_2$, $\neg\,\topP{B_2}$,
     and $\neg\,\topInCod{A'}$. From the premise $\topP{A}$
@@
   if there exists some type $B$ such that
-  $B \contains A$, $C <: \dom{B}$, $\cod{B} <: D$, and
+  $B \containedin A$, $C <: \dom{B}$, $\cod{B} <: D$, and
   $\neg\,\mathsf{topInCod}(B)$.
@@
     and so does $\dom{A_2} \to \cod{A_2}$.  So there exists
-    $B_1$ and $B_2$ such that $B_1 \contains B$, $\neg\,\topInCod{B_1}$,
+    $B_1$ and $B_2$ such that $B_1 \containedin B$, $\neg\,\topInCod{B_1}$,
     $\dom{A_1} <: \dom{B_1}$, $\cod{B_1} <: \cod{A_2}$ and similarly
@@
     $B$. We choose the witness $B_1 \cap B_2$.  Clearly we have $B_1 \cap
-    B_2 \contains B$ and $\neg \topInCod{B_1 \cap B_2}$.  Also, we have
+    B_2 \containedin B$ and $\neg \topInCod{B_1 \cap B_2}$.  Also, we have
     \[
@@
 We now turn to the proof of transitivity, that if $A <: B$ and $B <:
-C$, then $A <: C$. The proof is by well-founded induction on the
-lexicographical ordering of the sum of the depths of $A$ and $C$
-followed by the sum of the sizes of $B$ and $C$. To be precise, we
-define this ordering as follows.
+C$, then $A <: C$.
+%
+The proof is by well-founded induction on the lexicographical ordering
+of the depth of $B$, the size of $B$, and then the size of $C$.  To be
+precise, we define this ordering as follows.
 \[
@@
 \begin{array}{l}
-\depth{A} + \depth{C} < \depth{A'} + \depth{C'}
-\text{ or } \\
-(\depth{A} + \depth{C}
-\leq
-\depth{A'} + \depth{C'} \\
-\text{ and }
-\size{B} + \size{C}
-<
-\size{B'} + \size{C'})
+\depth{B} < \depth{B'} \\
+\text{ or } \depth{B} \leq \depth{B'} \text{ and } \size{B} < \size{B'} \\
+\text{ or }
+  \depth{B} \leq \depth{B'} \text{ and } \size{B} \leq \size{B'}\\
+  \;\quad \text{ and } \size{C} < \size{C'}
 \end{array}
 \]
+%
+%% The proof is by well-founded induction on the lexicographical ordering
+%% of the sum of the depths of $A$ and $C$ followed by the sum of the
+%% sizes of $B$ and $C$. To be precise, we define this ordering as
+%% follows.
+%% \[
+%% \langle A, B, C \rangle \ll \langle A', B', C' \rangle
+%% =
+%% \begin{array}{l}
+%% \depth{A} + \depth{C} < \depth{A'} + \depth{C'}
+%% \text{ or } \\
+%% (\depth{A} + \depth{C}
+%% \leq
+%% \depth{A'} + \depth{C'} \\
+%% \text{ and }
+%% \size{B} + \size{C}
+%% <
+%% \size{B'} + \size{C'})
+%% \end{array}
+%% \]
 where $\size{A}$ is 
@@
     hypothesis, noting that $\langle A, B_1, C \rangle \ll \langle A,
-    B, C \rangle$ because $\size{B_1} < \size{B}$.
-  \item[Case (lb$_R$)] So $B = B_1 \cap B_2$, $B_2 <: C$, and $A <: B_1 \cap
-    B_2$.  We have $A <: B_2$ (Proposition~\ref{prop:subtyping} part
-    \ref{prop:⊔⊑-inv}), so we conclude that $A <: C$ by the induction
-    hypothesis, noting that $\langle A, B_2, C \rangle \ll \langle A,
-    B, C \rangle$ because $\depth{B_2} \leq \depth{B}$ and $\size{B_2}
-    < \size{B}$.
+    B, C \rangle$
+    %because $\size{B_1} < \size{B}$.
+    because $\depth{B_1} \le \depth{B}$
+    and $\size{B_1} < \size{B}$.
+  \item[Case (lb$_R$)] So $B = B_1 \cap B_2$, $B_2 <: C$, and $A <:
+    B_1 \cap B_2$.  We have $A <: B_2$
+    (Proposition~\ref{prop:subtyping} part \ref{prop:⊔⊑-inv}), so we
+    conclude that $A <: C$ by the induction hypothesis, noting that
+    $\langle A, B_2, C \rangle \ll \langle A, B, C \rangle$ because
+    $\depth{B_2} \leq \depth{B}$ and $\size{B_2} < \size{B}$.
+
   \item[Case (glb)] We have $C = C_1 \cap C_2$, $B <: C_1$, and $B <:
@@
     C_2$, noting that $\langle A, B, C_1 \rangle \ll \langle A, B, C
-    \rangle$ $\langle A, B, C_2 \rangle \ll \langle A, B, C\rangle$
-    because $\depth{C_1} \leq \depth{C}$, $\depth{C_2} \leq
-    \depth{C}$, $\size{C_1} < C$, and $\size{C_2} < C$.
+    \rangle$ and $\langle A, B, C_2 \rangle \ll \langle A, B, C\rangle$
+    %% because $\depth{C_1} \leq \depth{C}$, $\depth{C_2} \leq
+    %% \depth{C}$, $\size{C_1} < C$, and $\size{C_2} < C$.
+    because $\size{C_1} < C$ and $\size{C_2} < C$.
+    %
     We conclude $A <: C_1 \cap C_2$ by rule (glb).
@@
     there exists $B'$ such that $C_1 <: \dom{B'}$, $\cod{B'} <: C_2$,
-    $B' \contains B$, and $\neg \topInCod{B'}$. From $A <: B$ and $B' \contains B$, we
+    $B' \containedin B$, and $\neg \topInCod{B'}$. From $A <: B$ and $B' \containedin B$, we
     have $A <: B'$ (Proposition~\ref{prop:subtyping} part
@@
     $\dom{B'} \to \cod{B'}$ factors $A$. That means there exists $A'$
-    such that $A' \contains A$, $\neg\,\topInCod{A'}$, $\dom{B'} <: \dom{A'}$,
+    such that $A' \containedin A$, $\neg\,\topInCod{A'}$, $\dom{B'} <: \dom{A'}$,
     and $\cod{A'} <: \cod{B'}$. Then by the induction hypothesis, we
@@
     noting that
-    $\langle C_1, \dom{B'}, \dom{A'} \rangle \ll \langle A, B, C \rangle$
-    because $\depth{C_1} + \depth{\dom{A'}} < \depth{A} + \depth{C}$
+    \[
+    \langle C_1, \dom{B'}, \dom{A'} \rangle \ll \langle A, B, C \rangle
+    \]
     and
-    $\langle \cod{A'}, \cod{B'}, C_2 \rangle \ll \langle A, B, C \rangle$
-    because $\depth{\cod{A'}} + \depth{C_2} < \depth{A} + \depth{C}$.
+    \[
+    \langle \cod{A'}, \cod{B'}, C_2 \rangle \ll \langle A, B, C \rangle
+    \]
+%    because $\depth{C_1} + \depth{\dom{A'}} < \depth{A} + \depth{C}$
+    because $\depth{\dom{B'}} < \depth{B}$
+%    because $\depth{\cod{A'}} + \depth{C_2} < \depth{A} + \depth{C}$.
+    and $\depth{\cod{B'}} < \depth{B}$.
     We conclude that $A <: C_1 \to C_2$ by rule $(\to')$ witnessed by $A'$.
@@
       $C \leq \dom{B}$ and also $\cod{B} \leq D$.
-      From $B \contains A$ we have $A \leq B$.      
+      From $B \containedin A$ we have $A \leq B$.      
       Then by Lemma~\ref{lem:dv↦cv<:v} we have

### 2020-05-14 papers/LMCS-subtyping-intersections/response.txt
check

@@
 
-The revision fixes all of the above.
+The revision fixes all of the above except for the last item on page 3
+regarding eta-reduction. The reviewer is mistaken. BCD typing is
+indeed preserved under eta reduction (see Proposition 12.1.15 and
+Corollary 14.2.9 of Lambda Calculus with Types by Barendregt, Dekkers,
+and Statman). Perhaps the reviewer was thinking of eta-expansion,
+which is not preserved by BCD typing.
 
@@
 
+I've rephrased to remove mention of the STLC, as that was not necessary.
+
 >> B. page 2, you write that the "BCD subtyping judgment does not enjoy
@@
 
+The mistaken uses of "judgment" have been replaced with "subtyping system"
+(i.e. proof system).
+
 >> C. page 3: you say that A \in B "indicates whether A is syntactically
@@
 >>    avoid these fiddly syntactic definitions.
+
+I've changed the notation for \in and \subseteq to make it clear
+that these notions are different from the usual ones.
 

### 2020-05-15 papers/LMCS-subtyping-intersections/response.txt
additions to response

@@
 
-True. The revision fixes this problem being more precise regarding the
+The revision fixes this problem by being more precise regarding
 terminology, defining the "subformula conjunction property".  Also the
@@
 
-The revision fixes all of the above except for the last item on page 3
-regarding eta-reduction. The reviewer is mistaken. BCD typing is
-indeed preserved under eta reduction (see Proposition 12.1.15 and
-Corollary 14.2.9 of Lambda Calculus with Types by Barendregt, Dekkers,
-and Statman). Perhaps the reviewer was thinking of eta-expansion,
-which is not preserved by BCD typing.
+The revision addresses all of the above except for the last item on
+page 3 regarding eta-reduction. BCD typing is indeed preserved under
+eta reduction (see Proposition 12.1.15 and Corollary 14.2.9 of Lambda
+Calculus with Types by Barendregt, Dekkers, and Statman). Perhaps the
+reviewer was thinking of eta-expansion, which is not preserved by BCD
+typing.
 
@@
 ==========
-
->> Despite these strong points, I am sorry to have to recommend rejection
->> of the paper in its current form.  There are two main reasons for
->> this:
 
@@
 >>    starting point for this work, to consider the deep line of work on
->>    logical treatments of subtyping.  I will mention papers in this
->>    line below.  The result is that Siek's formulation, while
->>    technically correct, is set up in a pretty delicate (and, in my
->>    view, not aesthetically or foundationally pleasing) way that I fear
->>    is unlikely to be robust to changes in the type system.
+>>    logical treatments of subtyping.
+>>
+>> ...  In contrast, Siek's system handles function types through some
+>> very particular meta-level operations on type expressions and side
+>> conditions to the function subtyping rule (->').  I worry that
+>> these conditions will be brittle, and not work well in a setting
+>> where compositionality at the type level is required.  Imagine, for
+>> example, allowing type-level functions as in System F-omega.  Then
+>> particular checks on the forms of type expressions will fail in the
+>> presence of applications of type variables (i.e., abstracted type
+>> expressions), since we will not have a concrete type expression to
+>> apply these various checks and meta-level operations to.  So an
+>> approach that is based just on syntactic conditions on type
+>> expressions seems fragile and not likely to be extensible to richer
+>> type systems.
 
->> Main criticism 2:
-
->> The second criticism is about positioning the work with respect to an
->> important line of previous research concerned with the logical
->> underpinnings of subtyping. A very well written and informative
->> example is Tiuryn's "A Sequent Calculus for Subtyping Polymorphic
->> Types", which devises several subtyping systems as logics, and relates
->> to previous efforts in that direction. These works have a number of
->> interesting advantages over the proposals of Laurent and Siek.
-
->> First, they deal in a more compositional way with function types,
->> using ideas from substructural logic. The need for substructuralism is
->> due to the fact that usually one does not want to consider (a -> b ->
->> c) to be a subtype of (b -> a -> c), though as intuitionistic
->> formulas, these are equivalent.  In contrast, Siek's system handles
->> function types through some very particular meta-level operations on
->> type expressions and side conditions to the function subtyping rule
->> (->').  I worry that these conditions will be brittle, and not work
->> well in a setting where compositionality at the type level is
->> required.  Imagine, for example, allowing type-level functions as in
->> System F-omega.  Then particular checks on the forms of type
->> expressions will fail in the presence of applications of type
->> variables (i.e., abstracted type expressions), since we will not have
->> a concrete type expression to apply these various checks and meta-level
->> operations to.  So an approach that is based just on syntactic conditions
->> on type expressions seems fragile and not likely to be extensible to
->> richer type systems.
-
->> A point worth noting in passing is that Tiuryn's paper actually
->> presents a Curry-Howard view of subtyping and gives rules that derive
->> coercions. In fact, the line of work goes back to a paper of
->> Mitchell's which relates subtyping to typing by coercions which
->> eta-reduce to the identity. (One can find similar ideas in Miquel's
->> "Implicit Calculus of Constructions".)
-
->> @article{DBLP:journals/iandc/Tiuryn01, author = {Jerzy Tiuryn}, title = {A Sequent Calculus for Subtyping Polymorphic Types}, journal = {Inf.
-Comput.}, volume = {164}, number = {2}, pages = {345--369}, year = {2001}}
+The logical treatments of subtyping are indeed useful for extending to
+richer type systems for programming languages. The author has made use
+of them in the context of research on gradual typing.  However, the
+motivation for the present work is rather different: constructing
+filter models (denotational semantics) for programming languages. For
+that purpose, there is no need for richer type systems such as System
+F-omega; intersection types already provide what is needed to express
+the runtime behavior of functional programs. The revision includes
+changes to the introduction to differentiate the varied uses of
+intersection type systems and to make it clear that the present
+article's motivation comes from denotational semantics, not type
+systems for programming languages.
 
@@
 >>    every formula occurring in the proof is a subformula of the theorem
->>    proved by the proof.  So it is a (meta-level) type error to refer
->>    to a judgment as having the subformula.  To be clear: it is not
->>    even correct to say that the set of valid subtyping judgments has
->>    the property, for different proof systems (some with, some without
->>    the property) could give rise to the same valid judgments.  We are
->>    speaking about whether some particular set of proof rules has the
->>    property.  Even there we could imagine a distinction (which I did
->>    not find in a brief literature search) between the *rules* having
->>    the property (we could say intrinsic subformula property?) and the
->>    proofs derivable from the rules having the property (extrinsic, the
->>    definition I use just above). For say that the proof system
->>    includes a spurious rule with some premise that is not a subformula
->>    of the conclusion, but where that rule can never actually be
->>    successfully used (due to other constraints in it) in a closed
->>    derivation.  Then the proof system would have the subformula
->>    property extrinsically but not intrinsically.
+>>    proved by the proof.
 
@@
 
-I've changed the notation for \in and \subseteq to make it clear
-that these notions are different from the usual ones.
+I've changed the notation for \in and \subseteq to make it clear that
+these notions are different from the usual ones and added some
+explanation to provide an intuition for what they accomplish.
 

### 2020-05-15 papers/LMCS-subtyping-intersections/paper.tex
additions to response

@@
   provides a direct proof of transitivity, significantly reducing the
-  length of the proof. The new system satisfies something like the
-  subformula property: every type occuring in the derivation of $A <:
-  B$ is a subformula of $A$ or $B$, or an intersection of such
-  subformulas.  Borrowing from Laurent's system, the rule for function
-  types is essentially the $\beta$-soundness property. The article
-  proves that the new subtyping system is equivalent to that of
-  Barendregt, Coppo, and Dezani-Ciancaglini.
+  length of the proof. Inspired by Laurent's system, the rule for
+  function types is essentially the $\beta$-soundness property.  The
+  new system satisfies the ``subformula conjunction property'': every
+  type occuring in the derivation of $A <: B$ is a subformula of $A$
+  or $B$, or an intersection of such subformulas. The article proves
+  that the new subtyping system is equivalent to that of Barendregt,
+  Coppo, and Dezani-Ciancaglini.
 \end{abstract}
@@
 \end{itemize}
+The motivation for this article comes from the third use of
+intersection types, in denotational semantics.  We are interested in
+constructing filter models for functional programming languages in
+support of mechanized compiler
+verification. Section~\ref{sec:motivation} describes how this line of
+research motivated the results in this article.
 
@@
 \section{Motivations from Verified Compilation}
+\label{sec:motivation}
 
@@
 %
-The relations $A \inside B$ and $A \containedin B$ enable the treatment
-of a sequence of intersections $B_1 \cap \cdots \cap B_n$ (modulo
-associativity and commutativity) as if it was set of types $\{ B_1,
-\ldots, B_n \}$, where each $B_i$ is an atom or a function type.  We
-say that $A$ is a part of $B$ if $A \inside B$ and we say that $B$
-\emph{contains} $A$ if $A \containedin B$.
+The relations $A \inside B$ and $A \containedin B$ enable the
+treatment of a sequence of intersections $B_1 \cap \cdots \cap B_n$
+(modulo associativity and commutativity) as if it were a set of types
+$\{ B_1, \ldots, B_n \}$, where each $B_i$ is an atom or a function
+type.  We say that $A$ is a part of $B$ if $A \inside B$ and we say
+that $B$ \emph{contains} $A$ if $A \containedin B$.
 

### 2020-05-15 papers/LMCS-subtyping-intersections/response.txt
tweak

@@
 
->> “Transitivity of Subtyping for Intersection Types”
-
->> The paper considers intersection type systems in the style of Coppo,
->> Dezani-Ciancaglini, and Salle, and more closely Barendregt, Dekkers,
->> and Statman (BCD). These systems typically include an explicit
->> transitivity rule in the Subtyping relation, which, among other
->> things, poses issues for inversion lemmas. The paper follows up on a
->> recent result by Laurent who demonstrated that equivalent systems can
->> be formulated without explicit transitivity. In comparison to Laurent,
->> the paper presents a much simplified proof strategy, reducing six
->> lemmas to one, and eliminating the need for a separate sequent-style
->> system.
-
->> With the adoption of intersection types into mainstream programming
->> languages like TypeScript and Scala, the paper is timely, and it
->> addresses an important question with a convincing solution, that may
->> be useful for future large-scale language formalization efforts.
-
->> The main technical innovation appears to be the design of the new
->> subtyping judgment "<:" in Figure 4 (later shown to be equivalent to
->> "≤" in BCD), and the induction metrics that enable the proof of
->> Theorem 7 (transitivity).
-
->> The paper is overall well written and clear, with a good level of
->> detail provided for the proofs. The formal development is mechanized
->> in Agda, which provides additional confidence in the results.
-
 >> The only question I had is specifically related to the recent adoption
@@
 >> transitivity, it does not seem to pose any obvious major difficulties:
-
 >>    A' ≤ A   (B ∩ C) ≤ D
@@
 >> (A → B) ∩ (A → C) ≤ A' → D
+
+I suspect that the above rule handles the case when (→∩) derives the
+A <: B premise of (trans), but not when (→∩) derives the B <: C premise
+of (trans).  In the 2012 article "Intersection types with subtyping by
+means of cut elimination", Laurent comes up with another rule that
+fixes that problem:
+
+D ≤ C → A    D ≤ C → B
+----------------------
+D ≤ C → (A ∩ B)
+
+but then he was forced to add another rule
+
+C ≤ A → D   D ≤ B
+-----------------
+C ≤ A → B
+
+that does not satisfy the subformula property.
 

### 2020-05-15 papers/LMCS-subtyping-intersections/paper.tex
tweak

@@
   provides a direct proof of transitivity, significantly reducing the
-  length of the proof. Inspired by Laurent's system, the rule for
-  function types is essentially the $\beta$-soundness property.  The
-  new system satisfies the ``subformula conjunction property'': every
-  type occuring in the derivation of $A <: B$ is a subformula of $A$
-  or $B$, or an intersection of such subformulas. The article proves
-  that the new subtyping system is equivalent to that of Barendregt,
-  Coppo, and Dezani-Ciancaglini.
+  length of the proof, exchanging the six lemmas for just
+  one. Inspired by Laurent's system, the rule for function types is
+  essentially the $\beta$-soundness property.  The new system
+  satisfies the ``subformula conjunction property'': every type
+  occuring in the derivation of $A <: B$ is a subformula of $A$ or
+  $B$, or an intersection of such subformulas. The article proves that
+  the new subtyping system is equivalent to that of Barendregt, Coppo,
+  and Dezani-Ciancaglini.
 \end{abstract}
@@
 Salle to study normalization in the lambda
-calculus~\citep{Coppo:1979aa}. Subsequently intersection type systems
-have been used for at least three purposes:
+calculus~\citep{Coppo:1979aa}. Subsequently intersection types have
+been used for at least three purposes:
 \begin{itemize}
-  \item as type systems for programming
-    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
-  \item to formulate precise static
+  \item in type systems for programming
+    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Amin:2017aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
+  \item in precise static
     analyses~\citep{Turbak:1997aa,Palsberg:1998aa,Mossin:2003aa,Simoes:2007aa},
     and
-  \item to express filter models for the denotational semantics of a
-    variety of lambda
+  \item in filter models for the denotational semantics of a variety
+    of lambda
     calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
@@
 support of mechanized compiler
-verification. Section~\ref{sec:motivation} describes how this line of
-research motivated the results in this article.
+verification. Section~\ref{sec:motivation} describes how this research
+motivates the results in this article.
 
@@
 and then prove transitivity.  Unfortunately, the $({\to}{\cap})$ rule
-of the BCD system significantly complicates the situation.
+of the BCD system significantly complicates the situation.  One might
+hope to omit the $({\to}{\cap})$ rule, but it plays an import role in
+filter models; without it they are not adequate with respect to
+reduction semantics.
 
@@
 Our interest in intersection type systems stems from our work on
-mechanized proofs of correctness for compiling functional programming
-languages. We are exploring whether the use of denotational semantics
-can streamline such proofs compared to the operational techniques,
-e.g., used to prove the correctness of
+mechanized correctness proofs for compilers for functional programming
+languages. We are exploring whether denotational semantics can
+streamline such proofs compared to the operational techniques such as
+those used to prove the correctness of
 CakeML~\citep{Kumar:2014aa,Owens:2017aa}. In particular, we are
@@
 \]
-If one takes each type constant $c_i$ to be a singleton type for the
-number $i$, then the set of types assigned to the identity function
-starts to look like the graph of the function. Indeed, filter models
-are closely related to graph
-models~\citep{Scott:1976lq,Engeler:1981aa,Plotkin:1993ab}.
+If one takes each type constant $c_i$ to represent a singleton type
+for the number $i$, then the set of types assigned to the identity
+function starts to look like the graph of the function. Indeed, filter
+models are closely related to graph
+models~\citep{Scott:1976lq,Engeler:1981aa,Plotkin:1993ab,Barendregt:2013aa}.
 
 One of the standard compiler transformations that we are interested in
-verifying is common subexpression elimination. For example, this
-transformation would replace the duplicated $(f \; x)$ terms in the
-following
+verifying is common subexpression
+elimination~\citep{Downey:JACM:1980,Appel:1992fk,Tarditi:1996aa}. For
+example, this transformation would replace the duplicated $(f \; x)$
+terms in the following
 \[
@@
 can be accomplished by requiring every entry in a function's graph to
-be ``consistent''. Two function types, $A \to B$ and $A' \to B'$, are
-consistent, written $(A \to B) \sim (A' \to B')$, when either $A
-\not\sim A'$ or $A \sim A'$ and $B \sim B'$. Two type constants are
-consistent when they are equal. One would then hope to prove that for
-any term $M$, if $\Gamma \vdash M : C$ and $\Gamma \vdash M : C'$,
-then $C \sim C'$ as long as $C$, $C'$, and $\Gamma$ are self
-consistent. A key lemma needed for that proof is that consistency is
-upward closed with respect to subtyping: if $A \sim B$, $A <: C$, $B
-<: D$, then $C \sim D$.
+be ``consistent''. Two entries are consistent, written $(A \to B) \sim
+(A' \to B')$, when either $A \not\sim A'$ or $A \sim A'$ and $B \sim
+B'$. Two type constants are consistent when they are equal. One would
+then hope to prove that for any term $M$, if $\Gamma \vdash M : C$ and
+$\Gamma \vdash M : C'$, then $C \sim C'$ as long as $C$, $C'$, and
+$\Gamma$ are self consistent. A key lemma needed for that proof is
+that consistency is upward closed with respect to subtyping: if $A
+\sim B$, $A <: C$, $B <: D$, then $C \sim D$.
 
 This brings us back to the transitivity rule and the subformula
-property. The proof of this lemma hits a snag in the cases where $A <:
-C$ and $B <: D$ are derived using the transitivity rule. The
+property. The proof of the above lemma hits a snag in the cases where
+$A <: C$ and $B <: D$ are derived using the transitivity rule. The
 intermediate type is not guaranteed to be self consistent. We explored
 adding this guarantee to the transitivity rule, but it caused a
-significant increase in obligations elsewhere in the proof. By some
+significant increase in obligations elsewhere in our proofs. By some
 serendipity, around that moment in time \citet{Laurent:2018aa}
@@
 A year later we began to port this filter model to the Agda proof
-assistant, motivated by the availability of dependent types to make
-our development more reusable: we needed several variations on the
-filter model to give semantics to different languages, including ones
-that are call-by-value and call-by-name, and to the different
-intermediate languages used in our compiler. We found it tedious to
-repeat Laurent's proof of transitivity for each of these variations,
-and in the process developed an intuition that the sequent-style
-system was not necessary. The present article discusses how we were
-able to extract the key feature from Laurent's system, the rule for
-subtyping between function types and adapt the BCD subtyping system,
-replacing the function $(\to)$, distributivity $({\to}{\cap})$, and
-transitivity (trans) rules.  We then prove transitivity directly,
-without the sequence of six lemmas, instead using one lemma that has
-been used to prove the inversion principle for subtyping of function
+assistant. We needed several variations on the filter model to give
+semantics to different languages, including ones that are
+call-by-value and call-by-name, and to the different intermediate
+languages used in our compiler.  The availability of dependent types
+in Agda promised to make our development more reusable.  However, we
+still found it tedious to repeat Laurent's proof of transitivity for
+each of the filter models, and in the process we developed an
+intuition that the sequent-style system and the six lemmas were not
+necessary.
+
+The present article discusses how we were able to import the key
+feature from Laurent's system, the rule for subtyping between function
+types, into the BCD subtyping system, replacing the function $(\to)$,
+distributivity $({\to}{\cap})$, and transitivity (trans) rules.  We
+then prove transitivity without the sequence of six lemmas, instead
+using one lemma that has been used previously to prove the inversion
+principle for subtyping of function
 types~\citep{Barendregt:2013aa}. The new subtyping system does not
@@
 
-We discuss two recent developments regarding subtyping for
-intersection types in Section~\ref{sec:other-developments}.  We
-present our new subtyping system in Section~\ref{sec:new-subtyping},
-prove transitivity in Section~\ref{sec:trans}, and prove its
-equivalence to BCD subtyping in Section~\ref{sec:equiv}. We make some
-concluding remarks in Section~\ref{sec:conclude}.
+\section{Road Map}
+\label{sec:road-map}
+
+The rest of this article is organized as follows. We discuss two
+recent developments regarding subtyping for intersection types in
+Section~\ref{sec:recent-developments}.  We present our new subtyping
+system in Section~\ref{sec:new-subtyping}, prove transitivity in
+Section~\ref{sec:trans}, and prove its equivalence to BCD subtyping in
+Section~\ref{sec:equiv}. We conclude the artcle in
+Section~\ref{sec:conclude}.
 
@@
 
-\section{Other Developments}
-\label{sec:other-developments}
+\section{Recent Developments}
+\label{sec:recent-developments}
 
 There has been a recent flurry of interest in subtyping for
-intersection types that has led to two more systems that satisfy the
-subformula property, in addition to the one by \citet{Laurent:2018aa}
-discussed above. These two systems were developed for the purposes of
-type checking programming languages, and not for the construction of
-filter models, so the goals are somewhat different.
+intersection types that has led to two algorithms that effectively
+satisfy the subformula property, in addition to the one by
+\citet{Laurent:2018aa} discussed above. These two algorithms were
+developed for the purposes of type checking programming languages, and
+not for the construction of filter models, so the goals are somewhat
+different. In short, the algorithms introduce complications for the
+sake of efficiency, which would be undesirable for use in a filter
+model.
 
@@
 ${\to}{\cap}$ (left to right) as much as possible.
-%[todo: say something about the proof of transitivity]
-
-%$λ^{\mathrm{BCD}}_\cap$
 

### 2020-05-15 papers/LMCS-subtyping-intersections/response.txt
check

@@
 intersection type systems and to make it clear that the present
-article's motivation comes from denotational semantics, not type
+article's motivation comes from the realm of filter models, not type
 systems for programming languages.
@@
 
+That is an interesting connection. I think that this style of proof
+would work using the new function rule (→′) but not with the old rules
+(→) and (→∩). I've added a discussion of this connection to the
+revision.
+
 >> As far as I remember, DOT does not include the rule "(→∩)", however,
@@
 >> type assignment systems for call-by-value -models." missing lambda
+
+All of the above have been fixed in the revision.

### 2020-05-15 papers/LMCS-subtyping-intersections/paper.tex
check

@@
 been used for at least three purposes:
-\begin{itemize}
+\begin{enumerate}
   \item in type systems for programming
@@
     calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
-\end{itemize}
+\end{enumerate}
 The motivation for this article comes from the third use of
@@
 support of mechanized compiler
-verification. Section~\ref{sec:motivation} describes how this research
+verification. Section~\ref{sec:motivation} describes how that research
 motivates the results in this article.
@@
 that if $A$ and $B$ are self consistent, then so are all the types
-that appear in the derivation of $A <: B$. As a result, the new
-subtyping system is highly suitable for mechanizing filter models.
+that appear in the derivation of $A <: B$.  Thus, the new subtyping
+system is suitable for mechanizing filter models.
 
@@
 There has been a recent flurry of interest in subtyping for
-intersection types that has led to two algorithms that effectively
-satisfy the subformula property, in addition to the one by
-\citet{Laurent:2018aa} discussed above. These two algorithms were
-developed for the purposes of type checking programming languages, and
-not for the construction of filter models, so the goals are somewhat
-different. In short, the algorithms introduce complications for the
-sake of efficiency, which would be undesirable for use in a filter
-model.
+intersection types. We begin by discussing two algorithms that
+effectively satisfy the subformula property. We then discuss issues
+surrounding subtyping and transitivity in the Scala language.
 
@@
 
-\section{A New Subtyping Judgment}
+These two algorithms were developed for the purposes of type checking
+programming languages, and not for the construction of filter models,
+so the goals are somewhat different. In short, the algorithms
+introduce complications for the sake of efficiency, which would be
+undesirable for use in a filter model.
+
+Research on the Scala programming language~\citep{Odersky:2004aa} has
+led to the development of the Dependent Object Types (DOT) calculus,
+which includes path-dependent types and intersection
+types~\citep{Amin:2016aa,Amin:2017aa}.  The interplay between
+path-dependent types, subtyping, and transitivity in DOT has proved a
+challenge in the proof of type soundness. Nevertheless,
+\citet{Rompf:2016aa} prove the inversion principle for function types
+using a ``pushback'' lemma that reorganizes any subtyping derivation
+such that the last rule to be applied is never transitivity. Applying
+such a lemma exhaustively might provide an alternative route to
+transitivity for the new subtyping system presented in this
+article. Conversely, the new subtyping system might enable the
+addition of distributivity of function and intersection types to
+DOT. However, we suspect that the pushback approach would encounter
+difficulties if applied to the original BCD subtyping rules.
+
+\section{A New Subtyping System}
 \label{sec:new-subtyping}
 
-Our new subtyping judgment relies on a few definitions that are given
-in Figure~\ref{fig:aux}. These include the partial functions $\dom{A}$
+Our new subtyping system relies on a few definitions that are given in
+Figure~\ref{fig:aux}. These include the partial functions $\dom{A}$
 and $\cod{A}$, the $\topP{A}$ and $\topInCod{A}$ predicates, and the
@@
   \dom{A \to B} &= A \\
-  \dom{A \cap B} &= \dom{A} \cap \dom {B} \\
+  \dom{A \cap B} &= \dom{A} \,\cap\, \dom {B} \\
   \\
   \cod{A \to B} &= B \\
-  \cod{A \cap B} &= \cod{A} \cap \cod {B}
+  \cod{A \cap B} &= \cod{A} \,\cap\, \cod {B}
   \end{align*}
@@
   \end{gather*}
-  \caption{The New Subtyping Judgment}
+  \caption{The New Subtyping System}
   \label{fig:new-subtyping}
@@
 
-The new intersection subtyping judgment, $A <: B$, is defined in
-Figure~\ref{fig:new-subtyping}. First, it does not include the (trans)
-rule.  It also replaces the (refl) rule with reflexivity for atoms
-(refl$_\alpha$). The most important rule is the one for function types
-$(\to')$, which subsumes $(\to)$ and $({\to}{\cap})$ in BCD subtyping.
-The $(\to')$ rule essentially turns the $\beta$-soundness property
-into a subtyping rule~\citep{Barendregt:2013aa}. The $(\to')$ rule
-says that a type $A$ is a subtype of a function type $C \to D$ if a
-type contained in $A$, call it $B$, has domain and codomain that are
-larger and smaller than $C$ and $D$, respectively. The use of
-containment enables this rule to absorb uses of (incl$_L$) and
-(incl$_R$) on the left.  The side conditions $\neg\;\topP{B}$ and
-$\neg\;\topInCod{D}$ are needed because of the $(\TOP{\to}')$ rule,
-which in turn is needed to preserve types under $\eta$-reduction.  In
-the many systems that do not involve $\eta$-reduction, the
-$(\TOP{\to}')$ rule can be omitted, as well as these side
-conditions. The rules (lb$_L$) and (lb$_R$) adapt (incl$_L$) and
-(incl$_R$) to a system without transitivity, and have appeared many
-times in the literature~\citep{Bakel:1995aa}.  The $(\TOP{\to}')$ rule
-generalizes the $(\TOP{\to})$ rule, replacing the $\TOP$ on the left
-with any type $A$, because for transitivity, any type is below
-$\TOP$. The $(\TOP{\to}')$ rule replaces the $\TOP$ in the
-codomain on the right with any type $D$ that is equivalent to $\TOP$.
+The new intersection subtyping system, with judgements of the form $A
+<: B$, is defined in Figure~\ref{fig:new-subtyping}. First, it does
+not include the (trans) rule.  It also replaces the (refl) rule with
+reflexivity for atoms (refl$_\alpha$). The most important rule is the
+one for function types $(\to')$, which subsumes $(\to)$ and
+$({\to}{\cap})$ in BCD subtyping.  The $(\to')$ rule essentially turns
+the $\beta$-soundness property into a subtyping
+rule~\citep{Barendregt:2013aa}. The $(\to')$ rule says that a type $A$
+is a subtype of a function type $C \to D$ if a type contained in $A$,
+call it $B$, has domain and codomain that are larger and smaller than
+$C$ and $D$, respectively. The use of containment enables this rule to
+absorb uses of (incl$_L$) and (incl$_R$) on the left.  The side
+conditions $\neg\;\topP{B}$ and $\neg\;\topInCod{D}$ are needed
+because of the $(\TOP{\to}')$ rule, which in turn is needed to
+preserve types under $\eta$-reduction.  In the many systems that do
+not involve $\eta$-reduction, the $(\TOP{\to}')$ rule can be omitted,
+as well as these side conditions. The rules (lb$_L$) and (lb$_R$)
+adapt (incl$_L$) and (incl$_R$) to a system without transitivity, and
+have appeared many times in the literature~\citep{Bakel:1995aa}.  The
+$(\TOP{\to}')$ rule generalizes the $(\TOP{\to})$ rule, replacing the
+$\TOP$ on the left with any type $A$, because for transitivity, any
+type is below $\TOP$. The $(\TOP{\to}')$ rule replaces the $\TOP$ in
+the codomain on the right with any type $D$ that is equivalent to
+$\TOP$.
 
@@
 
-The proof of transitivity relies on the following lemma, which is
+Our proof of transitivity relies on the following lemma, which is
 traditionally needed to prove the inversion principle for function
-types. However, it was not needed for our system because the rule
-$(\to')$ is already quite close to the inversion principle.
-The lemma states that if every function type $C \to D$ in A
-factors $B$, then $\dom{A} \to \cod{A}$ also factors $B$.
+types. (In our system this lemma is not needed to prove the inversion
+principle because the rule $(\to')$ is already quite close to the
+inversion principle.)  The lemma states that if every function type $C
+\to D$ in A factors $B$, then $\dom{A} \to \cod{A}$ also factors $B$.
 
@@
     B$ we immediately conclude that $A <: \alpha$.
-  \item[Case (lb$_L$)] So $B = B_1 \cap B_2$, $B_1 <: C$, and $A <: B_1 \cap
-    B_2$.  We have $A <: B_1$ (Proposition~\ref{prop:subtyping} part
+  \item[Case (lb$_L$)] So $B = B_1 \cap B_2$, $A <: B_1 \cap
+    B_2$, and $B_1 <: C$.
+    We have $A <: B_1$ (Proposition~\ref{prop:subtyping} part
     \ref{prop:⊔⊑-inv}), so we conclude that $A <: C$ by the induction
@@
     and $\size{B_1} < \size{B}$.
-  \item[Case (lb$_R$)] So $B = B_1 \cap B_2$, $B_2 <: C$, and $A <:
-    B_1 \cap B_2$.  We have $A <: B_2$
+  \item[Case (lb$_R$)] So $B = B_1 \cap B_2$, $A <:
+    B_1 \cap B_2$, and $B_2 <: C$.  We have $A <: B_2$
     (Proposition~\ref{prop:subtyping} part \ref{prop:⊔⊑-inv}), so we

### 2020-05-15 papers/LMCS-subtyping-intersections/paper.tex
check

@@
 hope to omit the $({\to}{\cap})$ rule, but it plays an import role in
-filter models; without it they are not adequate with respect to
-reduction semantics.
+ensuring that the filter model produces a compositional semantics.
 

### 2020-05-15 papers/LMCS-subtyping-intersections/paper.tex
check

@@
   \item in type systems for programming
-    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Amin:2017aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa,Dotty:2020aa},
+    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Amin:2017aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa},
   \item in precise static
@@
 
-\section{Transitivity}
+\section{Proof of Transitivity of Subtyping}
 \label{sec:trans}

### 2020-05-19 papers/LMCS-subtyping-intersections/response.txt
check

@@
 
-The logical treatments of subtyping are indeed useful for extending to
-richer type systems for programming languages. The author has made use
-of them in the context of research on gradual typing.  However, the
-motivation for the present work is rather different: constructing
-filter models (denotational semantics) for programming languages. For
-that purpose, there is no need for richer type systems such as System
-F-omega; intersection types already provide what is needed to express
-the runtime behavior of functional programs. The revision includes
-changes to the introduction to differentiate the varied uses of
-intersection type systems and to make it clear that the present
-article's motivation comes from the realm of filter models, not type
-systems for programming languages.
+Yes, the logical treatments of subtyping are useful for extending to
+richer type systems for programming languages. However, the motivation
+for the present work is in constructing filter models (denotational
+semantics) for programming languages. For that purpose, there is no
+need for richer type systems such as System F-omega; intersection
+types already provide what is needed to express the runtime behavior
+of functional programs. The revision includes changes to the
+introduction to differentiate the varied uses of intersection type
+systems and to make it clear that the present article's motivation
+comes from the realm of filter models, not type systems for
+programming languages.
 
@@
 That is an interesting connection. I think that this style of proof
-would work using the new function rule (→′) but not with the old rules
+would work using the new function rule (→′) but not with the BCD rules
 (→) and (→∩). I've added a discussion of this connection to the
@@
 >> Minor notes:
-
 >> The paper has a few formatting issues with text running into the
 >> margin (p1, p8, some URLs in the bibliography).
-
 >> Typos: 
-
 >> "and then then prove transitivity" (p2) double "then"
-
 >> "Before moving on, we make note of some basic facts" perhaps "take
 >> note" or "make a note of"?
-
 >> Bibliography:
-
 >> "M. Dezani-Ciancaglini, F. Honsell, and Y. Motohama. Compositional
 >> characterisations of -terms" missing a lambda symbol
-
 >> "E. Engeler. Algebras and combinators. algebra universalis" capitalize
 >> journal title (?)
-
 >> "H. Ishihara and T. Kurata. Completeness of intersection and union

### 2020-05-19 papers/LMCS-subtyping-intersections/paper.tex
check

@@
 %\usepackage{cmbright}
-\usepackage{natbib}
+\usepackage[numbers]{natbib}
 \usepackage{amsthm}
@@
 satisfies the ``subformula conjunction property'': every type
-  occuring in the derivation of $A <: B$ is a subformula of $A$ or
+occurring in the derivation of $A <: B$ is a subformula of $A$ or
 $B$, or an intersection of such subformulas. The article proves that
@@
 \begin{enumerate}
-  \item in type systems for programming
-    languages~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Amin:2017aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa},
+  \item in type systems~\citep{Reynolds:1988aa,Pierce:1991aa,Castagna:2014aa,Chaudhuri:2014aa,Oliveira:2016aa,Amin:2017aa,Muehlboeck:2018aa,Bi:2019aa,Dunfield:2019aa,Microsoft:TypeScript2020aa},
   \item in precise static
@@
     and
-  \item in filter models for the denotational semantics of a variety
-    of lambda
+  \item in the denotational semantics for a variety of lambda
     calculi~\citep{Coppo:1980ab,Coppo:1981aa,Coppo:1984aa,Honsell:1992aa,Abramsky:1993fk,Honsell:1999aa,Ishihara:2002aa,Rocca:2004aa,Dezani-Ciancaglini:2005aa,Alessi:2006aa,Barendregt:2013aa}.
@@
 intersection types, in denotational semantics. We are interested in
-constructing filter models for functional programming languages in
-support of mechanized compiler
+constructing filter models for the denotational semantics of
+functional programming languages in support of mechanized compiler
 verification. Section~\ref{sec:motivation} describes how that research
@@
 CakeML~\citep{Kumar:2014aa,Owens:2017aa}. In particular, we are
-constructing filter models (based on intersection type systems)
-because they are relatively straightforward to mechanize in a proof
-assistant compared to traditional domain
+constructing filter models because they are relatively straightforward
+to mechanize in a proof assistant compared to traditional domain
 theory~\citep{Benton:2009ab,Dockins:2014aa}.  In a filter model, the
@@
 
-This problem is not conceptually difficult to solve; one just needs to
-restrict the graphs to only allow approximations of functions. This
-can be accomplished by requiring every entry in a function's graph to
-be ``consistent''. Two entries are consistent, written $(A \to B) \sim
-(A' \to B')$, when either $A \not\sim A'$ or $A \sim A'$ and $B \sim
-B'$. Two type constants are consistent when they are equal. One would
-then hope to prove that for any term $M$, if $\Gamma \vdash M : C$ and
-$\Gamma \vdash M : C'$, then $C \sim C'$ as long as $C$, $C'$, and
-$\Gamma$ are self consistent. A key lemma needed for that proof is
-that consistency is upward closed with respect to subtyping: if $A
-\sim B$, $A <: C$, $B <: D$, then $C \sim D$.
+To solve this problem, one needs to restrict the graphs to only allow
+approximations of functions. This can be accomplished by requiring
+every entry in a function's graph to be ``consistent''. Two entries
+are consistent, written $(A \to B) \sim (A' \to B')$, when either $A
+\not\sim A'$ or $A \sim A'$ and $B \sim B'$. Two type constants are
+consistent when they are equal. One would then hope to prove that for
+any term $M$, if $\Gamma \vdash M : C$ and $\Gamma \vdash M : C'$,
+then $C \sim C'$ as long as $C$, $C'$, and $\Gamma$ are self
+consistent. A key lemma needed for that proof is that consistency is
+upward closed with respect to subtyping: if $A \sim B$, $A <: C$, $B
+<: D$, then $C \sim D$.
 
@@
 significant increase in obligations elsewhere in our proofs. By some
-serendipity, around that moment in time \citet{Laurent:2018aa}
-published the first subtyping system for intersection types that
-removes the transitivity rule and satisfies the subformula property.
-Laurent developed a sequent-style subtyping system, written $\Gamma
-\vdash B$, where $\Gamma$ is a sequence of types $A_1,\ldots,A_n$. The
-intuition is that $A_1,\ldots,A_n \vdash B$ corresponds to $A_1 \cap
-\cdots \cap A_n \leq B$. This system satisfies the subformula property
-and is equivalent to the BCD system. To prove this, Laurent
-establishes six lemmas that culminate in cut-elimination, from which
-transitivity follows. Laurent mechanized these proofs in Coq.
+serendipity, around that time \citet{Laurent:2018aa} published a
+subtyping system that removes the transitivity rule and satisfies the
+subformula property.  Laurent developed a sequent-style subtyping
+system, written $\Gamma \vdash B$, where $\Gamma$ is a sequence of
+types $A_1,\ldots,A_n$. The intuition is that $A_1,\ldots,A_n \vdash
+B$ corresponds to $A_1 \cap \cdots \cap A_n \leq B$. This system
+satisfies the subformula property and is equivalent to the BCD
+system. To prove this, Laurent establishes six lemmas that culminate
+in cut-elimination, from which transitivity follows. Laurent
+mechanized these proofs in Coq.
 
-We immediately took advantage of Laurent's result to complete a filter
-model for the ISWIM
-language~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa}.
-We mechanized the formal development and proofs using the Isabelle
-proof assistant~\citep{Siek:2018aa}.
+We immediately used Laurent's result to complete a filter model for
+ISWIM~\citep{Landin:1966la,G.-D.-Plotkin:1975on,Felleisen:2009aa} and
+mechanized the results in the Isabelle proof
+assistant~\citep{Siek:2018aa}.
 
@@
 technically satisfy the subformula property but it satisfies what we
-call the \emph{subformula conjunction property}: every type occuring
+call the \emph{subformula conjunction property}: every type occurring
 in the derivation of $A <: B$ is a subformula of $A$ or $B$ or an
@@
 Section~\ref{sec:trans}, and prove its equivalence to BCD subtyping in
-Section~\ref{sec:equiv}. We conclude the artcle in
+Section~\ref{sec:equiv}. We conclude the article in
 Section~\ref{sec:conclude}.
@@
 repository.
-
 \begin{center}
@@
 \prec: B$, where $\mathcal{L}$ is a queue of types that are peeled off
-from the domain of the type on the right. So the judgement
+from the domain of the type on the right. So the judgment
 $\mathcal{L} \vdash A \prec: B$ is equivalent to $A \leq \mathcal{L}
@@
 but it corrects some errors and introduces more lemmas concerning an
-auxilliary notion called reflexive supertypes.
+auxiliary notion called reflexive supertypes.
 
@@
 The $\topP{A}$ predicate identifies types that are equivalent to
-$\TOP$. (See Propostion~\ref{prop:top}(\ref{prop:AllBot-⊑-any}).) The
+$\TOP$. (See Proposition~\ref{prop:top}(\ref{prop:AllBot-⊑-any}).) The
 $\topInCod{A}$ predicate identifies types that have $\TOP$ in their
@@
 
-The new intersection subtyping system, with judgements of the form $A
+The new intersection subtyping system, with judgments of the form $A
 <: B$, is defined in Figure~\ref{fig:new-subtyping}. First, it does
@@
 
-\bibliographystyle{abbrvnat}
+%\bibliographystyle{abbrvnat}
+\bibliographystyle{plainnat}
 \bibliography{all}
@@
 % LocalWords:  BCD lclr refl glb topInCod Coppo Dezani Ciancaglini fk
-% LocalWords:  Salle Engeler Honsell Abramsky Plotkin Rocca Alessi
+% LocalWords:  Salle Engeler Honsell Abramsky Plotkin Rocca Alessi lq
 % LocalWords:  subsumption subtype denotational ISWIM Landin Ishihara
-% LocalWords:  Felleisen Muehlboeck ISC Agda codomain reflexivity
-% LocalWords:  Bakel
+% LocalWords:  Felleisen Muehlboeck ISC Agda codomain reflexivity Coq
+% LocalWords:  Bakel subformulas Castagna Chaudhuri Oliveira Amin
+% LocalWords:  Dunfield TypeScript Turbak Palsberg Mossin Simoes Siek
+% LocalWords:  compositional CakeML Kumar Dockins runtime Appel agda
+% LocalWords:  subexpression Tarditi denotationally distributivity
+% LocalWords:  Scala supertypes Odersky Rompf pushback associativity
+% LocalWords:  commutativity

### 2021-08-05 blog/filter-vs-graph.tex
check

--- 2019-06-25 papers/denot-ISWIM.tex
@@
 
-\title{An Elementary Denotational \\ Semantics for ISWIM}
+\title{Filter versus Graph Models}
 \author{Jeremy G. Siek}
@@
 
-\begin{abstract}
-  This article takes a baby step toward developing denotational
-  semantics that are suitable for widespread use by programming
-  language practitioners and researchers who work outside of
-  semantics, e.g, in compiler correctness, static analysis, language
-  standards, etc. The current practice in such areas is to use
-  operational semantics, and it is quite successful. However, such
-  semantics are not compositional, which complicates reasoning about
-  them.  Denotational semantics are compositional, but their
-  mathematical sophistication has made them too time consuming for
-  practitioners to understand and customize to their uses. One purpose
-  of this article is to see how little mathematics is necessary to
-  develop compositional semantics.
+Filter models and graph models are two different ways to give a
+denotational semantics to the lambda calculus. At first glance they
+appear quite different, but many of those differences do not matter.
+On the other hand, there are some differences that do matter.
+Differences that matter are ones that cause significant differences in
+the proofs about the semantics, such as proofs of soundness and
+adequacy with respect to the reduction semantics.
 
-  As a case study, this article develops a denotational semantics for
-  the ISWIM programming language, that is, for a call-by-value lambda
-  calculus extended with arbitrary primitive operations and literals.
-  It models lambda expressions with lookup tables, a simple idea that
-  goes back to the graph and filter models for the lambda
-  calculus. Unfortunately, those models cannot be used as-is for
-  ISWIM: lookup tables are too loose in that they can represent
-  arbitrary relations, not just functions. So the article presents a
-  way to restrict lookup tables to just functions. The article proves
-  the soundness and adequacy of the resulting denotational semantics,
-  proves that denotational equality implies contextual equivalence,
-  and demostrates uses of the semantics in compiler correctness proofs
-  (inlining and closure conversion).
-\end{abstract}
-
-\section{Introduction}
-
-\citet{Barendregt:2013aa}
-
-\section{Semantics of ISWIM}
-
-Values
-\[
-\begin{array}{lcll}
-  \mathbb{A} & ::= & \mathbb{Z} \mid \mathbb{B} \mid \cdots & \text{atomic types}\\
-  \mathbb{P} & ::= & \mathbb{A} \mid \mathbb{A} \to \mathbb{P} & \text{primitive types}\\
-  c & \in & \mathbb{P} & \text{constants} \\
-  u,v,w & ::= & c \mid \bot \mid v \mapsto w \mid u \sqcup v & \text{values}
-\end{array}
-\]
-
-\begin{align*}
-  \curry{D}\, \gamma\, (v \mapsto w) &= D \,(\gamma,v)\, w\\
-  \curry{D}\, \gamma\, (u \sqcup v) &= D \,\gamma\, u \text{ and } D\, \gamma\, v \\
-  \curry{D}\, \gamma\, \bot &= \true \\
-\end{align*}
-
-\begin{align*}
-  \apply{D}{E}\,\gamma\,w &=
-    \exists v.\, D\,\gamma\,(v \mapsto w) \text{ and } E \, \gamma \, v
-\end{align*}
-
-\section{Conclusion}
+\bibliographystyle{abbrvnat}
+\bibliography{all}
 
